---
title: "Batch correction (diaPASEF) — removeBatchEffect for clustering, PCA and curve plots"
output: html_notebook
---

# Purpose

Produce a **batch-corrected intensity matrix** for the `hme1_lfq` / `hme1_diaPASEF` dataset, to be
consumed by the downstream Python clustering, PCA and visualisation code.

The 219 injections were acquired in **three enrichment + LC batches**. Each batch carries an overall
additive offset — the whole run sits a little higher or lower on the log2 scale than its neighbours —
and that offset is a technical artefact of when the samples were enriched and injected, not biology.
Left in, it becomes one of the largest directions of variance in the matrix, which means PC1 of a PCA
and the first split of any clustering can be *"which batch was this run in"* rather than *"how does
this site respond to EGF"*.

This notebook removes that offset with `limma::removeBatchEffect`, and validates the removal against
the three pooled reference injections.

> This notebook and `notebooks/01_preprocessing/batch_correction_diapasef.R` are the two drivers
> of the same analysis — the `.R` for `Rscript` / cron use, this one for reading the diagnostics
> as they are produced. **They now share one implementation** (`src/batch_correction_diapasef.R`),
> so the helper functions can no longer drift apart.
>
> ⚠️ They are **not** configured the same, and as of 2026-09-14 they are not equivalent:
>
> | | this notebook | the `.R` script |
> |---|---|---|
> | `OUTPUT_SCALE` | `"linear"` | `"log2"` |
> | suffixes | `_batchcorrected_linear*` | `_batchcorrected*` |
> | what is written | **all** injections, references included, in input-file column order | biological samples only; the references are held out |
> | paths | from `PROJECT_ROOT` | from `PROJECT_ROOT`, overridable by command-line arguments |
>
> The write-out difference is a real behavioural difference, not a setting — decide which one
> is wanted and make the other match.

# The two tracks: subtract for the eyes, adjust for the test

There are two legitimate ways to deal with an additive batch offset, and this project uses **both** —
for different consumers, on different matrices. Getting them mixed up is the single most likely way
to misuse this notebook's output, so it is stated first.

| | ADJUST for the test | SUBTRACT for the eyes |
|---|---|---|
| where | `limma_for_pvalues_diapasef.rmd` | **this notebook** |
| data fitted | the **uncorrected** values | — |
| how batch is handled | a term in the design matrix, `~ ... + batch` | subtracted from the data up front |
| consumers | p-values, FDR, responsiveness filters | clustering, PCA, UMAP, profile and curve plots |

**Why the split.** limma has a design matrix to put a nuisance term into, so it can estimate the
batch offsets *jointly* with the biology and charge the degrees of freedom for having done so — the
residual df, the moderated variance and therefore the p-values all stay correct. Clustering and PCA
have no design matrix. They see whatever numbers they are handed, so for them the offset has to be
gone from the numbers themselves.

⚠ **The matrix this notebook writes must never be fed into limma, or into any other inference step.**
Its batch degrees of freedom have already been spent, but a model fitted on it does not know that, so
it would treat corrected residuals as if they were free — anti-conservative p-values. If the
inference needs changing, that is a deliberate change to the limma notebook, not a change of input
file.

# Why `removeBatchEffect`, and why location-only

- **Same model as the inference.** `removeBatchEffect` removes exactly the additive batch term that
  `~ ... + batch` estimates. The matrix we cluster and the matrix we test then differ only in *where*
  the batch term was accounted for, not in *what the batch term is*. ComBat/sva would fit a different
  model (empirical-Bayes, and for ComBat also multiplicative), so the picture and the statistics
  would start telling slightly different stories about the same data.
- **The effect we have evidence for is a shift.** An enrichment-run + LC-session difference moves the
  whole run up or down; that is a location effect. A scale or variance correction would be fitting
  something we have not observed, and on DIA data with per-run missingness a per-batch *variance*
  estimate is exactly the kind of quantity that is dominated by which sites happened to be quantified
  rather than by any real dispersion difference.
- **`batch` is a plain fixed effect with 3 levels.** Three levels is far too few for empirical-Bayes
  shrinkage of the batch term to mean anything. A plain fixed term gets the bookkeeping right, and
  nothing is gained by being clever.
- **`design` carries the biology**, which is what tells `removeBatchEffect` what to leave alone
  (see *The model* below).

# What is deliberately not done

- **No normalisation** — no median centring, no quantile normalisation, no scaling. The `_None`
  unnormalised FragPipe variant is the deliberate input; between-sample scaling is a separate
  decision that is still open on this dataset (`NORMALIZE` in the limma notebook) and must not be
  smuggled in here.
- **No zero-filling, no imputation.** Missing intensities stay `NA` from load to write (project
  rule). Zeros would corrupt the linear fit and destroy the reference CVs — a zero is a
  non-detection, not a measurement of nothing.
- **No injection-order term.** Order was randomised within each batch, so LC drift is orthogonal to
  the biology: it does not bias any contrast, it only inflates within-batch variance. Adding the term
  would spend degrees of freedom to buy nothing, and on sparse DIA sites it would start fitting the
  missingness pattern. A disabled hook is left in the model chunk; wire it in **only if** a QC plot
  of total phospho-signal against injection order shows clear within-batch monotonic drift.
- **The pooled references are never clustered or tested.** They are held out for QC and dropped from
  the written matrix.

# Inputs

**Input 1 — the filtered abundance matrix.** The single-site diaPASEF table *after* the project's
standard filtering (`n:reps`, contaminants CON/REV, localization, incomplete time series), sites in
rows, one column per injection plus the usual annotation columns.

**Input 2 — the sample sheet**, one row per injection, written by
`notebooks/01_preprocessing/LFQ_diaPASEF_batch_effect_correction.ipynb`:

| column | notes |
|---|---|
| `sample_id` | **must match the abundance-matrix column header exactly** — this is the join key |
| `cell_line` | BRAF biological replicates 1 and 2 are **distinct** levels (`BRAFS151A1`, `BRAFS151A2`) |
| `condition` | includes the controls (`full`, `starve`) as their own levels |
| `timepoint` | numeric minutes; controls are `NA` or 0 |
| `batch` | `1` / `2` / `3` — the only factor that gets removed |
| `is_reference` | `TRUE` for the pooled references (`mix`, `mixb`, `mixc`), else `FALSE` |
| `injection_order` | optional; **not used** unless the disabled drift hook is turned on |

The generating notebook writes the id column as `sample_Id`; the specification calls it `sample_id`.
`canonicalise_sheet()` accepts either spelling and normalises it, so neither file has to change.

**Sample columns are identified only by joining the header against `sample_id`** — never by
string-pattern matching on column names. Anything that does not join is treated as an annotation
column, and a *numeric* column that fails to join is a hard error (it is almost always an injection
missing from the sheet, which would otherwise silently escape correction).

```{r setup}
# ── Helper functions ────────────────────────────────────────────────────────────────────
# They live in src/ so this notebook and its sibling script
# notebooks/01_preprocessing/batch_correction_diapasef.R share ONE implementation — before the
# move the two carried their own copies and had already drifted apart. src/ has to be found
# before find_project_root() is available (it is defined in r_utils.R), hence the candidate
# loop: knitting runs with this file's folder as the working directory, running the chunks
# from the IDE console runs with the project root, and one relative path cannot mean both.

SRC_DIR <- NULL
for (candidate in c("src", file.path("..", "..", "src"))) {
  if (file.exists(file.path(candidate, "r_utils.R"))) {
    SRC_DIR <- normalizePath(candidate)
    break
  }
}
if (is.null(SRC_DIR)) {
  stop("setup: cannot find src/ from '", getwd(),
       "'. Set the working directory to the project root or to notebooks/01_preprocessing/.")
}
source(file.path(SRC_DIR, "r_utils.R"))                    # find_project_root, say
source(file.path(SRC_DIR, "batch_correction_diapasef.R"))  # output_path, read_tsv, ...

PROJECT_ROOT <- find_project_root()

# --- inputs ---------------------------------------------------------------------------
# Input 1: the filtered single-site diaPASEF abundance matrix. Deliberately the
#          UNNORMALISED `_None` variant — do not swap it for `_Norm`.
ABUNDANCE_FILE <- file.path(PROJECT_ROOT, "Experiment", "hme1_diaPASEF", "Data", "Processed",
                            "20260821_peptide_MS2quant_None_renamed.tsv")

# Input 2: the sample sheet, one row per injection.
SAMPLE_SHEET_FILE <- file.path(PROJECT_ROOT, "Experiment", "hme1_diaPASEF", "Data",
                               "batch_info.tsv")

# --- outputs --------------------------------------------------------------------------
# Default: next to the abundance file. The suffixes guarantee the input is never the output.
OUT_DIR <- dirname(ABUNDANCE_FILE)

# The suffixes carry OUTPUT_SCALE, so the log2 and linear runs cannot collide and neither
# can overwrite the other. The reference QC is computed on the log2 matrix and is therefore
# IDENTICAL between the two runs — the scale token in its name is for pairing, not content.
OUT_SUFFIX_MATRIX  <- "_batchcorrected_linear.tsv"        # the clustering matrix (references dropped)
OUT_SUFFIX_REF_QC  <- "_batchcorrected_linear_refQC.tsv"  # per-site reference spread, before vs after
OUT_SUFFIX_REF_PNG <- "_batchcorrected_linear_refQC.png"  # before/after diagnostic figure

# Refuse to clobber anything that already exists (project rule: never overwrite).
OVERWRITE <- FALSE

# --- sample sheet schema --------------------------------------------------------------
# Both id spellings are accepted; whichever is present is canonicalised to `sample_id`.
ID_COLUMN_CANDIDATES <- c("sample_id", "sample_Id", "sample_ID", "sampleId")

REQUIRED_SHEET_COLUMNS <- c("cell_line", "condition", "timepoint", "batch", "is_reference")

# --- scale detection ------------------------------------------------------------------
# removeBatchEffect operates on log-scale data. Real log2 phospho intensities live in
# roughly [5, 35]; linear detector intensities run to 1e5-1e9. Anything above this
# threshold is therefore unambiguously linear and gets log2-transformed exactly once.
LOG2_MAX_PLAUSIBLE <- 64

# --- output scale ---------------------------------------------------------------------
# "log2"   -> write the corrected values on the log2 scale the correction was done on, and
#             rename the DataType token of every sample column `raw:abs` -> `log2:abs`, so
#             the header does not claim to hold linear intensities that it does not.
# "linear" -> back-transform with 2^x so the file can be fed straight into
#             `run_diapasef_transformations()` in src/transformations.py, which expects
#             `raw:abs` linear intensities and does its own log2.
# The correction happens on log2 either way; this only chooses what is written to disk.
#
# NOTE on "linear": 2^x is the corrected value back on a linear scale, NOT the original
# measurement. Each cell has been multiplied by a per-(site, batch) factor — that is the
# whole point of the correction. The column keeps the `raw:abs` DataType token because the
# SCALE is genuinely linear, but the numbers are no longer "as measured by the detector".
# Feeding them to run_diapasef_transformations() is exact, not approximate: that function
# takes log2 immediately, and log2(2^x) == x to floating-point precision.
OUTPUT_SCALE <- "linear"

RENAME_DATA_TYPE <- c(from = "raw:abs", to = "log2:abs")

# --- annotation columns that are legitimately numeric ---------------------------------
# Used ONLY by the safety check in split_matrix_columns(): a numeric column that is neither
# a known sample nor listed here is almost certainly an injection missing from the sheet.
NUMERIC_ANNOTATION_COLUMNS <- c("Index", "peptide_index", "Start", "End", "Peptide Length",
                                "Probability", "Multiplicity", "Precursor Count",
                                "Best Localization", "Best Scan for Localization",
                                "protein_length", "nrPeptides", "nr_tryptic_peptides",
                                "NumPhos", "LocalizedNumPhos", "MaxPepProb", "AScore",
                                "n:reps", "ReferenceIntensity", "startModSite", "endModSite")

# --- switches -------------------------------------------------------------------------
# Injection-order drift hook — see the disabled block in the model chunk. Leave FALSE.
USE_INJECTION_ORDER <- FALSE

# Re-fit the same linear model once more, purely to count how many sites had a batch
# coefficient that could not be estimated. Costs one extra lmFit over the whole matrix;
# set FALSE to skip it on a very large file.
REPORT_NONESTIMABLE <- TRUE

VERBOSE <- TRUE
options(tmt.verbose = VERBOSE)   # read by say() in src/r_utils.R
```

# Helper functions

Every function below is pure except `say()` (prints) and `plot_reference_qc()` (draws). The
correction itself is a single call to `limma::removeBatchEffect` in the *Correct* chunk — nothing
here re-implements it.

Every helper lives in **`src/batch_correction_diapasef.R`**, sourced by the `setup` chunk
(`say()` and `find_project_root()` come from **`src/r_utils.R`**, shared with the limma and
Trendy notebooks). The **same file is sourced by the sibling script**
`notebooks/01_preprocessing/batch_correction_diapasef.R`, so the two drivers can no longer
drift apart — only their configuration differs.

| function | what it does |
|---|---|
| `output_path` | input stem + descriptive suffix; refuses to overwrite |
| `read_tsv` | project TSV reader — `check.names = FALSE`, missing values stay `NA` |
| `canonicalise_sheet` | validate the sample sheet, normalise the id spelling and the types |
| `split_matrix_columns` | sample vs annotation columns, by joining the header against `sample_id` |
| `to_log2_if_needed` | put the matrix on the log2 scale exactly once |
| `build_bio_group` | the biological factor the correction must protect |
| `check_batch_orthogonality` | is every biological group present in every batch? |
| `count_nonestimable` | per site, how many batch coefficients limma could not estimate |
| `reference_spread` | per-site SD / CV across the pooled references |
| `plot_reference_qc` | the before/after diagnostic figure |

Every function is pure except `say()` (prints) and `plot_reference_qc()` (draws). The correction
itself is a single call to `limma::removeBatchEffect` in the *Correct* chunk — nothing in the
module re-implements it.

⚠️ Arguments that used to default to a notebook global (`OUT_DIR`, `OVERWRITE`,
`ID_COLUMN_CANDIDATES`, `REQUIRED_SHEET_COLUMNS`, `NUMERIC_ANNOTATION_COLUMNS`,
`LOG2_MAX_PLAUSIBLE`, `VERBOSE`) are now explicit arguments with literal defaults, and the
chunks below pass the configuration in. A sourced file must not silently depend on variables
defined by whoever sourced it. ⚠️ `read_tsv()` shadows `readr::read_tsv` — do not attach readr
in a session that sources the module.

The originals are kept below, commented out, for comparison. The chunk is never evaluated and
never rendered.

```{r functions-moved-to-src, eval=FALSE, include=FALSE}
# ─────────────────────────────────────────────────────────────────────────────────────
# MOVED to src/batch_correction_diapasef.R (and say() to src/r_utils.R).
# Kept commented out so the two versions can be compared.
# ─────────────────────────────────────────────────────────────────────────────────────
# say <- function(...) {
#   # Print a progress line, honouring the VERBOSE switch.
#   #
#   # Args:
#   #   ...: objects pasted together to form the message.
#   #
#   # Returns:
#   #   NULL, invisibly. Called for the side effect of writing to stdout.
#
#   if (VERBOSE) cat(..., "\n", sep = "")
#   invisible(NULL)
# }
#
#
# output_path <- function(input_path,
#                         suffix,
#                         out_dir = OUT_DIR) {
#   # Build an output path from the input filename plus a descriptive suffix.
#   #
#   # The input extension is stripped and the suffix appended, so the input file can never be
#   # the output file. Stops if the target exists and OVERWRITE is FALSE (project rule: never
#   # overwrite a processed data file, always version the name).
#   #
#   # Args:
#   #   input_path: path of the abundance matrix the output is derived from.
#   #   suffix: string appended to the input stem, including its own extension.
#   #   out_dir: directory to write into.
#   #
#   # Returns:
#   #   Absolute path of the file to write.
#
#   stem <- sub("\\.[A-Za-z0-9]+$", "", basename(input_path))
#   path <- file.path(out_dir, paste0(stem, suffix))
#
#   if (file.exists(path) && !OVERWRITE) {
#     stop("output_path: '", path, "' already exists and OVERWRITE is FALSE. ",
#          "Move it aside or add a version tag to the suffix.")
#   }
#   if (normalizePath(path, mustWork = FALSE) == normalizePath(input_path, mustWork = FALSE)) {
#     stop("output_path: refusing to write over the input file '", input_path, "'.")
#   }
#   path
# }
#
#
# read_tsv <- function(path,
#                      what) {
#   # Read a project TSV, keeping column names verbatim and missing values as NA.
#   #
#   # `check.names = FALSE` is essential: the sample column names contain ':' and the header
#   # would otherwise be mangled into `WT_raw.abs_EGF_full_r1`, breaking the join against the
#   # sample sheet. Empty strings and the usual sentinels become NA — nothing is ever
#   # zero-filled (project rule).
#   #
#   # Args:
#   #   path: file to read.
#   #   what: short human-readable description used in error messages.
#   #
#   # Returns:
#   #   A data.frame with the file's columns in file order.
#
#   if (!file.exists(path)) {
#     stop("read_tsv: ", what, " not found at '", path, "'.")
#   }
#   df <- utils::read.delim(path,
#                           sep = "\t",
#                           header = TRUE,
#                           check.names = FALSE,
#                           stringsAsFactors = FALSE,
#                           na.strings = c("", "NA", "NaN", "nan", "None"))
#   say("  loaded ", what, ": ", nrow(df), " rows x ", ncol(df), " columns")
#   df
# }
#
#
# canonicalise_sheet <- function(sheet) {
#   # Validate the sample sheet and normalise its column names and types.
#   #
#   # Renames whichever id spelling is present to `sample_id`, checks the required columns
#   # exist, coerces `is_reference` from R logicals / "TRUE"/"FALSE" / 0/1 to a logical
#   # vector, and forces `batch` to a factor. Stops loudly on anything missing or ambiguous
#   # rather than guessing.
#   #
#   # Args:
#   #   sheet: data.frame as read from the sample-sheet TSV.
#   #
#   # Returns:
#   #   The same data.frame with `sample_id` (character), `batch` (factor), `is_reference`
#   #   (logical), and the remaining required columns present.
#
#   id_present <- intersect(ID_COLUMN_CANDIDATES, colnames(sheet))
#   if (length(id_present) == 0L) {
#     stop("canonicalise_sheet: the sample sheet has no id column. Expected one of: ",
#          paste(ID_COLUMN_CANDIDATES, collapse = ", "),
#          ". Found: ", paste(colnames(sheet), collapse = ", "))
#   }
#   if (length(id_present) > 1L) {
#     stop("canonicalise_sheet: the sample sheet has more than one id column (",
#          paste(id_present, collapse = ", "), "). Keep exactly one.")
#   }
#   colnames(sheet)[colnames(sheet) == id_present[[1L]]] <- "sample_id"
#
#   missing_cols <- setdiff(REQUIRED_SHEET_COLUMNS, colnames(sheet))
#   if (length(missing_cols) > 0L) {
#     stop("canonicalise_sheet: sample sheet is missing required column(s): ",
#          paste(missing_cols, collapse = ", "))
#   }
#
#   sheet$sample_id <- as.character(sheet$sample_id)
#   if (anyDuplicated(sheet$sample_id) > 0L) {
#     dups <- unique(sheet$sample_id[duplicated(sheet$sample_id)])
#     stop("canonicalise_sheet: duplicated sample_id in the sheet: ",
#          paste(dups, collapse = ", "))
#   }
#
#   # is_reference: accept logical, "TRUE"/"FALSE"/"True"/"true", or 0/1.
#   ref_raw <- sheet$is_reference
#   ref <- if (is.logical(ref_raw)) {
#     ref_raw
#   } else if (is.numeric(ref_raw)) {
#     ref_raw != 0
#   } else {
#     tolower(trimws(as.character(ref_raw))) %in% c("true", "t", "yes", "1")
#   }
#   if (anyNA(ref)) {
#     stop("canonicalise_sheet: `is_reference` could not be read as a logical for ",
#          sum(is.na(ref)), " row(s).")
#   }
#   sheet$is_reference <- ref
#
#   sheet$batch <- factor(as.character(sheet$batch))
#   if (nlevels(sheet$batch) < 2L) {
#     stop("canonicalise_sheet: `batch` has ", nlevels(sheet$batch),
#          " level(s). Nothing to correct — check the sample sheet.")
#   }
#
#   say("  sample sheet: ", nrow(sheet), " injections, ",
#       nlevels(sheet$batch), " batches (", paste(levels(sheet$batch), collapse = "/"), "), ",
#       sum(sheet$is_reference), " pooled references")
#   sheet
# }
#
#
# split_matrix_columns <- function(abundance,
#                                  sheet) {
#   # Split the abundance table into sample columns and annotation columns.
#   #
#   # Sample columns are identified ONLY by matching the header against the sheet's
#   # `sample_id` — never by pattern-matching on the column name. Two failures are fatal: a
#   # `sample_id` with no matching column, and a numeric column that is in neither the sheet
#   # nor NUMERIC_ANNOTATION_COLUMNS (which almost always means an injection was left out of
#   # the sheet and would silently escape correction).
#   #
#   # Args:
#   #   abundance: data.frame as read from the abundance TSV.
#   #   sheet: canonicalised sample sheet.
#   #
#   # Returns:
#   #   A list with `sample_cols` (character, in sheet order) and `annot_cols` (character, in
#   #   file order).
#
#   sample_cols <- sheet$sample_id
#   missing_in_matrix <- setdiff(sample_cols, colnames(abundance))
#   if (length(missing_in_matrix) > 0L) {
#     stop("split_matrix_columns: ", length(missing_in_matrix),
#          " sample_id(s) in the sheet have no column in the abundance matrix: ",
#          paste(utils::head(missing_in_matrix, 20L), collapse = ", "),
#          if (length(missing_in_matrix) > 20L) " ..." else "")
#   }
#
#   annot_cols <- setdiff(colnames(abundance), sample_cols)
#
#   unexplained <- annot_cols[vapply(abundance[annot_cols], is.numeric, logical(1))]
#   unexplained <- setdiff(unexplained, NUMERIC_ANNOTATION_COLUMNS)
#   if (length(unexplained) > 0L) {
#     stop("split_matrix_columns: ", length(unexplained),
#          " numeric column(s) are neither a sample_id in the sheet nor a known annotation: ",
#          paste(unexplained, collapse = ", "),
#          ". If these are injections, add them to the sample sheet; if they are annotations, ",
#          "add them to NUMERIC_ANNOTATION_COLUMNS.")
#   }
#
#   say("  columns: ", length(sample_cols), " samples, ", length(annot_cols), " annotations")
#   list(sample_cols = sample_cols, annot_cols = annot_cols)
# }
#
#
# to_log2_if_needed <- function(x) {
#   # Put the intensity matrix on the log2 scale, exactly once.
#   #
#   # `removeBatchEffect` fits a linear model, so an additive batch offset only *is* additive
#   # on the log scale. Linear intensities are detected by their range (see
#   # LOG2_MAX_PLAUSIBLE) and transformed; values that are already log2 are left alone so the
#   # matrix is never double-logged. Non-positive values become NA, NOT zero — a zero
#   # intensity in DIA is a non-detection, and log2(0) = -Inf would drag the linear fit. No
#   # normalisation of any kind is applied.
#   #
#   # Args:
#   #   x: numeric matrix of intensities, injections in columns.
#   #
#   # Returns:
#   #   A list with `x` (the log2-scale matrix) and `was_linear` (logical, which branch ran).
#
#   finite_vals <- x[is.finite(x)]
#   if (length(finite_vals) == 0L) {
#     stop("to_log2_if_needed: the sample matrix contains no finite values.")
#   }
#
#   rng <- range(finite_vals)
#   was_linear <- rng[[2L]] > LOG2_MAX_PLAUSIBLE
#
#   say("  value range: [", signif(rng[[1L]], 4), ", ", signif(rng[[2L]], 4), "], ",
#       "median ", signif(stats::median(finite_vals), 4))
#
#   if (was_linear) {
#     say("  scale check -> LINEAR intensities: applying log2 (non-positive -> NA)")
#     x[!is.na(x) & x <= 0] <- NA_real_
#     x <- log2(x)
#   } else {
#     say("  scale check -> already LOG2: no transformation applied (not double-logging)")
#     x[!is.finite(x)] <- NA_real_
#   }
#
#   say("  missing values: ", sum(is.na(x)), " / ", length(x),
#       " (", sprintf("%.1f%%", 100 * mean(is.na(x))), ") — kept as NA, never zero-filled")
#
#   list(x = x, was_linear = was_linear)
# }
#
#
# build_bio_group <- function(sheet) {
#   # Build the biological factor that the correction must PROTECT.
#   #
#   # One level per (cell_line, condition, timepoint) combination — the finest biological unit
#   # in the experiment, so nothing biological is pooled and therefore nothing biological can
#   # be absorbed into the batch term. The pooled references are overridden to a single
#   # "reference" level: they are the same material in all three batches, so giving them their
#   # own free mean stops that material's overall abundance from being read as a biological
#   # group, while still letting them ride along and be corrected.
#   #
#   # BRAF biological replicates 1 and 2 are distinct `cell_line` levels in the sheet, so they
#   # stay distinct here — they are not pooled.
#   #
#   # Args:
#   #   sheet: canonicalised sample sheet, rows in the order the matrix columns will be in.
#   #
#   # Returns:
#   #   A factor with one entry per injection, levels made syntactically valid.
#
#   grp <- paste(sheet$cell_line, sheet$condition, sheet$timepoint, sep = ".")
#   grp[sheet$is_reference] <- "reference"
#   factor(make.names(grp))
# }
#
#
# check_batch_orthogonality <- function(bio_group,
#                                       batch) {
#   # Report how batch and biology are laid out against each other.
#   #
#   # The design of this experiment puts every biological group in every batch exactly once,
#   # which makes `batch` orthogonal to `bio_group`. That is the reason the correction cannot
#   # erode cell-line / condition / time effects: there is no biological contrast that is even
#   # partly a batch contrast, so subtracting the batch offsets removes a direction the
#   # biology does not live in. This function checks that assumption instead of assuming it,
#   # and warns (rather than stops) if the layout is imperfect — a few groups missing a batch
#   # is a missingness problem, not a design failure.
#   #
#   # Args:
#   #   bio_group: factor of biological groups, one entry per injection.
#   #   batch: factor of batches, one entry per injection.
#   #
#   # Returns:
#   #   The group x batch contingency table, invisibly.
#
#   tab <- table(bio_group, batch)
#   balanced <- all(tab == 1L)
#
#   if (balanced) {
#     say("  design: each biological group appears exactly once per batch -> ",
#         "batch is orthogonal to bio_group; the correction cannot touch the biology")
#   } else {
#     n_unbal <- sum(apply(tab, 1L, function(r) length(unique(r)) > 1L))
#     warning("check_batch_orthogonality: ", n_unbal, " of ", nrow(tab),
#             " biological groups are not evenly spread across batches. ",
#             "Batch and biology are partially confounded for those groups; ",
#             "the correction there is a compromise, not a clean removal.",
#             call. = FALSE)
#     say("  design: NOT fully balanced — ", n_unbal, " group(s) unevenly spread over batches")
#   }
#
#   invisible(tab)
# }
#
#
# count_nonestimable <- function(x,
#                                design,
#                                batch) {
#   # Count, per site, how many batch coefficients could not be estimated.
#   #
#   # A site too sparsely quantified in some batch leaves that batch's offset undetermined;
#   # limma returns NA and `removeBatchEffect` silently substitutes 0, i.e. that part of the
#   # correction is simply not applied. That is expected under DIA missingness and acceptable
#   # — but it should be visible rather than silent, which limma's single "Partial NA
#   # coefficients" warning does not make it (that warning also fires for NA *biological*
#   # coefficients, which are harmless here, so it cannot be read as a batch diagnostic).
#   #
#   # This re-fits the exact model `removeBatchEffect` fits internally (the batch factor in
#   # sum-to-zero coding, projected onto the orthogonal complement of `design`) purely to read
#   # the NA pattern out of the coefficients. It does NOT produce the corrected values — those
#   # come from `removeBatchEffect` itself, so there is one implementation of the correction,
#   # not two.
#   #
#   # Args:
#   #   x: log2-scale matrix, sites in rows, injections in columns.
#   #   design: biological design matrix passed to removeBatchEffect.
#   #   batch: factor of batches, one entry per injection.
#   #
#   # Returns:
#   #   Integer vector, one entry per site: the number of NA batch coefficients (0 = fully
#   #   corrected, ncol(batch contrasts) = not corrected at all).
#
#   b <- as.factor(batch)
#   stats::contrasts(b) <- stats::contr.sum(levels(b))
#   batch_mat <- stats::model.matrix(~ b)[, -1L, drop = FALSE]
#   batch_mat <- qr.resid(qr(design), batch_mat)
#
#   fit <- suppressWarnings(limma::lmFit(x, cbind(design, batch_mat)))
#   beta <- fit$coefficients[, -seq_len(ncol(design)), drop = FALSE]
#   rowSums(is.na(beta))
# }
#
#
# reference_spread <- function(x,
#                              ref_cols) {
#   # Per-site technical spread across the pooled reference injections.
#   #
#   # The references are the same physical material injected once per batch, so every
#   # difference between them is technical. Only sites quantified in ALL references are
#   # evaluable — under DIA missingness that is a subset, which is fine: this is a spot-check,
#   # not a per-site gate.
#   #
#   # CV is reported on the log2 scale as sd/|mean| * 100, which is a ratio of log2 units, not
#   # the linear-scale CV. SD in log2 units is the primary number; the CV column is kept
#   # because it is scale-free across sites.
#   #
#   # Args:
#   #   x: log2-scale matrix, sites in rows.
#   #   ref_cols: column names of the reference injections.
#   #
#   # Returns:
#   #   A data.frame with one row per site: `n_obs`, `mean`, `sd`, `cv` (NA where the site is
#   #   not observed in every reference).
#
#   m <- x[, ref_cols, drop = FALSE]
#   n_obs <- rowSums(!is.na(m))
#   complete <- n_obs == length(ref_cols)
#
#   mu <- rep(NA_real_, nrow(m))
#   sd <- rep(NA_real_, nrow(m))
#   mu[complete] <- rowMeans(m[complete, , drop = FALSE])
#   sd[complete] <- apply(m[complete, , drop = FALSE], 1L, stats::sd)
#
#   data.frame(n_obs = n_obs,
#              mean = mu,
#              sd = sd,
#              cv = 100 * sd / abs(mu),
#              row.names = rownames(m),
#              stringsAsFactors = FALSE)
# }
#
#
# plot_reference_qc <- function(before,
#                               after,
#                               path = NULL) {
#   # Draw the before/after reference-spread diagnostic, optionally saving it.
#   #
#   # Left: overlaid histograms of per-site SD across the references. Right: scatter of before
#   # vs after with the identity line — points below the line are sites whose technical spread
#   # shrank. Success is the whole cloud sitting below the diagonal.
#   #
#   # Args:
#   #   before: data.frame from reference_spread() on the uncorrected values.
#   #   after: data.frame from reference_spread() on the corrected values.
#   #   path: PNG file to write; NULL draws to the current device (inline in the notebook).
#   #
#   # Returns:
#   #   `path`, invisibly (NULL when drawing inline).
#
#   ok <- !is.na(before$sd) & !is.na(after$sd)
#   b <- before$sd[ok]
#   a <- after$sd[ok]
#
#   if (!is.null(path)) {
#     grDevices::png(path, width = 1600, height = 750, res = 140)
#     on.exit(grDevices::dev.off(), add = TRUE)
#   }
#   old_par <- graphics::par(mfrow = c(1L, 2L), mar = c(4.5, 4.5, 3, 1))
#   on.exit(graphics::par(old_par), add = TRUE)
#
#   brks <- graphics::hist(c(b, a), breaks = 40L, plot = FALSE)$breaks
#   hb <- graphics::hist(b, breaks = brks, plot = FALSE)
#   ha <- graphics::hist(a, breaks = brks, plot = FALSE)
#   graphics::plot(hb, col = grDevices::adjustcolor("grey40", 0.6), border = NA,
#                  ylim = c(0, max(hb$counts, ha$counts)),
#                  main = "Reference spread per site",
#                  xlab = "SD across mix / mixb / mixc (log2)", ylab = "sites")
#   graphics::plot(ha, col = grDevices::adjustcolor("#1f77b4", 0.6), border = NA, add = TRUE)
#   graphics::abline(v = stats::median(b), col = "grey20", lwd = 2, lty = 2)
#   graphics::abline(v = stats::median(a), col = "#1f77b4", lwd = 2, lty = 2)
#   graphics::legend("topright", bty = "n",
#                    fill = c(grDevices::adjustcolor("grey40", 0.6),
#                             grDevices::adjustcolor("#1f77b4", 0.6)),
#                    legend = c(sprintf("before (median %.3f)", stats::median(b)),
#                               sprintf("after  (median %.3f)", stats::median(a))))
#
#   lim <- range(c(b, a), finite = TRUE)
#   graphics::plot(b, a, pch = 16, cex = 0.35,
#                  col = grDevices::adjustcolor("black", 0.25),
#                  xlim = lim, ylim = lim,
#                  main = sprintf("%.1f%% of sites improved", 100 * mean(a < b)),
#                  xlab = "SD before (log2)", ylab = "SD after (log2)")
#   graphics::abline(0, 1, col = "red", lwd = 2)
#
#   invisible(path)
# }
```

# 1. Load

The two inputs are joined by name and then **ordered identically** — the matrix columns are taken in
sheet order, so every positional index downstream (`batch[j]`, `bio_group[j]`) refers to the same
injection as column `j`. This is the single point where the two files are aligned, and it is done by
name, so it cannot silently mis-pair.

```{r load}
say("=== 1. Load ===")
say("  abundance : ", ABUNDANCE_FILE)
say("  sheet     : ", SAMPLE_SHEET_FILE)

abundance <- read_tsv(ABUNDANCE_FILE, "abundance matrix")
sheet     <- canonicalise_sheet(read_tsv(SAMPLE_SHEET_FILE, "sample sheet"),
                                id_candidates    = ID_COLUMN_CANDIDATES,
                                required_columns = REQUIRED_SHEET_COLUMNS)

cols <- split_matrix_columns(abundance,
                             sheet,
                             numeric_annotation_columns = NUMERIC_ANNOTATION_COLUMNS)
sample_cols <- cols$sample_cols
annot_cols  <- cols$annot_cols

x <- as.matrix(abundance[, sample_cols, drop = FALSE])
storage.mode(x) <- "double"
rownames(x) <- if ("site" %in% annot_cols) abundance[["site"]] else rownames(abundance)
stopifnot(identical(colnames(x), sheet$sample_id))
```

# 2. Scale check

`removeBatchEffect` fits a linear model, so an additive batch offset only **is** additive on the log
scale. The branch taken is printed, and the matrix is never double-logged: if the input is already
`log2:abs`, nothing happens.

```{r scale-check}
say("=== 2. Scale check ===")
scaled <- to_log2_if_needed(x, log2_max_plausible = LOG2_MAX_PLAUSIBLE)
x <- scaled$x
```

# 3. The model

Two factors, doing opposite jobs.

**`bio_group` — what to protect.** A single factor with one level per
`(cell_line, condition, timepoint)` combination, plus `"reference"` for the pooled injections. This
is the finest biological unit in the experiment: because nothing biological is pooled into a shared
level, nothing biological can be absorbed into the batch term. Giving the references their own free
mean stops that pooled material — which is not any cell line at any timepoint — from being read as a
biological group of its own.

**`batch` — what to remove.** A 3-level fixed factor, unshrunk.

**Why this is safe.** `removeBatchEffect` projects the batch columns onto the orthogonal complement
of `design` before fitting, so any variance a batch column *shares* with a biological group is
credited to the biology and left alone. In this experiment each biological group appears exactly once
per batch, so batch is orthogonal to `bio_group` to begin with — there is no biological contrast that
is even partly a batch contrast. `check_batch_orthogonality()` verifies that from the sheet rather
than assuming it, and warns if the layout is not what is claimed here.

```{r model}
say("=== 3. Model ===")

bio_group <- build_bio_group(sheet)
batch     <- sheet$batch

say("  bio_group: ", nlevels(bio_group), " levels (incl. 'reference'), ",
    "batch: ", nlevels(batch), " levels")

check_batch_orthogonality(bio_group, batch)

design <- stats::model.matrix(~ 0 + bio_group)
colnames(design) <- levels(bio_group)

rank_full <- qr(design)$rank
if (rank_full < ncol(design)) {
  stop("model: the biological design matrix is rank deficient (rank ", rank_full,
       " < ", ncol(design), " columns). Check for empty bio_group levels in the sheet.")
}
say("  design: ", nrow(design), " injections x ", ncol(design),
    " biological coefficients (full rank), plus ", nlevels(batch) - 1L,
    " batch coefficient(s)")

# ---------------------------------------------------------------------------------------
# INJECTION ORDER — DISABLED HOOK, do not enable without evidence
# ---------------------------------------------------------------------------------------
# Injection order was randomised WITHIN each batch, so LC drift is orthogonal to the
# biology: it does not bias any contrast, it only inflates within-batch variance. Adding an
# order term would spend degrees of freedom to buy nothing, and on sparse DIA sites it would
# start fitting the missingness pattern.
#
# Wire this in ONLY IF a QC plot of total phospho-signal vs injection order shows a clear
# within-batch monotonic drift. It would then go in as a `covariates` term, NOT as a second
# batch factor:
#
# if (USE_INJECTION_ORDER) {
#   stopifnot("injection_order" %in% colnames(sheet))
#   # centre within batch so the term captures drift, not a second batch offset
#   drift <- unlist(lapply(split(sheet$injection_order, sheet$batch),
#                          function(v) scale(v, center = TRUE, scale = FALSE)))
#   drift <- drift[order(unlist(split(seq_len(nrow(sheet)), sheet$batch)))]
#   covariates <- matrix(drift, ncol = 1L, dimnames = list(NULL, "injection_order"))
# } else {
#   covariates <- NULL
# }
# ---------------------------------------------------------------------------------------
```

# 4. Correct

One call. `batch` is the axis removed, `design` is what is left alone.

Sites too sparse in a batch to estimate its offset get an `NA` coefficient, which `removeBatchEffect`
sets to `0` — those sites are simply left uncorrected in that direction. **This is expected under DIA
missingness and is not an error**; the count is reported so it is visible rather than implied.

Two assertions close the chunk: the correction is location-only, so nothing may appear or disappear.
The dimensions and the entire missingness pattern must be identical before and after — if either
changes, something imputed, and the run should be stopped.

```{r correct}
say("=== 4. removeBatchEffect ===")

corrected <- limma::removeBatchEffect(x,
                                      batch = batch,
                                      design = design)

say("  corrected ", nrow(corrected), " sites x ", ncol(corrected), " injections")

n_untouched <- sum(rowSums(abs(corrected - x), na.rm = TRUE) == 0)
say("  sites left completely unchanged (no batch coefficient estimable): ",
    n_untouched, " (", sprintf("%.1f%%", 100 * n_untouched / nrow(corrected)), ")")

if (REPORT_NONESTIMABLE) {
  na_counts <- count_nonestimable(x, design, batch)
  say("  sites with a non-estimable batch coefficient (NA -> 0, partial correction): ",
      sum(na_counts > 0), " (", sprintf("%.1f%%", 100 * mean(na_counts > 0)), ")",
      " — expected under DIA missingness, not an error")
}

stopifnot(identical(dim(corrected), dim(x)))
stopifnot(identical(is.na(corrected), is.na(x)))
say("  missingness pattern unchanged (nothing imputed)")
```

# 5. Split and write

**Every injection is written, references included**, in the input file's column order, with the
annotation columns re-attached **unchanged**. The output filename carries a descriptive suffix so the
input can never be overwritten.

The three pooled `MIX_*` injections are kept deliberately: they are the technical control, and having
them in the same file means the reference collapse can be re-checked downstream in Python instead of
being taken on trust from section 6. They are corrected exactly like every other column — they carry
their own `reference` level in `bio_group`, so the correction never pooled them with a cell line.

⚠️ **They are not biological samples: drop them before clustering, PCA or any per-cell-line
statistic.** They belong to no cell line and no timepoint, so anything that groups by those fields
will either mis-assign them or silently create a spurious group. Select on the `MIX` cell-line token
(or re-read `is_reference` from `batch_info.tsv`) and exclude them at load time.

**On `OUTPUT_SCALE`.** The correction is done on log2 either way; this only decides what lands on
disk. With the default `"log2"` the sample columns are renamed `raw:abs` → `log2:abs`, because a
column called `raw:abs` that holds log2 values is a trap for anything reading the file later. Set
`OUTPUT_SCALE <- "linear"` to back-transform with `2^x` instead, which is what
`run_diapasef_transformations()` in `src/transformations.py` expects — it wants `raw:abs` linear
intensities and does its own log2.

```{r write-matrix}
say("=== 5. Split and write ===")

ref_cols <- sheet$sample_id[sheet$is_reference]
bio_cols <- sheet$sample_id[!sheet$is_reference]

# Columns to write: EVERY injection, references included, in the INPUT FILE's order.
# `sheet$sample_id` is in injection order, so writing it as-is would silently reshuffle the
# header relative to the input; `intersect()` keeps the order of its first argument, so this
# restores the file's own layout. Nothing else is affected — the matrix is indexed by name
# everywhere, so the ordering is presentation only.
out_cols <- intersect(colnames(abundance), sheet$sample_id)
stopifnot(setequal(out_cols, sheet$sample_id))

if (length(ref_cols) < 2L) {
  warning("write: fewer than 2 reference injections found — the reference QC below will ",
          "be uninformative.", call. = FALSE)
}
say("  writing ", length(out_cols), " injections: ", length(bio_cols),
    " biological + ", length(ref_cols), " pooled reference(s) kept as controls",
    " — drop the references before clustering")

out_matrix <- corrected[, out_cols, drop = FALSE]

out_names <- out_cols
if (identical(OUTPUT_SCALE, "linear")) {
  say("  OUTPUT_SCALE = 'linear': back-transforming with 2^x (correction was done on log2)")
  out_matrix <- 2 ^ out_matrix
} else if (identical(OUTPUT_SCALE, "log2")) {
  if (scaled$was_linear) {
    out_names <- sub(RENAME_DATA_TYPE[["from"]], RENAME_DATA_TYPE[["to"]],
                     out_names, fixed = TRUE)
    say("  OUTPUT_SCALE = 'log2': DataType token renamed '",
        RENAME_DATA_TYPE[["from"]], "' -> '", RENAME_DATA_TYPE[["to"]], "' in ",
        sum(out_names != out_cols), " column name(s)")
  }
} else {
  stop("OUTPUT_SCALE must be 'log2' or 'linear', got '", OUTPUT_SCALE, "'.")
}

out_df <- cbind(abundance[, annot_cols, drop = FALSE],
                as.data.frame(out_matrix, check.names = FALSE))
colnames(out_df) <- c(annot_cols, out_names)

matrix_path <- output_path(ABUNDANCE_FILE, OUT_SUFFIX_MATRIX,
                           out_dir = OUT_DIR, overwrite = OVERWRITE)
utils::write.table(out_df, matrix_path,
                   sep = "\t", quote = FALSE, row.names = FALSE, na = "")
say("  wrote clustering matrix: ", matrix_path,
    "  (", nrow(out_df), " sites x ", ncol(out_df), " columns, ",
    round(file.size(matrix_path) / 1e6), " MB)")
```

# 6. Reference QC — did the correction work?

The three pooled references (`mix`, `mixb`, `mixc`) are the **same physical material** injected once
per batch. Any spread across them is therefore purely technical, which makes them the validation:
if the batch offsets are real and were estimated well, the references must collapse toward each other
after correction.

**Read it as:** median SD across sites should drop substantially, and the scatter cloud should sit
below the identity line. It will not drop to zero — what remains is the ordinary run-to-run
measurement noise the correction was never trying to remove.

**The caveat that makes this near-independent rather than circular.** The references are 3 of ~219
injections, so they contribute almost nothing to the batch estimate — that estimate is driven by the
~216 biological samples. The correction is therefore not "fitted to" the references in any meaningful
sense, and their convergence is close to out-of-sample evidence. It is still only a spot-check: only
sites quantified in **all three** references are evaluable, and that subset is biased toward abundant
sites.

If the references do **not** collapse, the chunk warns. That means either the batch effect is not an
additive shift (in which case this whole approach is the wrong one, not just badly tuned) or the
`batch` assignment in the sheet is wrong. Do not use the corrected matrix until that is understood.

```{r reference-qc, fig.width = 11, fig.height = 5}
say("=== 6. Reference QC ===")

if (length(ref_cols) >= 2L) {
  before <- reference_spread(x, ref_cols)
  after  <- reference_spread(corrected, ref_cols)

  evaluable <- !is.na(before$sd)
  say("  evaluable sites (quantified in all ", length(ref_cols), " references): ",
      sum(evaluable), " / ", nrow(x),
      " (", sprintf("%.1f%%", 100 * mean(evaluable)), ")")

  if (sum(evaluable) > 0L) {
    med_sd_before <- stats::median(before$sd[evaluable])
    med_sd_after  <- stats::median(after$sd[evaluable])
    med_cv_before <- stats::median(before$cv[evaluable], na.rm = TRUE)
    med_cv_after  <- stats::median(after$cv[evaluable], na.rm = TRUE)

    say("  median SD across references (log2):  ",
        sprintf("%.4f -> %.4f  (%+.1f%%)", med_sd_before, med_sd_after,
                100 * (med_sd_after / med_sd_before - 1)))
    say("  median CV across references (log2 units, %): ",
        sprintf("%.3f -> %.3f", med_cv_before, med_cv_after))
    say("  sites whose spread shrank: ",
        sprintf("%.1f%%", 100 * mean(after$sd[evaluable] < before$sd[evaluable])))

    if (med_sd_after >= med_sd_before) {
      warning("Reference QC: the references did NOT collapse (median SD ",
              sprintf("%.4f -> %.4f", med_sd_before, med_sd_after),
              "). Either the batch effect is not an additive shift, or the sample sheet's ",
              "`batch` assignment is wrong. Do not use the corrected matrix until this is ",
              "understood.", call. = FALSE)
    }

    qc_df <- data.frame(site = rownames(x),
                        n_refs_observed = before$n_obs,
                        sd_before = before$sd,
                        sd_after = after$sd,
                        cv_before = before$cv,
                        cv_after = after$cv,
                        stringsAsFactors = FALSE)
    qc_path <- output_path(ABUNDANCE_FILE, OUT_SUFFIX_REF_QC,
                           out_dir = OUT_DIR, overwrite = OVERWRITE)
    utils::write.table(qc_df, qc_path,
                       sep = "\t", quote = FALSE, row.names = FALSE, na = "")
    say("  wrote reference QC table: ", qc_path)

    png_path <- output_path(ABUNDANCE_FILE, OUT_SUFFIX_REF_PNG,
                            out_dir = OUT_DIR, overwrite = OVERWRITE)
    plot_reference_qc(before, after, png_path)          # saved copy
    plot_reference_qc(before, after)                    # inline, below
    say("  wrote reference QC figure: ", png_path)
  } else {
    warning("Reference QC: no site is quantified in every reference — QC skipped.",
            call. = FALSE)
  }
} else {
  say("  skipped: fewer than 2 reference injections in the sheet")
}
```

# Outputs, and what may be done with them

| file | contents | consumers |
|---|---|---|
| `..._batchcorrected_linear.tsv` | annotation columns unchanged + **all 219 corrected injections** (216 biological + 3 `MIX_*` references), in input-file column order, back-transformed to the linear scale | clustering, PCA / UMAP, profile and curve plots — **after dropping the `MIX_*` columns** |
| `..._batchcorrected_linear_refQC.tsv` | per-site reference spread before and after (`sd`, `cv`, `n_refs_observed`) | QC only |
| `..._batchcorrected_linear_refQC.png` | the two-panel diagnostic above | QC only |

Reading it in Python:

```python
import pandas as pd
df = pd.read_csv(path, sep="\t", low_memory=False)   # never .fillna(0)
```

Two rules carry over to whatever reads this file:

- **Missing values are `NA` and must stay `NaN`.** They are non-detections, not zeros. This applies
  with extra force to any table this is later merged with that carries limma statistics columns — a
  `FFDR` of `NaN` means *not tested*, and `0.0` reads as *perfectly significant*.
- ⚠ **Do not test on this matrix.** Statistical testing stays on the **uncorrected** values with
  `batch` in the design — see *The two tracks* at the top. This file exists so that a PCA shows
  biology instead of acquisition order; it does not exist to make p-values smaller.

# Expected warnings

`Partial NA coefficients for N probe(s)` — emitted by `lmFit` inside `removeBatchEffect`. It fires
whenever *any* coefficient is non-estimable for a site, and under DIA missingness that is usually a
**biological** coefficient (a site never observed in one cell-line/timepoint group), which is
harmless: that group simply has no fitted mean, and nothing is subtracted on its behalf. It does
**not** mean the batch correction failed. The `sites with a non-estimable batch coefficient` line in
the *Correct* chunk is the diagnostic that actually answers that question, and it is usually a much
smaller number — often zero.

```{r session-info}
sessionInfo()
```
