---
title: "limma — p-values of each timepoint against the starve control"
output: html_notebook
---

# Purpose

Compute, for every phosphosite, the statistical significance of each stimulation timepoint
against the **starve** control, using the `log2:abs` data produced by
`TMT_dataset_preprocessing.ipynb` / `src/transformations.py`.

This replaces the Welch t-test / Benjamini-Hochberg functions that used to live in
`src/transformations.py` (`compute_pvalues`, `compute_fdr`, `compute_log10_fdr`, now commented
out there). With only 4 replicates the per-site variance estimate has 3 degrees of freedom and
is very unstable; **limma** shrinks each site's variance toward a prior fitted across all sites
(Smyth 2004), which is the whole reason for moving the statistics to R.

Three families of results are produced:

1. **Per timepoint × condition** — moderated *t* contrast against starve: log fold change,
   p-value, BH-FDR and adjusted p-value (`-log10(FDR)`).
2. **Per condition** — omnibus moderated *F* across that condition's stimulation timepoints:
   *"is this site EGF-responsive at all?"*
3. **Global, per site** — omnibus moderated *F* across every timepoint of every condition:
   *"does this site respond to anything, at any time?"*

**The data is already log2-transformed** (`log2:abs`). No further transformation is applied —
`lmFit` expects data on the log scale, which is exactly what it gets here.

# The design: why the plex has to be in the model

Each replicate `r1 … r4` **is one TMT plex**. Each plex contains 17 real channels —
`full`, `starve`, and 5 stimulation timepoints for each of EGF / INS / EGFnINS — so all
comparisons of interest live *inside* a single plex.

The requirement is that each timepoint is compared to the starve control **of its own
replicate**, never to another replicate's. In limma this is the standard paired/blocked design
(limma User's Guide §9.4):

```
~ 0 + group + plex
```

The `plex` term absorbs each plex's overall offset, so every `group − starve` contrast is
estimated *within* plex. The residual variance is then pooled across plexes and shrunk across
sites by `eBayes`. This is strictly better than subtracting the starve channel by hand and
testing the ratios, because the pooled fit keeps the degrees of freedom that hand-made ratios
throw away.

## A structural detail of these files: the controls are duplicated

There are 84 `log2:abs` columns, but only **68 real samples**. `full` and `starve` are
physically one channel per plex, yet they appear in the table three times — once under each
condition arm (`..._EGF_starve_r1`, `..._INS_starve_r1`, `..._EGFnINS_starve_r1` hold identical
values).

Feeding all 84 columns to `lmFit` would enter the starve control **three times**, which is
pseudo-replication: it would inflate the residual degrees of freedom and shrink the variance
estimate, making everything look more significant than it is. The code below therefore keeps
only one copy of each control and *asserts* that the discarded copies were identical.

# Caveat: no between-sample normalisation is applied

This notebook fits `log2:abs` **exactly as stored on disk** (`NORMALIZE <- "none"`), which is a
deliberate choice, not an oversight.

The consequence should be stated plainly: the 68 channel medians span about **1.58 log2** in
`hme1_2`, and still span 0.49–1.11 log2 *within* a single plex. The `plex` term removes the
plex-level offset but **not** per-channel loading differences, so each contrast inherits a
systematic shift in its log fold change. This is the defect documented in
`notebooks/03_clustering/clustering_method_decision.md` §1 (median `log2:FC` of +0.566 across
all sites at EGF 5 min). It biases the *point estimate*, not just the variance.

`NORMALIZE` is left as a parameter so median-centring or quantile normalisation can be switched
on later without touching any of the analysis code.

# Parameters

Paths are resolved from the **project root**, which is located by walking up from the current
working directory until a folder containing `data/`, `src/` and `notebooks/` is found. Relative
paths like `../../data/...` cannot be used here: knitting the notebook runs with this file's
folder as the working directory, but running the chunks from the PyCharm / RStudio console runs
with the project root, and the same relative path cannot mean both.

Each dataset keeps its own folder, `Experiment/{dataset}/Data/Processed/`, so only the file
*name* is listed per dataset and the folder is built from the dataset key. Outputs go back into
that same folder — one limma table per experiment rather than a flat pile in `Experiment/` —
and are named after the input's stem, so the date prefix carries over
(`20260807_hTERT_HEM1_2_log2_processed.tsv` → `20260807_hTERT_HEM1_2_limma_pvalues.tsv`).

```{r setup}
suppressPackageStartupMessages(library(limma))

find_project_root <- function(start = getwd(),
                              markers = c("data", "src", "notebooks")) {
  # Locate the project root by walking up the directory tree from `start`.
  #
  # Makes the notebook runnable from any working directory — knitted (working directory =
  # this file's folder) or run chunk-by-chunk from a console rooted at the project.
  #
  # Args:
  #   start: directory to start searching from, default the current working directory.
  #   markers: directory names that must all be present for a folder to count as the root.
  #
  # Returns:
  #   Absolute path to the project root. Raises an error if no ancestor directory qualifies.

  path <- normalizePath(start, mustWork = TRUE)

  repeat {
    if (all(dir.exists(file.path(path, markers)))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("find_project_root: no ancestor of '", start, "' contains all of: ",
           paste(markers, collapse = ", "),
           ". Set PROJECT_ROOT manually to the TMT_Data_analysis folder.")
    }
    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()
DATA_DIR     <- file.path(PROJECT_ROOT, "Experiment")

# Each dataset lives in its own experiment folder, with the processed input under
# Experiment/{dataset}/Data/Processed/. Only the file name has to be listed here; the folder is
# built from the dataset key, so nothing in this notebook holds an absolute path.
PROCESSED_SUBDIR <- file.path("Data", "Processed")

DATASET_FILES <- list(hek_1  = "20260807_HEK293T_log2_processed.tsv",
                      hme1_1 = "20260807_hTERT_HEM1_1_log2_processed.tsv",
                      hme1_2 = "20260807_hTERT_HEM1_2_log2_processed.tsv")

DATA_FILES <- vapply(names(DATASET_FILES),
                     function(dataset_name) {
                       file.path(DATA_DIR,
                                 dataset_name,
                                 PROCESSED_SUBDIR,
                                 DATASET_FILES[[dataset_name]])
                     },
                     character(1))

CELL_LINE  <- "WT"          # column prefix of the cell line to analyse
CONTROL    <- "starve"      # reference timepoint for every contrast
CONDITIONS <- c("EGF", "INS", "EGFnINS")
MIN_PLEX   <- 2             # minimum number of plexes a site must be detected in (see below)
NORMALIZE  <- "none"        # "none" | "median" | "quantile"
OUT_SUFFIX <- "_limma_pvalues.tsv"

# --- debugging switches ---------------------------------------------------------------
# MAX_SITES caps the number of sites fitted, so the whole notebook can be smoke-tested in
# seconds before committing to a full run. Set to Inf for the real analysis.
MAX_SITES  <- Inf           # e.g. 2000 to time the pipeline end-to-end first
VERBOSE    <- TRUE          # per-stage timing checkpoints


limma_output_path <- function(dataset_name,
                              data_files = DATA_FILES,
                              out_suffix = OUT_SUFFIX) {
  # Build the output path of one dataset's limma table, next to that dataset's input file.
  #
  # Results are written into the same Experiment/{dataset}/Data/Processed/ folder the input was
  # read from, so each experiment folder stays self-contained. The name reuses the input's stem
  # (date + dataset), so the date prefix of the processed file carries over to the statistics.
  #
  # Args:
  #   dataset_name: dataset key, e.g. "hme1_2"; must be a name of `data_files`.
  #   data_files: named vector of input file paths.
  #   out_suffix: suffix appended to the input stem, including the extension.
  #
  # Returns:
  #   Absolute path of the .tsv to write. Never equal to the input path.

  input_path <- data_files[[dataset_name]]
  stem <- sub("_log2_processed\\.tsv$", "", basename(input_path))
  stem <- sub("\\.tsv$", "", stem)          # fallback if the input is named differently

  output_path <- file.path(dirname(input_path), paste0(stem, out_suffix))
  if (!dir.exists(dirname(output_path))) {
    stop("limma_output_path: output folder '", dirname(output_path), "' does not exist.")
  }
  if (identical(output_path, input_path)) {
    stop("limma_output_path: refusing to overwrite the input file '", input_path, "'.")
  }
  output_path
}

cat("limma version:", as.character(packageVersion("limma")), "\n")
cat("working directory:", getwd(), "\n")
cat("project root     :", PROJECT_ROOT, "\n\n")

missing_files <- DATA_FILES[!file.exists(DATA_FILES)]
if (length(missing_files) > 0) {
  stop("input file(s) not found:\n  ", paste(missing_files, collapse = "\n  "))
}
for (dataset_name in names(DATA_FILES)) {
  cat(sprintf("  %-8s in : %s\n", dataset_name, DATA_FILES[[dataset_name]]))
  cat(sprintf("  %-8s out: %s\n", "",           limma_output_path(dataset_name)))
}
```

## Why `MIN_PLEX = 2`

Missingness in TMT is **per plex, not per channel**: a peptide is either selected for MS2 in a
given run — yielding reporter ions in all 17 channels — or it is not. So a site is fully
present or fully absent within each plex, and `n:reps` is exactly the number of plexes it was
detected in.

A site detected in **one** plex contributes 17 observations to a model with 17 estimable
parameters, giving `df.residual = 0`. `eBayes` will still happily return a p-value for it,
computed entirely from the prior variance with *zero* site-specific evidence. Those numbers are
not defensible, so such sites are dropped. Raise `MIN_PLEX` to 3 or 4 for a stricter analysis.

# Progress checkpoints

Every stage announces itself **before** it runs and reports its duration after, and each line is
flushed immediately. So if the run stalls, the **last line printed is the stage it is stuck in** —
which is the whole point: a stage that never prints its `done` line is the culprit.

`report_memory()` is included because the most likely cause of a job that should take seconds
taking an hour is not arithmetic but **swapping**. If `used` climbs into the GB range, the machine
is paging and every subsequent operation is disk-bound.

`muffle_warnings()` addresses the other classic cause. `lmFit` can emit one warning *per site*, and
while `Rscript` shrugs that off, an IDE notebook re-rendering tens of thousands of warning lines can
take orders of magnitude longer than the computation itself. Warnings are counted and summarised
instead of streamed.

```{r checkpoints}
.timing <- new.env(parent = emptyenv())

checkpoint <- function(label,
                       reset = FALSE) {
  # Print a flushed, timestamped progress line with the time since the previous checkpoint.
  #
  # Args:
  #   label: description of the stage being reported.
  #   reset: TRUE to restart the total-elapsed clock (use at the start of a dataset).
  #
  # Returns:
  #   Invisibly, seconds elapsed since the previous checkpoint.

  if (!VERBOSE) {
    return(invisible(0))
  }

  now <- proc.time()[["elapsed"]]
  if (reset || is.null(.timing$start)) {
    .timing$start <- now
    .timing$last  <- now
  }
  since_last  <- now - .timing$last
  since_start <- now - .timing$start
  .timing$last <- now

  cat(sprintf("[%s | +%7.1fs | %7.1fs total] %s\n",
              format(Sys.time(), "%H:%M:%S"),
              since_last,
              since_start,
              label))
  flush.console()
  invisible(since_last)
}


stage <- function(label,
                  expr) {
  # Run one stage, announcing it before evaluation and timing it after.
  #
  # `expr` is passed unevaluated and forced inside, so the "starting" line is guaranteed to be
  # printed and flushed before any work begins. A stage that hangs therefore leaves its own name
  # as the last thing on screen.
  #
  # Args:
  #   label: short description of the stage, e.g. "lmFit".
  #   expr: the expression to evaluate.
  #
  # Returns:
  #   The value of `expr`.

  if (VERBOSE) {
    cat(sprintf("[%s]   -> %-42s ", format(Sys.time(), "%H:%M:%S"), paste0(label, " ...")))
    flush.console()
  }

  started <- proc.time()[["elapsed"]]
  value   <- force(expr)
  elapsed <- proc.time()[["elapsed"]] - started

  if (VERBOSE) {
    cat(sprintf("done in %7.1fs\n", elapsed))
    flush.console()
  }
  value
}


report_memory <- function(label = "") {
  # Print the memory currently used by R, to expose swapping as a cause of a slow run.
  #
  # Args:
  #   label: optional tag printed alongside the figures.
  #
  # Returns:
  #   Invisibly, total megabytes in use.

  if (!VERBOSE) {
    return(invisible(0))
  }

  # gc() returns a variable number of columns — a "limit (Mb)" column appears on some platforms,
  # so the megabyte columns must be located by name rather than by position. The first "(Mb)"
  # column is current usage, the last is the peak since the session started.
  usage   <- gc(verbose = FALSE)
  mb_cols <- which(colnames(usage) == "(Mb)")
  used    <- sum(usage[, mb_cols[1]])
  peak    <- if (length(mb_cols) > 1) sum(usage[, mb_cols[length(mb_cols)]]) else NA_real_

  cat(sprintf("[%s]      memory: %6.0f MB in use%s  %s\n",
              format(Sys.time(), "%H:%M:%S"),
              used,
              if (is.na(peak)) "" else sprintf(", %6.0f MB peak", peak),
              label))
  flush.console()
  invisible(used)
}


muffle_warnings <- function(expr,
                            label = "stage") {
  # Evaluate an expression, collecting warnings instead of emitting them one by one.
  #
  # limma emits one warning per problematic site; tens of thousands of them can dominate the
  # runtime of an IDE notebook even though the computation itself is fast. This counts them and
  # prints a summary of the distinct messages.
  #
  # Args:
  #   expr: expression to evaluate.
  #   label: tag used in the printed summary.
  #
  # Returns:
  #   The value of `expr`.

  collected <- character(0)

  value <- withCallingHandlers(
    expr,
    warning = function(w) {
      collected <<- c(collected, conditionMessage(w))
      invokeRestart("muffleWarning")
    })

  if (length(collected) > 0 && VERBOSE) {
    counts <- sort(table(collected), decreasing = TRUE)
    cat(sprintf("[%s]      %d warning(s) in %s:\n",
                format(Sys.time(), "%H:%M:%S"), length(collected), label))
    for (i in seq_len(min(5, length(counts)))) {
      cat(sprintf("           %6d x  %s\n", counts[[i]], names(counts)[i]))
    }
    flush.console()
  }
  value
}
```

# Functions

```{r functions}
parse_abs_columns <- function(column_names,
                              cell_line,
                              conditions) {
  # Parse the log2:abs column names of one cell line into their component fields.
  #
  # Args:
  #   column_names: character vector of all column names in the dataset.
  #   cell_line: cell line prefix to keep, e.g. "WT".
  #   conditions: character vector of condition names, e.g. c("EGF", "INS", "EGFnINS").
  #
  # Returns:
  #   data.frame with one row per log2:abs column and the columns
  #   `column`, `condition`, `timepoint`, `plex`.

  pattern <- paste0("^", cell_line, "_log2:abs_(",
                    paste(conditions, collapse = "|"),
                    ")_([^_]+)_(r[0-9]+)$")

  matched <- grep(pattern, column_names, value = TRUE)
  if (length(matched) == 0) {
    stop("parse_abs_columns: no log2:abs columns found for cell line '", cell_line, "'.")
  }

  fields <- regmatches(matched, regexec(pattern, matched))

  data.frame(column    = matched,
             condition = vapply(fields, `[`, character(1), 2),
             timepoint = vapply(fields, `[`, character(1), 3),
             plex      = vapply(fields, `[`, character(1), 4),
             stringsAsFactors = FALSE)
}


deduplicate_controls <- function(df,
                                 column_info,
                                 conditions,
                                 control_timepoints = c("full", "starve")) {
  # Remove the duplicated control columns that appear once per condition arm.
  #
  # `full` and `starve` are a single physical channel per plex but are written out under every
  # condition. Only the copy belonging to the first condition is kept, and the discarded copies
  # are checked to be identical to it first — if a future dataset stops duplicating them, this
  # raises rather than silently averaging different samples.
  #
  # Args:
  #   df: the full dataset, used only to compare the duplicated columns.
  #   column_info: data.frame returned by parse_abs_columns().
  #   conditions: character vector of condition names; the first one is kept.
  #   control_timepoints: timepoints that are shared across condition arms.
  #
  # Returns:
  #   column_info restricted to unique samples, with an added `group` column giving the
  #   experimental group of each sample (control timepoints keep their bare name).

  keep_condition <- conditions[1]

  for (timepoint in control_timepoints) {
    for (plex in unique(column_info$plex)) {
      reference <- column_info$column[column_info$condition == keep_condition &
                                      column_info$timepoint == timepoint &
                                      column_info$plex      == plex]
      if (length(reference) != 1) next

      for (other_condition in conditions[-1]) {
        duplicate <- column_info$column[column_info$condition == other_condition &
                                        column_info$timepoint == timepoint &
                                        column_info$plex      == plex]
        if (length(duplicate) != 1) next

        if (!isTRUE(all.equal(df[[reference]], df[[duplicate]]))) {
          stop("deduplicate_controls: '", duplicate, "' is not identical to '", reference,
               "'. The control channels are not duplicates in this dataset — ",
               "the design must be rebuilt before fitting.")
        }
      }
    }
  }

  is_duplicate_control <- column_info$timepoint %in% control_timepoints &
                          column_info$condition != keep_condition
  unique_info <- column_info[!is_duplicate_control, , drop = FALSE]

  unique_info$group <- ifelse(unique_info$timepoint %in% control_timepoints,
                              unique_info$timepoint,
                              paste(unique_info$condition, unique_info$timepoint, sep = "_"))

  unique_info
}


sort_timepoints <- function(timepoints) {
  # Sort timepoint labels in biological order: named controls first, then numerically.
  #
  # Args:
  #   timepoints: character vector of timepoint labels, e.g. c("10", "full", "2", "starve").
  #
  # Returns:
  #   The same labels, ordered with "full" and "starve" first and the numeric points ascending.

  named   <- c("full", "starve")
  present <- named[named %in% timepoints]
  numeric_points <- setdiff(timepoints, named)
  c(present, numeric_points[order(as.numeric(numeric_points))])
}


normalize_matrix <- function(M,
                             method) {
  # Apply between-sample normalisation to the expression matrix.
  #
  # Args:
  #   M: numeric matrix, sites in rows and samples in columns, on the log2 scale.
  #   method: "none" (return unchanged), "median" (centre every column on the grand median),
  #     or "quantile" (limma::normalizeBetweenArrays).
  #
  # Returns:
  #   The normalised matrix, same dimensions and dimnames as the input.

  if (method == "none") {
    return(M)
  }
  if (method == "median") {
    column_medians <- apply(M, 2, median, na.rm = TRUE)
    return(sweep(M, 2, column_medians) + mean(column_medians))
  }
  if (method == "quantile") {
    return(normalizeBetweenArrays(M, method = "quantile"))
  }
  stop("normalize_matrix: unknown method '", method, "'.")
}


build_contrasts <- function(groups,
                            control,
                            design) {
  # Build the contrast matrix of every group against the control group.
  #
  # Args:
  #   groups: character vector of group names to contrast against the control.
  #   control: name of the control group, e.g. "starve".
  #   design: design matrix whose column names define the coefficient names.
  #
  # Returns:
  #   Contrast matrix with one column per group, named after the group.

  expressions <- paste0("group", groups, " - group", control)
  contrast_matrix <- makeContrasts(contrasts = expressions, levels = design)
  colnames(contrast_matrix) <- groups
  contrast_matrix
}


order_by_datatype <- function(results,
                              cell_line,
                              labels,
                              f_names) {
  # Reorder the result columns so each data type forms one contiguous block.
  #
  # The natural assembly order is per contrast (all four statistics of EGF_2, then all four of
  # EGF_5, ...). This flips it to per data type: every limmaFC column with its timepoints in
  # order, then every pvalue column, then FDR, then adjustedFDR, then the F-test blocks.
  #
  # Args:
  #   results: assembled results data.frame, keyed by `site`.
  #   cell_line: cell line column prefix, e.g. "WT".
  #   labels: contrast labels ("EGF_full", "EGF_2", ...) in the desired timepoint order.
  #   f_names: names of the F-tests, e.g. c("EGF", "INS", "EGFnINS", "ALL").
  #
  # Returns:
  #   The same data.frame with its columns reordered. Raises if the reordering would drop or
  #   invent a column.

  contrast_stats <- c("limmaFC", "pvalue", "FDR", "adjustedFDR")
  f_stats        <- c("Fpvalue", "FFDR", "adjustedFFDR")

  ordered <- "site"
  for (statistic in contrast_stats) {
    ordered <- c(ordered, paste0(cell_line, "_log2:", statistic, "_", labels))
  }
  for (statistic in f_stats) {
    ordered <- c(ordered, paste0(cell_line, "_log2:", statistic, "_", f_names, "_omnibus"))
  }

  if (!setequal(ordered, colnames(results))) {
    stop("order_by_datatype: the reordered column list does not match the assembled one.")
  }

  results[, ordered, drop = FALSE]
}


add_test_columns <- function(results,
                             p_values,
                             prefix) {
  # Append p-value, BH-FDR and adjusted p-value columns for one test to the results frame.
  #
  # The adjusted p-value is -log10(FDR), matching the project's `adjustedFDR` convention.
  # An FDR of exactly 0 would give +Inf and is set to NA instead.
  #
  # Args:
  #   results: data.frame being assembled, one row per site.
  #   p_values: numeric vector of raw p-values, in the row order of `results`.
  #   prefix: column-name stem, e.g. "WT_log2:pvalue_EGF_5" without the statistic part;
  #     the three columns are built from the `pvalue` / `FDR` / `adjustedFDR` variants.
  #
  # Returns:
  #   The results data.frame with three new columns appended.

  fdr <- p.adjust(p_values, method = "BH")
  adjusted <- -log10(fdr)
  adjusted[is.infinite(adjusted)] <- NA_real_

  results[[prefix$pvalue]]      <- p_values
  results[[prefix$fdr]]         <- fdr
  results[[prefix$adjustedfdr]] <- adjusted
  results
}
```

# The analysis

```{r read-dataset}
read_dataset <- function(file_path,
                         dataset_name) {
  # Read one processed dataset and report everything needed to explain a slow or wrong run.
  #
  # The file is read once per dataset and reused by the fit and both verification steps, rather
  # than being re-read three times. The diagnostics printed here answer, in order: is the file
  # the size we think it is, did it parse into the expected shape, are the intensity columns
  # numeric, and are missing values stored as NA rather than as 0 (see the note below).
  #
  # Args:
  #   file_path: path to the processed .tsv.
  #   dataset_name: dataset key, used in the printed report.
  #
  # Returns:
  #   The dataset as a data.frame, with column names untouched.

  size_mb <- file.size(file_path) / 1e6
  cat(sprintf("[%s]      file: %.0f MB  %s\n",
              format(Sys.time(), "%H:%M:%S"), size_mb, basename(file_path)))
  flush.console()

  df <- stage(sprintf("read.delim (%.0f MB)", size_mb),
              read.delim(file_path,
                         sep = "\t",
                         check.names = FALSE,
                         na.strings = c("NA", "NaN", ""),
                         stringsAsFactors = FALSE))

  cat(sprintf("[%s]      parsed: %d rows x %d columns\n",
              format(Sys.time(), "%H:%M:%S"), nrow(df), ncol(df)))

  abs_columns <- grep("_log2:abs_", colnames(df), value = TRUE)
  non_numeric <- abs_columns[!vapply(df[abs_columns], is.numeric, logical(1))]
  if (length(non_numeric) > 0) {
    cat(sprintf("      WARNING: %d log2:abs column(s) did not parse as numeric, e.g. %s\n",
                length(non_numeric), paste(head(non_numeric, 3), collapse = ", ")))
  }

  # Missingness must be NA. If a fillna(0) reached this file, the zeros are read as real
  # intensities: limma then sees a complete matrix, takes its fast path, and returns confident
  # nonsense. This is the defect recorded in clustering_method_decision.md §1.
  numeric_abs <- abs_columns[vapply(df[abs_columns], is.numeric, logical(1))]
  if (length(numeric_abs) > 0) {
    values    <- as.matrix(df[numeric_abs])
    n_missing <- sum(is.na(values))
    n_zero    <- sum(values == 0, na.rm = TRUE)
    cat(sprintf("      log2:abs cells: %.1f%% NA, %.1f%% exactly zero\n",
                100 * n_missing / length(values),
                100 * n_zero / length(values)))
    if (n_zero > 0.01 * length(values)) {
      cat("      WARNING: many exact zeros in log2:abs — missing values may have been\n",
          "               written as 0 instead of NA. Fix before trusting any p-value.\n", sep = "")
    }
  }

  report_memory(sprintf("after reading %s", dataset_name))
  flush.console()
  df
}
```

```{r run-limma}
run_limma_dataset <- function(df,
                              dataset_name,
                              cell_line = CELL_LINE,
                              control = CONTROL,
                              conditions = CONDITIONS,
                              min_plex = MIN_PLEX,
                              normalize = NORMALIZE,
                              max_sites = MAX_SITES) {
  # Fit the blocked limma model on one dataset and return the full table of statistics.
  #
  # Every stimulation timepoint is contrasted against the control timepoint within its own
  # plex, using the design ~ 0 + group + plex. Three sets of statistics are produced: the
  # moderated t contrast per timepoint, an omnibus moderated F per condition, and one global
  # omnibus F per site across all conditions.
  #
  # Args:
  #   df: the dataset as returned by read_dataset() — read once and reused, so the file is not
  #     parsed again here.
  #   dataset_name: short name of the dataset, used in the printed report.
  #   cell_line: cell line column prefix to analyse.
  #   control: control timepoint every contrast is taken against.
  #   conditions: condition names present in the column headers.
  #   min_plex: minimum value of `n:reps` for a site to be tested.
  #   normalize: between-sample normalisation method passed to normalize_matrix().
  #   max_sites: cap on the number of sites fitted, for smoke-testing; Inf for a real run.
  #
  # Returns:
  #   data.frame with one row per tested site: the `site` key followed by the logFC / pvalue /
  #   FDR / adjustedFDR columns per timepoint and the F-test columns per condition and global.

  cat("\n=====", dataset_name, "=====\n")
  checkpoint(paste("fitting", dataset_name), reset = TRUE)

  stopifnot(!anyDuplicated(df$site))
  stopifnot("n:reps" %in% colnames(df))

  # --- samples -------------------------------------------------------------------------
  column_info <- stage("parse log2:abs column names",
                       parse_abs_columns(colnames(df), cell_line, conditions))
  cat("     log2:abs columns:", nrow(column_info), "\n")

  samples <- stage("de-duplicate control channels",
                   deduplicate_controls(df, column_info, conditions))
  cat("     unique sample channels after de-duplicating the controls:", nrow(samples), "\n")

  timepoints <- sort_timepoints(unique(column_info$timepoint))
  cat("     timepoints:", paste(timepoints, collapse = ", "), "\n")

  group_levels <- c(control,
                    setdiff(c("full", "starve"), control),
                    unlist(lapply(conditions, function(condition) {
                      paste(condition,
                            setdiff(timepoints, c("full", "starve")),
                            sep = "_")
                    })))
  group_levels <- group_levels[group_levels %in% samples$group]

  samples$group <- factor(samples$group, levels = group_levels)
  samples$plex  <- factor(samples$plex)

  # --- expression matrix ---------------------------------------------------------------
  tested <- df[["n:reps"]] >= min_plex
  tested[is.na(tested)] <- FALSE
  if (is.finite(max_sites) && sum(tested) > max_sites) {
    tested[which(tested)[-seq_len(max_sites)]] <- FALSE
    cat("     NOTE: MAX_SITES is set — only", max_sites, "sites will be fitted.\n")
  }
  cat("     sites tested (n:reps >=", min_plex, "):", sum(tested), "of", nrow(df),
      " | dropped:", sum(!tested), "\n")

  M <- stage("build expression matrix",
             as.matrix(df[tested, samples$column, drop = FALSE]))
  rownames(M) <- df$site[tested]
  storage.mode(M) <- "double"
  M <- stage(paste0("normalise (", normalize, ")"), normalize_matrix(M, normalize))
  cat("     matrix:", nrow(M), "sites x", ncol(M), "samples | ",
      sprintf("%.1f%% NA", 100 * sum(is.na(M)) / length(M)), "\n")

  # Missingness is supposed to be plex-wise: within one plex a site is either present in all 17
  # channels or absent from all of them. A "partial" plex block breaks the premise behind
  # MIN_PLEX and means n:reps no longer equals the number of plexes the site was seen in.
  # (The number of distinct patterns is not the right check — with 4 plexes and n:reps in 2..4
  # there are legitimately C(4,2)+C(4,3)+C(4,4) = 11 of them.)
  partial_blocks <- 0
  for (plex_level in levels(samples$plex)) {
    block   <- is.na(M[, samples$plex == plex_level, drop = FALSE])
    n_na    <- rowSums(block)
    partial_blocks <- partial_blocks + sum(n_na > 0 & n_na < ncol(block))
  }
  total_blocks <- nrow(M) * nlevels(samples$plex)
  cat(sprintf("     plex blocks partially missing: %d of %d (%.2f%%)%s\n",
              partial_blocks, total_blocks, 100 * partial_blocks / total_blocks,
              if (partial_blocks == 0) " — plex-wise, as expected" else
                " — missingness is NOT purely plex-wise; n:reps may not mean what MIN_PLEX assumes"))

  # --- design and fit ------------------------------------------------------------------
  design <- model.matrix(~ 0 + group + plex, data = samples)
  colnames(design) <- make.names(colnames(design))
  cat("     design:", nrow(design), "x", ncol(design),
      " | rank:", qr(design)$rank, "\n")
  stopifnot(qr(design)$rank == ncol(design))

  cat(sprintf("     lmFit will fit %d sites x %d coefficients%s\n",
              nrow(M), ncol(design),
              if (anyNA(M)) " (per-site loop: NAs present)" else " (single QR: no NAs)"))
  fit <- stage("lmFit", muffle_warnings(lmFit(M, design), "lmFit"))
  report_memory("after lmFit")

  # --- contrasts -----------------------------------------------------------------------
  stimulation_timepoints <- setdiff(timepoints, c("full", "starve"))
  contrast_groups <- setdiff(levels(samples$group), control)

  fit_t <- stage(sprintf("t-contrasts + eBayes (%d contrasts)", length(contrast_groups)),
                 muffle_warnings(
                   eBayes(contrasts.fit(fit,
                                        build_contrasts(contrast_groups, control, design)),
                          trend = TRUE,
                          robust = TRUE),
                   "t-contrasts"))

  f_groups <- list()
  for (condition in conditions) {
    f_groups[[condition]] <- paste(condition, stimulation_timepoints, sep = "_")
  }
  f_groups[["ALL"]] <- unlist(f_groups, use.names = FALSE)

  f_fits <- list()
  for (f_name in names(f_groups)) {
    f_fits[[f_name]] <- stage(sprintf("omnibus F: %s", f_name),
                              muffle_warnings(
                                eBayes(contrasts.fit(fit,
                                                     build_contrasts(f_groups[[f_name]],
                                                                     control,
                                                                     design)),
                                       trend = TRUE,
                                       robust = TRUE),
                                paste("F test", f_name)))
  }
  report_memory("after all fits")

  # --- assemble ------------------------------------------------------------------------
  # `contrast_groups` is already in reporting order: full first, then each condition with its
  # timepoints ascending. `labels` keeps that order and is reused to order the output columns.
  labels <- vapply(contrast_groups,
                   function(group) {
                     if (group %in% timepoints) paste0(conditions[1], "_", group) else group
                   },
                   character(1),
                   USE.NAMES = FALSE)

  results <- stage("assemble result columns", {
    assembled <- data.frame(site = rownames(M), stringsAsFactors = FALSE)

    for (i in seq_along(contrast_groups)) {
      group <- contrast_groups[i]
      label <- labels[i]

      assembled[[paste0(cell_line, "_log2:limmaFC_", label)]] <- fit_t$coefficients[, group]
      assembled <- add_test_columns(assembled,
                                    fit_t$p.value[, group],
                                    list(pvalue      = paste0(cell_line, "_log2:pvalue_", label),
                                         fdr         = paste0(cell_line, "_log2:FDR_", label),
                                         adjustedfdr = paste0(cell_line, "_log2:adjustedFDR_", label)))
    }

    for (name in names(f_fits)) {
      assembled <- add_test_columns(assembled,
                                    f_fits[[name]]$F.p.value,
                                    list(pvalue      = paste0(cell_line, "_log2:Fpvalue_", name, "_omnibus"),
                                         fdr         = paste0(cell_line, "_log2:FFDR_", name, "_omnibus"),
                                         adjustedfdr = paste0(cell_line, "_log2:adjustedFFDR_", name, "_omnibus")))
    }
    assembled
  })

  results <- stage("reorder columns by data type",
                   order_by_datatype(results, cell_line, labels, names(f_fits)))
  checkpoint(sprintf("%s fitted: %d sites x %d columns",
                     dataset_name, nrow(results), ncol(results)))

  # --- report --------------------------------------------------------------------------
  cat("\nsites at FDR < 0.05 per contrast:\n")
  for (label in labels) {
    fdr_col <- paste0(cell_line, "_log2:FDR_", label)
    cat(sprintf("  %-16s %4d / %d\n", label,
                sum(results[[fdr_col]] < 0.05, na.rm = TRUE), nrow(results)))
  }
  cat("sites at F-test FDR < 0.05:\n")
  for (name in names(f_fits)) {
    fdr_col <- paste0(cell_line, "_log2:FFDR_", name, "_omnibus")
    cat(sprintf("  %-16s %4d / %d\n", name,
                sum(results[[fdr_col]] < 0.05, na.rm = TRUE), nrow(results)))
  }

  results
}
```
```{r}

```

The `Partial NA coefficients for N probe(s)` warning below is **expected and benign**. When a
site is absent from a whole plex, that plex's indicator column becomes all-zero in the reduced
design and its coefficient is inestimable — but the missing coefficient is a *plex* nuisance
term, never a group term, so the contrasts remain valid and `df.residual` is correctly reduced
(48 for a 4-plex site, 32 for 3 plexes, 16 for 2 plexes).

Each file is read **once** and the parsed frame is kept in `limma_inputs`, so the fit and both
verification steps below reuse it instead of re-parsing the same TSV three times.

```{r fit-all-datasets}
limma_inputs  <- list()
limma_results <- list()

checkpoint("starting all datasets", reset = TRUE)
for (dataset_name in names(DATA_FILES)) {
  cat("\n===== reading", dataset_name, "=====\n")
  limma_inputs[[dataset_name]]  <- read_dataset(DATA_FILES[[dataset_name]], dataset_name)
  limma_results[[dataset_name]] <- run_limma_dataset(limma_inputs[[dataset_name]], dataset_name)
}
checkpoint("all datasets fitted")
report_memory("end of fitting")
```

# Verification

The critical claim is that each timepoint is compared to the starve control **of its own
replicate**. That is checked numerically rather than asserted: for a site detected in all four
plexes, limma's `logFC` must equal the mean of the four within-plex differences
`x[timepoint, ri] − x[starve, ri]`.

```{r verify-within-plex}
verify_within_plex <- function(df,
                               results,
                               cell_line = CELL_LINE,
                               control = CONTROL,
                               checks = c("EGF_5", "INS_10", "EGFnINS_90")) {
  # Check that limma's logFC equals the mean of the per-plex differences against the control.
  #
  # Args:
  #   df: the dataset that was fitted, as returned by read_dataset() — reused rather than
  #     re-read from disk.
  #   results: results data.frame returned by run_limma_dataset().
  #   cell_line: cell line column prefix.
  #   control: control timepoint used as the contrast reference.
  #   checks: contrasts to verify, each named "{condition}_{timepoint}".
  #
  # Returns:
  #   Invisibly, the maximum absolute discrepancy found across the checked sites.

  complete <- df[df[["n:reps"]] == 4, , drop = FALSE]
  complete <- complete[complete$site %in% results$site, , drop = FALSE]
  if (nrow(complete) == 0) {
    cat("no complete (4-plex) sites available to verify\n")
    return(invisible(NA_real_))
  }

  plexes <- c("r1", "r2", "r3", "r4")
  worst  <- 0

  for (check in checks) {
    parts     <- strsplit(check, "_")[[1]]
    condition <- parts[1]
    timepoint <- parts[2]

    fc_col <- paste0(cell_line, "_log2:limmaFC_", check)
    if (!fc_col %in% colnames(results)) next

    manual <- rowMeans(sapply(plexes, function(plex) {
      complete[[paste0(cell_line, "_log2:abs_", condition, "_", timepoint, "_", plex)]] -
      complete[[paste0(cell_line, "_log2:abs_", condition, "_", control, "_", plex)]]
    }))

    from_limma <- results[[fc_col]][match(complete$site, results$site)]
    discrepancy <- max(abs(manual - from_limma), na.rm = TRUE)
    worst <- max(worst, discrepancy)

    cat(sprintf("  %-12s n=%d  max |limma - within-plex mean| = %.3e\n",
                check, nrow(complete), discrepancy))
  }

  invisible(worst)
}

for (dataset_name in names(DATA_FILES)) {
  cat("\n", dataset_name, "\n", sep = "")
  worst <- stage(paste("verify within-plex contrast:", dataset_name),
                 verify_within_plex(limma_inputs[[dataset_name]],
                                    limma_results[[dataset_name]]))
  stopifnot(is.na(worst) || worst < 1e-9)
}
checkpoint("within-plex contrast verified for all datasets")
```

A second sanity check: the limma log fold change should track the existing mean-based
`log2:FC` column computed by the Python pipeline. They are not identical — limma estimates the
contrast inside the blocked model and only over the plexes where the site was seen — but on
complete sites they should agree closely.

```{r verify-against-pipeline}
for (dataset_name in names(DATA_FILES)) {
  df      <- limma_inputs[[dataset_name]]
  results <- limma_results[[dataset_name]]

  matched   <- match(results$site, df$site)
  pipeline  <- df[[paste0(CELL_LINE, "_log2:FC_EGF_5")]][matched]
  new_value <- results[[paste0(CELL_LINE, "_log2:limmaFC_EGF_5")]]

  cat(sprintf("%-8s correlation(limmaFC, log2:FC) at EGF 5 min = %.4f\n",
              dataset_name,
              cor(pipeline, new_value, use = "complete.obs")))
}
```

```{r verify-ranges}
for (dataset_name in names(DATA_FILES)) {
  results <- limma_results[[dataset_name]]
  p_cols <- grep("pvalue|FDR", colnames(results), value = TRUE)
  p_cols <- grep("adjusted", p_cols, value = TRUE, invert = TRUE)
  values <- unlist(results[, p_cols], use.names = FALSE)

  stopifnot(all(is.na(values) | (values >= 0 & values <= 1)))
  cat(sprintf("%-8s %d sites x %d columns | all p-values and FDRs within [0, 1]\n",
              dataset_name, nrow(results), ncol(results)))
}
```

# Write the results

Results are written to a **separate** file keyed by `site`, in the same
`Experiment/{dataset}/Data/Processed/` folder the input came from, never back over the input
(`limma_output_path()` raises if the two paths ever collide). Merging them into the Python
frames (`pandas.merge(df, limma, on="site", how="left")`) is a downstream step, not part of
this notebook.

```{r write-results}
for (dataset_name in names(DATA_FILES)) {
  out_path <- limma_output_path(dataset_name)
  stage(paste("write", basename(out_path)),
        write.table(limma_results[[dataset_name]],
                    file = out_path,
                    sep = "\t",
                    row.names = FALSE,
                    quote = FALSE,
                    na = "NA"))
  cat("     wrote", out_path,
      "  (", nrow(limma_results[[dataset_name]]), "sites x",
      ncol(limma_results[[dataset_name]]), "columns,",
      round(file.size(out_path) / 1e6), "MB )\n")
}
checkpoint("all results written")
```

# If the run is slow

Benchmarked on simulated matrices of exactly this shape (68 channels, 20 coefficients, ~25%
missing), the arithmetic is not the expensive part: `lmFit` plus all five `eBayes` fits take
**~7 s at 50 000 sites**, assembly ~0.1 s and `write.table` ~2 s. A full three-dataset run should
be on the order of a minute. If it takes far longer, the checkpoints above will name the stage —
and the likely causes are, in order:

1. **Warning flooding in the IDE.** `lmFit` can emit one warning per site. `Rscript` absorbs that;
   an IDE notebook re-rendering tens of thousands of console lines does not. `muffle_warnings()`
   now collects and summarises them. If this was the cause, the fix is already in place.
2. **Swapping.** Watch the `memory:` lines. `read.delim` can transiently use several times the
   file size, and once the machine pages, everything becomes disk-bound. Reading each file once
   instead of three times (above) cuts the peak substantially.
3. **A much wider file than expected.** The `parsed: N rows x M columns` line says so directly.
4. **Missing values stored as `0` rather than `NA`.** This does not make the run slow — it makes it
   *fast and wrong*, because limma sees a complete matrix and takes its single-QR path. The
   `% exactly zero` diagnostic in `read_dataset()` catches it.

To time it cheaply before committing to a full run, set `MAX_SITES <- 2000` in the parameters
chunk. And prefer running the whole file with `Rscript -e 'rmarkdown::render("limma_for_pvalues.rmd")'`
over stepping through chunks in the IDE — the console rendering, not the statistics, is the part
that scales badly.

# Output columns

For each condition × timepoint, contrasted against `starve` within plex:

| Column | Meaning |
|---|---|
| `WT_log2:limmaFC_EGF_5` | log2 fold change vs starve, estimated inside the blocked model. Named `limmaFC` rather than `FC` so it does not collide with the mean-based `log2:FC` already produced by the Python pipeline |
| `WT_log2:pvalue_EGF_5` | moderated *t* p-value |
| `WT_log2:FDR_EGF_5` | Benjamini-Hochberg FDR, applied across sites within that contrast |
| `WT_log2:adjustedFDR_EGF_5` | `-log10(FDR)` |

Omnibus *F*-tests, per site:

| Column | Meaning |
|---|---|
| `WT_log2:Fpvalue_EGF_omnibus` | is this site different from starve at **any** EGF timepoint? (likewise `INS`, `EGFnINS`) |
| `WT_log2:Fpvalue_ALL_omnibus` | is this site different from starve at **any** timepoint of **any** condition? |
| `..._FFDR_..._omnibus`, `..._adjustedFFDR_..._omnibus` | BH-FDR and `-log10(FDR)` of the corresponding F-test |

The `full` timepoint gets its own *t* contrast but is deliberately **excluded from the
F-tests**: it is a media control, not a stimulation timepoint, so including it would let a site
qualify as "responsive" purely because full media differs from starve. Because `full` is a
single shared channel per plex, its contrast is emitted once, under the first condition name
(`WT_log2:pvalue_EGF_full`), rather than duplicated under all three arms.

## `limmaFC` and the existing `log2:FC` agree exactly

Worth recording, because it is easy to mistake for a bug: `WT_log2:limmaFC_EGF_5` comes out
**numerically identical** to the pipeline's `WT_log2:FC_EGF_5` (maximum difference 7e-15 on
`hme1_2`, at every value of `n:reps`, not only on complete sites).

That is the correct result rather than a coincidence. Because missingness is plex-wise, the
plexes a site is observed in always contribute a *complete and balanced* set of channels, and
for a balanced blocked design the fitted group difference reduces exactly to the mean of the
per-plex differences — which is what the Python `compute_fold_change` already computed.

So limma does not change the fold-change point estimate here. **What it adds is the variance
model**: the moderated *t* and *F* statistics, their p-values and the FDR — the part that
4 replicates cannot estimate reliably on their own. The `limmaFC` column is kept anyway so the
output file stands alone: each p-value ships with the effect size it refers to.

```{r session-info}
sessionInfo()
```
