---
title: "limma (diaPASEF) — p-values of each timepoint against the starve control"
output: html_notebook
---

# Purpose

Compute, for every phosphosite **of every cell line**, the statistical significance of each
stimulation timepoint against the **starve** control, using the `log2:abs` data produced by
`LFQ_diaPASEF.ipynb` / `run_diapasef_transformations()` (`src/transformations.py`).

This is the diaPASEF sibling of `limma_for_pvalues.rmd`. The statistics are the same — a moderated
*t* per timepoint plus omnibus moderated *F* tests, with `eBayes` shrinking each site's
variance toward a prior fitted across all sites (Smyth 2004) — but the experiment is not, and the
four differences that follow drive every change in this file.

| | TMT notebook (`limma_for_pvalues.rmd`) | this notebook (hme1_lfq diaPASEF) |
|---|---|---|
| unit of analysis | 3 datasets, 1 cell line each | 1 dataset, **8 cell lines** |
| conditions | EGF, INS, EGFnINS | **EGF only** |
| blocking | TMT **plex** (`r1…r4`) — a physical constraint | **none by default** — randomised injection order pairs nothing (`BLOCK_ON_REPLICATE`) |
| missingness | **per plex** — all 17 channels present or absent together | **per run** — every sample missing independently |

Three families of results are produced, per cell line:

1. **Per timepoint** — moderated *t* contrast against starve: log fold change, p-value, BH-FDR and
   adjusted p-value (`-log10(FDR)`).
2. **Per condition** — omnibus moderated *F* across that cell line's stimulation timepoints:
   *"is this site EGF-responsive at all, in this cell line?"*
3. **Global** — with a single condition this is the same test as (2), so the `ALL` column is emitted
   for naming compatibility with the TMT output and is **identical to the `EGF` column**. It is kept
   so that `filter_by_ffdr(conditions=["_ALL_"])` resolves on both dataset families.

**The data is already log2-transformed** (`log2:abs`). No further transformation is applied —
`lmFit` expects data on the log scale, which is exactly what it gets here.

# The design: the replicate is not a constraint here

Each cell line contributes 27 runs: 9 timepoints (`full`, `starve`, and 7 stimulation points)
× 3 replicates. The requirement is **different** from the TMT experiment, and the difference is
experimental, not statistical.

In TMT the 17 channels of one plex are labelled and injected together, so a channel can only be
compared against the starve channel **of its own plex** — the plex is a physical block, and
`~ 0 + group + plex` is not a modelling preference but a description of the experiment. Here every
sample is its own injection, loaded sequentially in randomised order, so nothing pairs the 5 min run
of replicate 2 with the starve run of replicate 2 rather than with the starve of replicate 1 or 3.
Any timepoint may be compared against any starve. The default design is therefore unblocked:

```
~ 0 + group
```

`group` is the timepoint (`starve`, `full`, `EGF_2` … `EGF_90`), and every contrast is that group
against `starve` pooled over all the runs where it was observed.

## But keep the choice visible: `BLOCK_ON_REPLICATE`

Randomised injection order makes the unblocked fit *valid* — it does not make blocking *wrong*.
`replicate` is still a plausible nuisance factor: if r1/r2/r3 are separate cultures or preparation
days, they carry an overall offset that randomisation does not remove, it only decouples from run
order. So the term is now an empirical question rather than a design requirement, and the notebook
treats it as one:

- `BLOCK_ON_REPLICATE <- FALSE` (default) fits `~ 0 + group`, matching the experimental design as
  described above. 9 coefficients, 18 residual df on a fully observed site.
- `BLOCK_ON_REPLICATE <- TRUE` fits `~ 0 + group + replicate`. 11 coefficients, 16 residual df.
- Either way `replicate_effect_report()` runs and prints **what the replicate term would absorb**:
  the fraction of sites whose replicate offsets differ from zero at FDR < 0.05, and the median
  largest offset in log2. Flip the switch on that evidence, not on principle.

**The one case where it is not a free choice.** With complete balanced data the two designs give the
*same* fold-change estimate and differ only in residual df. That equivalence needs the balance, and
DIA missingness breaks it per site: if a site was seen at 5 min only in r1 and at starve only in r2,
the unblocked contrast is `x[5,r1] − x[starve,r2]`, which carries the r1−r2 offset in full. The
blocked model estimates that offset from the site's other runs and removes it. So the sites where the
two designs disagree are exactly the sparsely observed ones — and they are a large minority in DIA.
The `replicate effect:` line tells you how big that error can be; if the offsets are negligible, the
unblocked fit is both simpler and slightly more powerful.

**Measured on the 200-row development sample — the offsets are not negligible.** Replicate offsets
differ from zero at FDR < 0.05 in 0% (BRAFS151A1) to 73% (RPS6KA3S375A) of tested sites, and the
median largest offset is **0.47–0.85 log2**, i.e. the same order as the biological fold changes being
tested. Two things follow, and both need re-measuring on the full file before they are acted on:

- It is **not** just sample loading. Re-running with `NORMALIZE = "median"` or `"quantile"` barely
  moves it (WT: 0.585 → 0.537 → 0.481 log2 median offset), so between-run scaling does not explain
  the replicate structure away — it looks like batch-level (culture / preparation) variation.
- The unblocked model does not *ignore* that variation, it puts it in the residual, which is a valid
  and conservative choice: p-values get larger, not smaller. What it cannot do is protect the point
  estimate on a site whose timepoint and control runs come from different replicates.

So the honest summary is that neither design is wrong: the experiment permits both, and the choice
trades a bias risk on sparse sites (unblocked) against 2 degrees of freedom and an assumption of
additive batch offsets (blocked). Worth running once each way on the full data and comparing the
responsive-site counts before settling.


## What is *not* duplicated here

The TMT files store `full` and `starve` three times, once per condition arm, and the TMT notebook
spends a function removing the copies. With a single condition there is nothing to remove:
`deduplicate_controls()` is kept below (unchanged, so both notebooks stay comparable, and so it
already works if INS / EGFnINS arms are added later) but on this dataset it is a **no-op**.

The `MIX_*` columns are not sample channels of any cell line and are simply never selected — the
column parser is anchored on the cell-line prefix.

# Missingness: per run, not per plex — and what replaces `MIN_PLEX`

This is the substantive statistical difference between the two notebooks.

In TMT, a peptide is either selected for MS2 in a run — yielding reporter ions in all 17 channels —
or it is not, so `n:reps` is exactly the number of plexes it was seen in and `MIN_PLEX = 2`
guarantees a well-posed fit. **None of that holds here.** In DIA every run is quantified
independently, so a site can be present in 14 of its 27 runs in an arbitrary pattern, and the
processed diaPASEF file does not even carry an `n:reps` column.

The replacement is a per-site, per-cell-line **observation count**, computed from the matrix that is
actually being fitted:

- `MIN_RESIDUAL_DF` is the quantity that matters — a site fitted with `df.residual = 0` still gets a
  p-value out of `eBayes`, computed entirely from the prior with *zero* site-specific evidence.
- Because `df.residual = n_observed − rank(design)` and the rank can never exceed the number of
  design columns, requiring `n_observed ≥ ncol(design) + MIN_RESIDUAL_DF` **guarantees** the wanted
  residual df. `MIN_OBS` is therefore derived, not guessed, and the notebook asserts the outcome
  after the fit rather than trusting the arithmetic.
- `MIN_CONTROL_OBS` additionally requires the **control** to be observed: with no `starve`
  observation every contrast of that site is inestimable, and the row would come back all-NA after
  a wasted fit.

Sites that pass are fitted with whatever they have. Timepoints a site never showed up at give NA
coefficients, which is correct — and has one consequence worth stating in advance, in
**Omnibus F and missing timepoints** below.

# Omnibus F and missing timepoints

limma computes the omnibus *F* from the vector of contrast *t*-statistics
(`classifyTestsF(fstat.only = TRUE)`), and that computation is **all-or-nothing**: if a site never
appeared at one timepoint, that contrast's *t* is NA and the site's *F* p-value is NA as well.

With TMT's plex-wise missingness this almost never bites — a site present in ≥ 2 plexes is present
at *every* timepoint of those plexes. With per-run missingness it bites often, so the notebook
reports two counts per cell line: how many sites have **every** group observed at least once (the
sites that can yield an F), and how many non-NA F p-values actually came out. Sites without a
complete group coverage are **not** dropped — their per-timepoint *t* contrasts are perfectly valid
and are reported; only their omnibus F is NA.

Downstream this is already handled: `filter_by_ffdr()` (`src/filters.py`) drops sites whose FFDR is
NA, and the project rule against `.fillna(0)` on statistics columns exists precisely so that these
untested sites are never waved through as FFDR = 0.

# Caveat: no between-sample normalisation is applied

`NORMALIZE <- "none"` fits `log2:abs` **exactly as stored on disk**, matching what the Python
pipeline computed, so `limmaFC` stays comparable to `log2:FC` (verified below). It is a deliberate
choice, and on **LFQ it is a more consequential one than on TMT**: TMT channels of one plex share a
labelling reaction and an injection, whereas here every sample is its own injection, and FragPipe
was run with normalisation `None` (the input file is `abundance_multi-site_MS2quant_None.tsv`).

The notebook therefore *measures* the exposure rather than asserting it: `read_dataset()` prints the
span of the per-run `log2:abs` medians, per cell line and overall. On the 200-row development sample
that span is **1.45 log2 within WT** and 4.29 log2 across all 216 runs (the latter inflated by each
column's median being taken over a different observed subset). Re-read those lines on the full file
before deciding.

Switching `NORMALIZE` to `"median"` or `"quantile"` needs no other change — but it moves the point
estimates, so the exact `limmaFC` ≈ `log2:FC` agreement checked at the end will loosen. That is
expected, not a failure.

# Parameters

Paths are resolved from the **project root**, located by walking up from the working directory until
a folder containing `data/`, `src/` and `notebooks/` is found. Relative paths cannot be used:
knitting runs with this file's folder as the working directory, running the chunks from the IDE
console runs with the project root, and one relative path cannot mean both.

`INPUT_FILE` selects between the 200-row development sample (for a smoke test, seconds) and the full
processed table. **Everything in `data/` is a truncated development sample** — never read biology off
a run made with `USE_SAMPLE <- TRUE`.

```{r setup}
suppressPackageStartupMessages(library(limma))

find_project_root <- function(start = getwd(),
                              markers = c("data", "src", "notebooks")) {
  # Locate the project root by walking up the directory tree from `start`.
  #
  # Makes the notebook runnable from any working directory — knitted (working directory =
  # this file's folder) or run chunk-by-chunk from a console rooted at the project.
  #
  # Args:
  #   start: directory to start searching from, default the current working directory.
  #   markers: directory names that must all be present for a folder to count as the root.
  #
  # Returns:
  #   Absolute path to the project root. Raises an error if no ancestor directory qualifies.

  path <- normalizePath(start, mustWork = TRUE)

  repeat {
    if (all(dir.exists(file.path(path, markers)))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("find_project_root: no ancestor of '", start, "' contains all of: ",
           paste(markers, collapse = ", "),
           ". Set PROJECT_ROOT manually to the TMT_Data_analysis folder.")
    }
    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

# --- input ----------------------------------------------------------------------------
# TRUE  -> data/…_sample.tsv, 200 rows, for smoke-testing the whole notebook in seconds.
# FALSE -> the full processed table in Experiment/hme1_diaPASEF/Data/Processed/.
USE_SAMPLE <- FALSE

SAMPLE_FILE <- file.path(PROJECT_ROOT, "data",
                         "20260818_peptide_MS2quant_None_transformed_nolimma_sample.tsv")
FULL_FILE   <- file.path(PROJECT_ROOT, "Experiment", "hme1_diaPASEF", "Data", "Processed",
                         "20260818_peptide_MS2quant_None_transformed_nolimma.tsv")

INPUT_FILE  <- if (USE_SAMPLE) SAMPLE_FILE else FULL_FILE
DATASET_NAME <- "hme1_diaPASEF"

# --- model ----------------------------------------------------------------------------
CELL_LINES <- c("WT", "EGFRT693A", "BRAFS151A1", "SOS1S1178A",
                "SHOC2T71A", "BRAFS151A2", "GAB1Y259A", "RPS6KA3S375A")
CONDITIONS <- c("EGF")      # single condition; the vector form is kept for future INS / EGFnINS arms
CONTROL    <- "starve"      # reference timepoint for every contrast
NORMALIZE  <- "none"        # "none" | "median" | "quantile"

# FALSE -> ~ 0 + group          (default; the runs are independent injections in random order)
# TRUE  -> ~ 0 + group + replicate
# See "The design: the replicate is not a constraint here". The replicate-effect diagnostic is
# printed either way, so the switch can be flipped on measurement rather than on principle.
BLOCK_ON_REPLICATE <- FALSE

# --- which sites are testable ---------------------------------------------------------
# See "Missingness: per run, not per plex". MIN_OBS is derived from MIN_RESIDUAL_DF inside
# run_limma_cell_line() as ncol(design) + MIN_RESIDUAL_DF, which is what guarantees the df.
MIN_RESIDUAL_DF <- 2        # residual df every fitted site is guaranteed to have
MIN_CONTROL_OBS <- 2        # runs the control timepoint must be observed in (1 = estimable but unreplicated)

OUT_SUFFIX <- "_limma_pvalues.tsv"

# --- debugging switches ---------------------------------------------------------------
# MAX_SITES caps the number of sites fitted, so the whole notebook can be smoke-tested in
# seconds before committing to a full run. Set to Inf for the real analysis.
MAX_SITES  <- Inf           # e.g. 2000 to time the pipeline end-to-end first
MAX_CELL_LINES <- Inf       # e.g. 2 to fit only the first cell lines while debugging
VERBOSE    <- TRUE          # per-stage timing checkpoints


limma_output_path <- function(input_path = INPUT_FILE,
                              out_suffix = OUT_SUFFIX) {
  # Build the output path of the limma table, next to the input file.
  #
  # The result is written into the folder the input was read from, so a run on the development
  # sample lands in data/ and a full run lands in Experiment/hme1_diaPASEF/Data/Processed/. The
  # name reuses the input's stem, so the date prefix carries over
  # (20260818_..._transformed_nolimma.tsv -> 20260818_..._limma_pvalues.tsv).
  #
  # Args:
  #   input_path: path of the processed .tsv being analysed.
  #   out_suffix: suffix appended to the input stem, including the extension.
  #
  # Returns:
  #   Absolute path of the .tsv to write. Never equal to the input path.

  stem <- basename(input_path)
  stem <- sub("_transformed_nolimma\\.tsv$", "", stem)
  stem <- sub("_log2_processed\\.tsv$", "", stem)
  stem <- sub("\\.tsv$", "", stem)               # fallback if the input is named differently

  output_path <- file.path(dirname(input_path), paste0(stem, out_suffix))
  if (!dir.exists(dirname(output_path))) {
    stop("limma_output_path: output folder '", dirname(output_path), "' does not exist.")
  }
  if (identical(output_path, input_path)) {
    stop("limma_output_path: refusing to overwrite the input file '", input_path, "'.")
  }
  output_path
}

cat("limma version:", as.character(packageVersion("limma")), "\n")
cat("working directory:", getwd(), "\n")
cat("project root     :", PROJECT_ROOT, "\n")
cat("input            :", INPUT_FILE, if (USE_SAMPLE) "  (DEVELOPMENT SAMPLE)" else "", "\n")
cat("output           :", limma_output_path(), "\n")
cat("cell lines       :", paste(CELL_LINES, collapse = ", "), "\n\n")

if (!file.exists(INPUT_FILE)) {
  stop("input file not found: ", INPUT_FILE)
}
```

# Progress checkpoints

Every stage announces itself **before** it runs and reports its duration after, and each line is
flushed immediately. So if the run stalls, the **last line printed is the stage it is stuck in** —
which is the whole point: a stage that never prints its `done` line is the culprit.

`report_memory()` is included because the most likely cause of a job that should take seconds
taking an hour is not arithmetic but **swapping**. If `used` climbs into the GB range, the machine
is paging and every subsequent operation is disk-bound.

`muffle_warnings()` addresses the other classic cause. `lmFit` can emit one warning *per site*, and
while `Rscript` shrugs that off, an IDE notebook re-rendering tens of thousands of warning lines can
take orders of magnitude longer than the computation itself. Warnings are counted and summarised
instead of streamed. Two of them are **expected on every cell line of this dataset** and are
explained under *Expected warnings* below.

```{r checkpoints}
.timing <- new.env(parent = emptyenv())

checkpoint <- function(label,
                       reset = FALSE) {
  # Print a flushed, timestamped progress line with the time since the previous checkpoint.
  #
  # Args:
  #   label: description of the stage being reported.
  #   reset: TRUE to restart the total-elapsed clock (use at the start of a cell line).
  #
  # Returns:
  #   Invisibly, seconds elapsed since the previous checkpoint.

  if (!VERBOSE) {
    return(invisible(0))
  }

  now <- proc.time()[["elapsed"]]
  if (reset || is.null(.timing$start)) {
    .timing$start <- now
    .timing$last  <- now
  }
  since_last  <- now - .timing$last
  since_start <- now - .timing$start
  .timing$last <- now

  cat(sprintf("[%s | +%7.1fs | %7.1fs total] %s\n",
              format(Sys.time(), "%H:%M:%S"),
              since_last,
              since_start,
              label))
  flush.console()
  invisible(since_last)
}


stage <- function(label,
                  expr) {
  # Run one stage, announcing it before evaluation and timing it after.
  #
  # `expr` is passed unevaluated and forced inside, so the "starting" line is guaranteed to be
  # printed and flushed before any work begins. A stage that hangs therefore leaves its own name
  # as the last thing on screen.
  #
  # Args:
  #   label: short description of the stage, e.g. "lmFit".
  #   expr: the expression to evaluate.
  #
  # Returns:
  #   The value of `expr`.

  if (VERBOSE) {
    cat(sprintf("[%s]   -> %-42s ", format(Sys.time(), "%H:%M:%S"), paste0(label, " ...")))
    flush.console()
  }

  started <- proc.time()[["elapsed"]]
  value   <- force(expr)
  elapsed <- proc.time()[["elapsed"]] - started

  if (VERBOSE) {
    cat(sprintf("done in %7.1fs\n", elapsed))
    flush.console()
  }
  value
}


report_memory <- function(label = "") {
  # Print the memory currently used by R, to expose swapping as a cause of a slow run.
  #
  # Args:
  #   label: optional tag printed alongside the figures.
  #
  # Returns:
  #   Invisibly, total megabytes in use.

  if (!VERBOSE) {
    return(invisible(0))
  }

  # gc() returns a variable number of columns — a "limit (Mb)" column appears on some platforms,
  # so the megabyte columns must be located by name rather than by position. The first "(Mb)"
  # column is current usage, the last is the peak since the session started.
  usage   <- gc(verbose = FALSE)
  mb_cols <- which(colnames(usage) == "(Mb)")
  used    <- sum(usage[, mb_cols[1]])
  peak    <- if (length(mb_cols) > 1) sum(usage[, mb_cols[length(mb_cols)]]) else NA_real_

  cat(sprintf("[%s]      memory: %6.0f MB in use%s  %s\n",
              format(Sys.time(), "%H:%M:%S"),
              used,
              if (is.na(peak)) "" else sprintf(", %6.0f MB peak", peak),
              label))
  flush.console()
  invisible(used)
}


muffle_warnings <- function(expr,
                            label = "stage") {
  # Evaluate an expression, collecting warnings instead of emitting them one by one.
  #
  # limma emits one warning per problematic site; tens of thousands of them can dominate the
  # runtime of an IDE notebook even though the computation itself is fast. This counts them and
  # prints a summary of the distinct messages.
  #
  # Args:
  #   expr: expression to evaluate.
  #   label: tag used in the printed summary.
  #
  # Returns:
  #   The value of `expr`.

  collected <- character(0)

  value <- withCallingHandlers(
    expr,
    warning = function(w) {
      collected <<- c(collected, conditionMessage(w))
      invokeRestart("muffleWarning")
    })

  if (length(collected) > 0 && VERBOSE) {
    counts <- sort(table(collected), decreasing = TRUE)
    cat(sprintf("[%s]      %d warning(s) in %s:\n",
                format(Sys.time(), "%H:%M:%S"), length(collected), label))
    for (i in seq_len(min(5, length(counts)))) {
      cat(sprintf("           %6d x  %s\n", counts[[i]], names(counts)[i]))
    }
    flush.console()
  }
  value
}
```

# Functions

`parse_abs_columns()`, `deduplicate_controls()`, `sort_timepoints()`, `normalize_matrix()`,
`build_contrasts()`, `order_by_datatype()` and `add_test_columns()` are the TMT notebook's functions,
carried over with one rename (`plex` → `replicate`, since these are LC-MS injections rather than
labelling reactions) so the two files stay line-by-line comparable. `site_coverage()` is new and does
the job `n:reps` did in TMT.

```{r functions}
parse_abs_columns <- function(column_names,
                              cell_line,
                              conditions) {
  # Parse the log2:abs column names of one cell line into their component fields.
  #
  # The cell line is matched as the whole first field (the pattern is anchored and the field
  # cannot contain "_"), so a name that prefixes another — BRAFS151A1 vs BRAFS151A2 — cannot pull
  # in both, and the MIX_* control channels are never selected.
  #
  # Args:
  #   column_names: character vector of all column names in the dataset.
  #   cell_line: cell line prefix to keep, e.g. "WT".
  #   conditions: character vector of condition names, e.g. c("EGF").
  #
  # Returns:
  #   data.frame with one row per log2:abs column and the columns
  #   `column`, `condition`, `timepoint`, `replicate`.

  pattern <- paste0("^", cell_line, "_log2:abs_(",
                    paste(conditions, collapse = "|"),
                    ")_([^_]+)_(r[0-9]+)$")

  matched <- grep(pattern, column_names, value = TRUE)
  if (length(matched) == 0) {
    stop("parse_abs_columns: no log2:abs columns found for cell line '", cell_line, "'.")
  }

  fields <- regmatches(matched, regexec(pattern, matched))

  data.frame(column    = matched,
             condition = vapply(fields, `[`, character(1), 2),
             timepoint = vapply(fields, `[`, character(1), 3),
             replicate = vapply(fields, `[`, character(1), 4),
             stringsAsFactors = FALSE)
}


deduplicate_controls <- function(df,
                                 column_info,
                                 conditions,
                                 control_timepoints = c("full", "starve")) {
  # Remove the duplicated control columns that appear once per condition arm.
  #
  # NO-OP ON THIS DATASET: with a single condition there is nothing to de-duplicate, and the
  # function only adds the `group` column. It is kept unchanged from the TMT notebook so that the
  # two files stay comparable and so that adding INS / EGFnINS arms later needs no new code — in
  # the TMT files `full` and `starve` are one physical channel written out under every condition,
  # and feeding all copies to lmFit would be pseudo-replication.
  #
  # Args:
  #   df: the full dataset, used only to compare the duplicated columns.
  #   column_info: data.frame returned by parse_abs_columns().
  #   conditions: character vector of condition names; the first one is kept.
  #   control_timepoints: timepoints that are shared across condition arms.
  #
  # Returns:
  #   column_info restricted to unique samples, with an added `group` column giving the
  #   experimental group of each sample (control timepoints keep their bare name).

  keep_condition <- conditions[1]

  for (timepoint in control_timepoints) {
    for (replicate in unique(column_info$replicate)) {
      reference <- column_info$column[column_info$condition == keep_condition &
                                      column_info$timepoint == timepoint &
                                      column_info$replicate == replicate]
      if (length(reference) != 1) next

      for (other_condition in conditions[-1]) {
        duplicate <- column_info$column[column_info$condition == other_condition &
                                        column_info$timepoint == timepoint &
                                        column_info$replicate == replicate]
        if (length(duplicate) != 1) next

        if (!isTRUE(all.equal(df[[reference]], df[[duplicate]]))) {
          stop("deduplicate_controls: '", duplicate, "' is not identical to '", reference,
               "'. The control channels are not duplicates in this dataset — ",
               "the design must be rebuilt before fitting.")
        }
      }
    }
  }

  is_duplicate_control <- column_info$timepoint %in% control_timepoints &
                          column_info$condition != keep_condition
  unique_info <- column_info[!is_duplicate_control, , drop = FALSE]

  unique_info$group <- ifelse(unique_info$timepoint %in% control_timepoints,
                              unique_info$timepoint,
                              paste(unique_info$condition, unique_info$timepoint, sep = "_"))

  unique_info
}


sort_timepoints <- function(timepoints) {
  # Sort timepoint labels in biological order: named controls first, then numerically.
  #
  # Numeric sorting rather than a fixed list is what puts the diaPASEF grid
  # {full, starve, 2, 5, 10, 15, 20, 30, 90} in experimental order — a hard-coded TMT ordering
  # has no 20 or 30 min and would emit them after 90.
  #
  # Args:
  #   timepoints: character vector of timepoint labels, e.g. c("10", "full", "2", "starve").
  #
  # Returns:
  #   The same labels, ordered with "full" and "starve" first and the numeric points ascending.

  named   <- c("full", "starve")
  present <- named[named %in% timepoints]
  numeric_points <- setdiff(timepoints, named)
  c(present, numeric_points[order(as.numeric(numeric_points))])
}


normalize_matrix <- function(M,
                             method) {
  # Apply between-sample normalisation to the expression matrix.
  #
  # Args:
  #   M: numeric matrix, sites in rows and samples in columns, on the log2 scale.
  #   method: "none" (return unchanged), "median" (centre every column on the grand median),
  #     or "quantile" (limma::normalizeBetweenArrays).
  #
  # Returns:
  #   The normalised matrix, same dimensions and dimnames as the input.

  if (method == "none") {
    return(M)
  }
  if (method == "median") {
    column_medians <- apply(M, 2, median, na.rm = TRUE)
    return(sweep(M, 2, column_medians) + mean(column_medians))
  }
  if (method == "quantile") {
    return(normalizeBetweenArrays(M, method = "quantile"))
  }
  stop("normalize_matrix: unknown method '", method, "'.")
}


site_coverage <- function(df,
                          samples,
                          control = CONTROL) {
  # Count, per site, how many of one cell line's runs observed it.
  #
  # This is the diaPASEF replacement for the TMT `n:reps` column. It cannot be read off a stored
  # column: `n:reps` does not exist in the processed diaPASEF table, and DIA missingness is
  # per run rather than per plex, so the number of replicates a site was seen in does not describe
  # which runs those were.
  #
  # Args:
  #   df: the dataset, containing the log2:abs columns listed in `samples`.
  #   samples: data.frame returned by deduplicate_controls(), one row per run, with the columns
  #     `column` and `group`.
  #   control: name of the control group, taken as an argument rather than read from the global so
  #     it cannot drift away from the `control` the fit is using.
  #
  # Returns:
  #   data.frame with one row per site of `df` and the columns
  #     n_obs          — runs of this cell line in which the site was quantified,
  #     n_control_obs  — runs of the control group in which it was quantified,
  #     n_groups_obs   — timepoint groups observed at least once,
  #     n_fgroups_obs  — the same, counting only the groups the omnibus F is built from
  #       (control + stimulation timepoints, i.e. everything except `full`). A site needs all of
  #       them to get an F; `full` is irrelevant to it, so counting `full` would under-report.

  observed <- !is.na(as.matrix(df[, samples$column, drop = FALSE]))

  group_levels <- unique(as.character(samples$group))
  per_group <- vapply(group_levels,
                      function(group) {
                        rowSums(observed[, samples$group == group, drop = FALSE]) > 0
                      },
                      logical(nrow(df)))
  per_group <- matrix(per_group, nrow = nrow(df),
                      dimnames = list(NULL, group_levels))   # keep the shape for a 1-site frame

  f_groups <- setdiff(group_levels, "full")

  data.frame(n_obs         = rowSums(observed),
             n_control_obs = rowSums(observed[, samples$group == control, drop = FALSE]),
             n_groups_obs  = rowSums(per_group),
             n_fgroups_obs = rowSums(per_group[, f_groups, drop = FALSE]))
}


replicate_effect_report <- function(M,
                                    samples,
                                    label = "") {
  # Measure what a replicate block term would absorb, whatever design is actually being fitted.
  #
  # The diaPASEF runs are independent injections in randomised order, so blocking on the replicate
  # is a choice rather than a description of the experiment (see "The design"). This makes the
  # choice empirical: it fits ~ 0 + group + replicate on the same matrix and tests, per site,
  # H0: every replicate offset is zero. Two numbers come out — how often the offsets are
  # detectable, and how large the largest one typically is, in log2 units directly comparable to a
  # fold change.
  #
  # It is deliberately run in both modes. With BLOCK_ON_REPLICATE = FALSE it reports the systematic
  # error the unblocked fit is exposed to on unevenly observed sites; with TRUE it reports what the
  # two spent degrees of freedom bought.
  #
  # Args:
  #   M: the expression matrix being fitted, sites in rows and runs in columns.
  #   samples: data.frame with one row per run, carrying the `group` and `replicate` factors.
  #   label: tag printed with the report, e.g. the cell line.
  #
  # Returns:
  #   Invisibly, a list with n_tested, n_significant and median_abs_offset, or NULL if the design
  #   has no replicate term to test (a single replicate).

  design <- model.matrix(~ 0 + group + replicate, data = samples)
  colnames(design) <- make.names(colnames(design))
  replicate_coefs <- grep("^replicate", colnames(design), value = TRUE)

  if (length(replicate_coefs) == 0 || qr(design)$rank < ncol(design)) {
    cat("     replicate effect: not estimable (single replicate or rank-deficient design)\n")
    return(invisible(NULL))
  }

  fit <- muffle_warnings(lmFit(M, design), "replicate-effect lmFit")
  fit_r <- muffle_warnings(eBayes(contrasts.fit(fit,
                                                makeContrasts(contrasts = replicate_coefs,
                                                              levels = design)),
                                  trend = TRUE,
                                  robust = TRUE),
                           "replicate-effect F")

  fdr <- p.adjust(fit_r$F.p.value, method = "BH")

  # Largest absolute offset per site, in log2 units: directly comparable to a fold change.
  offsets <- suppressWarnings(apply(abs(fit_r$coefficients), 1, max, na.rm = TRUE))
  offsets[!is.finite(offsets)] <- NA_real_

  n_tested      <- sum(!is.na(fdr))
  n_significant <- sum(fdr < 0.05, na.rm = TRUE)

  cat(sprintf(paste0("     replicate effect%s: offsets non-zero at FDR < 0.05 in %d of %d sites (%.1f%%)",
                     " | median largest |offset| = %.3f log2\n"),
              if (nzchar(label)) paste0(" (", label, ")") else "",
              n_significant, n_tested,
              if (n_tested > 0) 100 * n_significant / n_tested else NA_real_,
              median(offsets, na.rm = TRUE)))
  flush.console()

  invisible(list(n_tested = n_tested,
                 n_significant = n_significant,
                 median_abs_offset = median(offsets, na.rm = TRUE)))
}


build_contrasts <- function(groups,
                            control,
                            design) {
  # Build the contrast matrix of every group against the control group.
  #
  # Args:
  #   groups: character vector of group names to contrast against the control.
  #   control: name of the control group, e.g. "starve".
  #   design: design matrix whose column names define the coefficient names.
  #
  # Returns:
  #   Contrast matrix with one column per group, named after the group.

  expressions <- paste0("group", groups, " - group", control)
  contrast_matrix <- makeContrasts(contrasts = expressions, levels = design)
  colnames(contrast_matrix) <- groups
  contrast_matrix
}


order_by_datatype <- function(results,
                              cell_line,
                              labels,
                              f_names) {
  # Reorder the result columns so each data type forms one contiguous block.
  #
  # The natural assembly order is per contrast (all four statistics of EGF_2, then all four of
  # EGF_5, ...). This flips it to per data type: every limmaFC column with its timepoints in
  # order, then every pvalue column, then FDR, then adjustedFDR, then the F-test blocks.
  #
  # Args:
  #   results: assembled results data.frame, keyed by `site`.
  #   cell_line: cell line column prefix, e.g. "WT".
  #   labels: contrast labels ("EGF_full", "EGF_2", ...) in the desired timepoint order.
  #   f_names: names of the F-tests, e.g. c("EGF", "ALL").
  #
  # Returns:
  #   The same data.frame with its columns reordered. Raises if the reordering would drop or
  #   invent a column.

  contrast_stats <- c("limmaFC", "pvalue", "FDR", "adjustedFDR")
  f_stats        <- c("Fpvalue", "FFDR", "adjustedFFDR")

  ordered <- "site"
  for (statistic in contrast_stats) {
    ordered <- c(ordered, paste0(cell_line, "_log2:", statistic, "_", labels))
  }
  for (statistic in f_stats) {
    ordered <- c(ordered, paste0(cell_line, "_log2:", statistic, "_", f_names, "_omnibus"))
  }

  if (!setequal(ordered, colnames(results))) {
    stop("order_by_datatype: the reordered column list does not match the assembled one.")
  }

  results[, ordered, drop = FALSE]
}


add_test_columns <- function(results,
                             p_values,
                             prefix) {
  # Append p-value, BH-FDR and adjusted p-value columns for one test to the results frame.
  #
  # The adjusted p-value is -log10(FDR), matching the project's `adjustedFDR` convention.
  # An FDR of exactly 0 would give +Inf and is set to NA instead.
  #
  # BH is applied over the non-NA p-values only (p.adjust's default n is the number of non-missing
  # values), which is the intended behaviour here: a site whose contrast was inestimable — it never
  # appeared at that timepoint — was not tested and must not enlarge the multiple-testing burden.
  #
  # Args:
  #   results: data.frame being assembled, one row per site.
  #   p_values: numeric vector of raw p-values, in the row order of `results`.
  #   prefix: named list of the three column names to write (pvalue / fdr / adjustedfdr).
  #
  # Returns:
  #   The results data.frame with three new columns appended.

  fdr <- p.adjust(p_values, method = "BH")
  adjusted <- -log10(fdr)
  adjusted[is.infinite(adjusted)] <- NA_real_

  results[[prefix$pvalue]]      <- p_values
  results[[prefix$fdr]]         <- fdr
  results[[prefix$adjustedfdr]] <- adjusted
  results
}
```

# The analysis

```{r read-dataset}
read_dataset <- function(file_path,
                         dataset_name,
                         cell_lines = CELL_LINES) {
  # Read the processed dataset and report everything needed to explain a slow or wrong run.
  #
  # The file is read once and reused by every cell-line fit and by both verification steps. The
  # diagnostics answer, in order: is the file the size we think it is, did it parse into the
  # expected shape, are the intensity columns numeric, are missing values stored as NA rather than
  # as 0, and how far apart are the per-run medians (the normalisation exposure).
  #
  # Args:
  #   file_path: path to the processed .tsv.
  #   dataset_name: dataset key, used in the printed report.
  #   cell_lines: cell lines whose per-run median span is reported.
  #
  # Returns:
  #   The dataset as a data.frame, with column names untouched.

  size_mb <- file.size(file_path) / 1e6
  cat(sprintf("[%s]      file: %.0f MB  %s\n",
              format(Sys.time(), "%H:%M:%S"), size_mb, basename(file_path)))
  flush.console()

  df <- stage(sprintf("read.delim (%.0f MB)", size_mb),
              read.delim(file_path,
                         sep = "\t",
                         check.names = FALSE,
                         na.strings = c("NA", "NaN", ""),
                         stringsAsFactors = FALSE))

  cat(sprintf("[%s]      parsed: %d rows x %d columns\n",
              format(Sys.time(), "%H:%M:%S"), nrow(df), ncol(df)))

  abs_columns <- grep("_log2:abs_", colnames(df), value = TRUE)
  non_numeric <- abs_columns[!vapply(df[abs_columns], is.numeric, logical(1))]
  if (length(non_numeric) > 0) {
    cat(sprintf("      WARNING: %d log2:abs column(s) did not parse as numeric, e.g. %s\n",
                length(non_numeric), paste(head(non_numeric, 3), collapse = ", ")))
  }

  # Missingness must be NA. If a fillna(0) reached this file, the zeros are read as real
  # intensities: limma then sees a complete matrix, takes its fast path, and returns confident
  # nonsense. This is the defect recorded in clustering_method_decision.md §1.
  numeric_abs <- abs_columns[vapply(df[abs_columns], is.numeric, logical(1))]
  if (length(numeric_abs) > 0) {
    values    <- as.matrix(df[numeric_abs])
    n_missing <- sum(is.na(values))
    n_zero    <- sum(values == 0, na.rm = TRUE)
    cat(sprintf("      log2:abs cells: %.1f%% NA, %.1f%% exactly zero  (%d runs)\n",
                100 * n_missing / length(values),
                100 * n_zero / length(values),
                length(numeric_abs)))
    if (n_zero > 0.01 * length(values)) {
      cat("      WARNING: many exact zeros in log2:abs — missing values may have been\n",
          "               written as 0 instead of NA. Fix before trusting any p-value.\n", sep = "")
    }

    # Normalisation exposure: how far apart the per-run medians are. NORMALIZE = "none" fits the
    # data as stored, so whatever span is printed here is inherited by every log fold change.
    run_medians <- apply(values, 2, median, na.rm = TRUE)
    cat(sprintf("      per-run log2:abs medians: span %.2f log2 over all runs\n",
                diff(range(run_medians, na.rm = TRUE))))
    for (cell_line in cell_lines) {
      own <- run_medians[startsWith(names(run_medians), paste0(cell_line, "_"))]
      if (length(own) > 1) {
        cat(sprintf("        %-14s %5.2f log2 across its %d runs\n",
                    cell_line, diff(range(own, na.rm = TRUE)), length(own)))
      }
    }
  }

  report_memory(sprintf("after reading %s", dataset_name))
  flush.console()
  df
}
```

```{r run-limma}
run_limma_cell_line <- function(df,
                                cell_line,
                                control = CONTROL,
                                conditions = CONDITIONS,
                                min_residual_df = MIN_RESIDUAL_DF,
                                min_control_obs = MIN_CONTROL_OBS,
                                normalize = NORMALIZE,
                                block_on_replicate = BLOCK_ON_REPLICATE,
                                max_sites = MAX_SITES) {
  # Fit the limma model on one cell line and return its table of statistics.
  #
  # Every stimulation timepoint is contrasted against the control timepoint, using ~ 0 + group
  # (the default: the runs are independent injections in randomised order, so nothing pairs a
  # timepoint with a particular replicate's control) or ~ 0 + group + replicate when
  # block_on_replicate is TRUE. Three sets of statistics are produced: the moderated t contrast
  # per timepoint, an omnibus moderated F per condition, and the global omnibus F (identical to
  # the per-condition one while there is a single condition).
  #
  # Args:
  #   df: the dataset as returned by read_dataset() — read once and reused, so the file is not
  #     parsed again per cell line.
  #   cell_line: cell line column prefix to analyse, e.g. "BRAFS151A1".
  #   control: control timepoint every contrast is taken against.
  #   conditions: condition names present in the column headers.
  #   min_residual_df: residual degrees of freedom every fitted site is required to have; the
  #     observation threshold is derived from it as ncol(design) + min_residual_df.
  #   min_control_obs: runs of the control timepoint a site must be observed in.
  #   normalize: between-sample normalisation method passed to normalize_matrix().
  #   block_on_replicate: TRUE to add the replicate term to the design. FALSE (default) fits the
  #     unblocked model; the replicate-effect diagnostic is reported either way.
  #   max_sites: cap on the number of sites fitted, for smoke-testing; Inf for a real run.
  #
  # Returns:
  #   data.frame with one row per tested site: the `site` key followed by this cell line's
  #   logFC / pvalue / FDR / adjustedFDR columns per timepoint and its F-test columns. Sites that
  #   did not pass the coverage filter are absent, not NA-filled — they are re-introduced as NA
  #   when the cell lines are merged.

  cat("\n=====", cell_line, "=====\n")
  checkpoint(paste("fitting", cell_line), reset = TRUE)

  stopifnot(!anyDuplicated(df$site))

  # --- samples -------------------------------------------------------------------------
  column_info <- stage("parse log2:abs column names",
                       parse_abs_columns(colnames(df), cell_line, conditions))
  cat("     log2:abs columns:", nrow(column_info), "\n")

  samples <- stage("de-duplicate control channels",
                   deduplicate_controls(df, column_info, conditions))
  cat("     unique sample runs after de-duplicating the controls:", nrow(samples), "\n")

  timepoints <- sort_timepoints(unique(column_info$timepoint))
  cat("     timepoints:", paste(timepoints, collapse = ", "), "\n")

  group_levels <- c(control,
                    setdiff(c("full", "starve"), control),
                    unlist(lapply(conditions, function(condition) {
                      paste(condition,
                            setdiff(timepoints, c("full", "starve")),
                            sep = "_")
                    })))
  group_levels <- group_levels[group_levels %in% samples$group]

  samples$group     <- factor(samples$group, levels = group_levels)
  samples$replicate <- factor(samples$replicate)

  # --- which sites are testable --------------------------------------------------------
  # The design is built here, before the filtering, because MIN_OBS is derived from its width:
  # df.residual = n_obs - rank(design) and rank <= ncol(design), so n_obs >= ncol + min_residual_df
  # guarantees the residual df. This is asserted after the fit rather than trusted.
  design <- if (block_on_replicate) {
    model.matrix(~ 0 + group + replicate, data = samples)
  } else {
    model.matrix(~ 0 + group, data = samples)
  }
  colnames(design) <- make.names(colnames(design))
  min_obs <- ncol(design) + min_residual_df
  cat(sprintf("     design formula: ~ 0 + group%s\n",
              if (block_on_replicate) " + replicate" else "   (unblocked: any run contrasted against all observed controls)"))

  coverage <- stage("count per-site coverage", site_coverage(df, samples, control = control))
  tested   <- coverage$n_obs >= min_obs & coverage$n_control_obs >= min_control_obs
  tested[is.na(tested)] <- FALSE

  if (is.finite(max_sites) && sum(tested) > max_sites) {
    tested[which(tested)[-seq_len(max_sites)]] <- FALSE
    cat("     NOTE: MAX_SITES is set — only", max_sites, "sites will be fitted.\n")
  }

  cat(sprintf("     coverage filter: n_obs >= %d (%d design columns + %d df) and %s observed in >= %d runs\n",
              min_obs, ncol(design), min_residual_df, control, min_control_obs))
  # The two failure counts overlap — a site with almost no runs usually fails both.
  cat(sprintf("     sites tested: %d of %d  | dropped: %d (too few runs: %d; no usable %s: %d; the two overlap)\n",
              sum(tested), nrow(df), sum(!tested),
              sum(coverage$n_obs < min_obs),
              control,
              sum(coverage$n_control_obs < min_control_obs)))

  if (sum(tested) == 0) {
    warning("run_limma_cell_line: no site passes the coverage filter for '", cell_line,
            "' — skipped.")
    return(NULL)
  }

  # An omnibus F needs every group estimable for that site (see "Omnibus F and missing
  # timepoints"): limma computes it from the vector of contrast t-statistics, and one NA in that
  # vector makes the whole F NA. This is how many sites can possibly get one.
  n_f_eligible <- sum(tested & coverage$n_fgroups_obs == nlevels(samples$group) - ("full" %in% timepoints))
  cat(sprintf("     of those, %d (%.1f%%) were observed at every timepoint the F is built from — only these can yield an omnibus F\n",
              n_f_eligible, 100 * n_f_eligible / sum(tested)))

  # --- expression matrix ---------------------------------------------------------------
  M <- stage("build expression matrix",
             as.matrix(df[tested, samples$column, drop = FALSE]))
  rownames(M) <- df$site[tested]
  storage.mode(M) <- "double"
  M <- stage(paste0("normalise (", normalize, ")"), normalize_matrix(M, normalize))
  cat("     matrix:", nrow(M), "sites x", ncol(M), "runs | ",
      sprintf("%.1f%% NA", 100 * sum(is.na(M)) / length(M)), "\n")

  # What a replicate block would absorb — reported whichever design is being fitted, so the
  # BLOCK_ON_REPLICATE choice rests on this number rather than on principle.
  invisible(stage("replicate-effect diagnostic",
                  replicate_effect_report(M, samples, label = cell_line)))

  # --- design and fit ------------------------------------------------------------------
  cat("     design:", nrow(design), "x", ncol(design),
      " | rank:", qr(design)$rank, "\n")
  stopifnot(qr(design)$rank == ncol(design))

  cat(sprintf("     lmFit will fit %d sites x %d coefficients%s\n",
              nrow(M), ncol(design),
              if (anyNA(M)) " (per-site loop: NAs present)" else " (single QR: no NAs)"))
  fit <- stage("lmFit", muffle_warnings(lmFit(M, design), "lmFit"))
  report_memory("after lmFit")

  # The point of deriving min_obs from ncol(design): every fitted site must clear min_residual_df.
  cat(sprintf("     residual df: min %d, median %.0f, max %d\n",
              min(fit$df.residual), median(fit$df.residual), max(fit$df.residual)))
  stopifnot(min(fit$df.residual) >= min_residual_df)

  # --- contrasts -----------------------------------------------------------------------
  stimulation_timepoints <- setdiff(timepoints, c("full", "starve"))
  contrast_groups <- setdiff(levels(samples$group), control)

  fit_t <- stage(sprintf("t-contrasts + eBayes (%d contrasts)", length(contrast_groups)),
                 muffle_warnings(
                   eBayes(contrasts.fit(fit,
                                        build_contrasts(contrast_groups, control, design)),
                          trend = TRUE,
                          robust = TRUE),
                   "t-contrasts"))

  f_groups <- list()
  for (condition in conditions) {
    f_groups[[condition]] <- paste(condition, stimulation_timepoints, sep = "_")
  }
  f_groups[["ALL"]] <- unique(unlist(f_groups, use.names = FALSE))

  f_fits <- list()
  for (f_name in names(f_groups)) {
    f_fits[[f_name]] <- stage(sprintf("omnibus F: %s", f_name),
                              muffle_warnings(
                                eBayes(contrasts.fit(fit,
                                                     build_contrasts(f_groups[[f_name]],
                                                                     control,
                                                                     design)),
                                       trend = TRUE,
                                       robust = TRUE),
                                paste("F test", f_name)))
  }
  report_memory("after all fits")

  # --- assemble ------------------------------------------------------------------------
  # `contrast_groups` is already in reporting order: full first, then the condition with its
  # timepoints ascending. `labels` keeps that order and is reused to order the output columns.
  labels <- vapply(contrast_groups,
                   function(group) {
                     if (group %in% timepoints) paste0(conditions[1], "_", group) else group
                   },
                   character(1),
                   USE.NAMES = FALSE)

  results <- stage("assemble result columns", {
    assembled <- data.frame(site = rownames(M), stringsAsFactors = FALSE)

    for (i in seq_along(contrast_groups)) {
      group <- contrast_groups[i]
      label <- labels[i]

      assembled[[paste0(cell_line, "_log2:limmaFC_", label)]] <- fit_t$coefficients[, group]
      assembled <- add_test_columns(assembled,
                                    fit_t$p.value[, group],
                                    list(pvalue      = paste0(cell_line, "_log2:pvalue_", label),
                                         fdr         = paste0(cell_line, "_log2:FDR_", label),
                                         adjustedfdr = paste0(cell_line, "_log2:adjustedFDR_", label)))
    }

    for (name in names(f_fits)) {
      assembled <- add_test_columns(assembled,
                                    f_fits[[name]]$F.p.value,
                                    list(pvalue      = paste0(cell_line, "_log2:Fpvalue_", name, "_omnibus"),
                                         fdr         = paste0(cell_line, "_log2:FFDR_", name, "_omnibus"),
                                         adjustedfdr = paste0(cell_line, "_log2:adjustedFFDR_", name, "_omnibus")))
    }
    assembled
  })

  results <- stage("reorder columns by data type",
                   order_by_datatype(results, cell_line, labels, names(f_fits)))
  checkpoint(sprintf("%s fitted: %d sites x %d columns",
                     cell_line, nrow(results), ncol(results)))

  # --- report --------------------------------------------------------------------------
  cat("\nsites at FDR < 0.05 per contrast (of the sites where the contrast was estimable):\n")
  for (label in labels) {
    fdr_col <- paste0(cell_line, "_log2:FDR_", label)
    cat(sprintf("  %-16s %5d significant / %5d tested / %d fitted\n", label,
                sum(results[[fdr_col]] < 0.05, na.rm = TRUE),
                sum(!is.na(results[[fdr_col]])),
                nrow(results)))
  }
  cat("sites at F-test FDR < 0.05:\n")
  for (name in names(f_fits)) {
    fdr_col <- paste0(cell_line, "_log2:FFDR_", name, "_omnibus")
    cat(sprintf("  %-16s %5d significant / %5d with an F / %d fitted\n", name,
                sum(results[[fdr_col]] < 0.05, na.rm = TRUE),
                sum(!is.na(results[[fdr_col]])),
                nrow(results)))
  }

  results
}
```

## Expected warnings, and one silent approximation

`muffle_warnings()` summarises warnings instead of streaming them. On the development sample exactly
one kind appears, once per cell line:

- **`Partial NA coefficients for N probe(s)`** — a site missing from every run of one timepoint has
  an inestimable coefficient for that group. Its own contrast becomes NA (correctly), the others
  stay valid, and `df.residual` is reduced accordingly. In TMT this only ever hit *replicate*
  nuisance terms; here it hits group terms as well, which is exactly why the omnibus F can be NA.
- **`Zero sample variances detected`** may additionally appear on the full file — a site whose
  observed values are identical within groups. `eBayes` offsets them and the moderated statistic
  then rests on the prior, which is what the shrinkage is for.

The approximation worth knowing about is **not** warned at all. With missing values `lmFit` fits each
site against its own reduced design, so `stdev.unscaled`, `sigma` and `df.residual` — everything the
moderated *t* uses — are exact per site. But `cov.coefficients` is returned once, computed from the
**complete** design, and that is what `contrasts.fit` turns into the between-contrast correlation
matrix behind the omnibus *F*. For a site missing several runs, the correlation structure the F
assumes is therefore the one it would have had with complete data. The TMT notebook inherits the same
approximation on ~25% NA; here the NA fraction is much higher, so read the omnibus F as a screening
statistic rather than an exact test.

```{r fit-all-cell-lines}
limma_input <- read_dataset(INPUT_FILE, DATASET_NAME)

fit_cell_lines <- head(CELL_LINES, if (is.finite(MAX_CELL_LINES)) MAX_CELL_LINES else length(CELL_LINES))

limma_results <- list()
checkpoint("starting all cell lines", reset = TRUE)
for (cell_line in fit_cell_lines) {
  result <- run_limma_cell_line(limma_input, cell_line)
  if (!is.null(result)) {
    limma_results[[cell_line]] <- result
  }
}
checkpoint("all cell lines fitted")
report_memory("end of fitting")
```

## Merging the cell lines

Each cell line is fitted **separately**: its own starve control, its own residual variance, its own
`eBayes` prior and its own BH correction. The alternative — one joint model over all 216 runs with
`group = cell line × timepoint` — would buy residual degrees of freedom and would be the right
design for *mutant versus WT* contrasts, but it also assumes one residual variance per site across
all eight cell lines and is not needed for the question this notebook answers ("does this site
respond to EGF, within this cell line?"). Per-cell-line fitting is not df-starved here: 27 runs
against 9 coefficients leave 18 residual df on a fully observed site (16 if the replicate term is
switched on).

The consequence for the merge is that **the site sets differ between cell lines** — a site can pass
the coverage filter in WT and fail it in SHOC2T71A. The merge is therefore a full outer join on
`site`, and a site not tested in a given cell line carries NA in that cell line's columns. That is
the state `filter_by_ffdr()` already expects.

```{r merge-cell-lines}
limma_merged <- stage("merge cell lines on site",
                      Reduce(function(left, right) merge(left, right, by = "site", all = TRUE),
                             limma_results))

cat(sprintf("merged: %d sites x %d columns (%d cell lines)\n",
            nrow(limma_merged), ncol(limma_merged), length(limma_results)))
for (cell_line in names(limma_results)) {
  fdr_col <- paste0(cell_line, "_log2:FFDR_EGF_omnibus")
  cat(sprintf("  %-14s tested in %5d sites (%.1f%% of the merged table) | EGF-responsive at FFDR < 0.05: %d\n",
              cell_line,
              nrow(limma_results[[cell_line]]),
              100 * nrow(limma_results[[cell_line]]) / nrow(limma_merged),
              sum(limma_merged[[fdr_col]] < 0.05, na.rm = TRUE)))
}
```

# Verification

What is checked is that the contrast means what the design says it means, and the identity to check
differs between the two designs — so the check is derived from `BLOCK_ON_REPLICATE` rather than
hard-coded:

- **Unblocked (`~ 0 + group`, the default).** Each group coefficient is the mean of that group's
  *observed* runs, so `limmaFC` must equal `mean(x[timepoint, observed]) − mean(x[starve, observed])`
  exactly — for **every** tested site, sparse ones included. This is the stronger of the two checks:
  it covers the whole table rather than a complete-data corner of it.
- **Blocked (`~ 0 + group + replicate`).** The equality only holds where the site's data is balanced,
  i.e. on fully observed sites, where `limmaFC` must equal the mean of the within-replicate
  differences `x[timepoint, ri] − x[starve, ri]`. On an incomplete site the two quantities *should*
  differ — that is precisely what the block term does — so checking those would fail for the right
  reason and prove nothing.

```{r verify-contrast-definition}
verify_contrast_definition <- function(df,
                                       results,
                                       cell_line,
                                       control = CONTROL,
                                       condition = CONDITIONS[1],
                                       block_on_replicate = BLOCK_ON_REPLICATE,
                                       checks = c("2", "10", "90")) {
  # Check limma's logFC against the difference the fitted design defines it to be.
  #
  # Unblocked: the difference of the observed group means, checked on every tested site.
  # Blocked: the mean of the per-replicate differences, checked on fully observed sites only
  # (where the two definitions coincide; see the section above).
  #
  # Args:
  #   df: the dataset that was fitted, as returned by read_dataset() — reused rather than re-read.
  #   results: results data.frame returned by run_limma_cell_line() for this cell line.
  #   cell_line: cell line column prefix.
  #   control: control timepoint used as the contrast reference.
  #   condition: condition name in the column headers.
  #   block_on_replicate: the design that was fitted; selects which identity is checked.
  #   checks: timepoints to verify.
  #
  # Returns:
  #   Invisibly, the maximum absolute discrepancy found, or NA if no site could be checked.

  own_columns <- grep(paste0("^", cell_line, "_log2:abs_"), colnames(df), value = TRUE)
  replicates  <- unique(sub(".*_(r[0-9]+)$", "\\1", own_columns))

  # The rows the identity is expected to hold on.
  checked <- if (block_on_replicate) {
    df[rowSums(is.na(df[own_columns])) == 0, , drop = FALSE]
  } else {
    df
  }
  checked <- checked[checked$site %in% results$site, , drop = FALSE]

  if (nrow(checked) == 0) {
    cat(sprintf("  %-14s no site available to verify%s\n", cell_line,
                if (block_on_replicate) " (none fully observed)" else ""))
    return(invisible(NA_real_))
  }

  run_columns <- function(timepoint) {
    paste0(cell_line, "_log2:abs_", condition, "_", timepoint, "_", replicates)
  }

  worst <- 0
  for (timepoint in checks) {
    fc_col <- paste0(cell_line, "_log2:limmaFC_", condition, "_", timepoint)
    if (!fc_col %in% colnames(results)) next

    manual <- if (block_on_replicate) {
      rowMeans(as.matrix(checked[run_columns(timepoint)]) -
               as.matrix(checked[run_columns(control)]))
    } else {
      # Observed group means: rowMeans(na.rm = TRUE) is exactly what the unblocked fit estimates.
      rowMeans(checked[run_columns(timepoint)], na.rm = TRUE) -
      rowMeans(checked[run_columns(control)],   na.rm = TRUE)
    }

    from_limma  <- results[[fc_col]][match(checked$site, results$site)]
    comparable  <- is.finite(manual) & is.finite(from_limma)
    discrepancy <- if (any(comparable)) max(abs(manual[comparable] - from_limma[comparable])) else NA_real_
    worst <- max(worst, discrepancy, na.rm = TRUE)

    cat(sprintf("  %-14s %-4s n=%5d  max |limma - %s| = %.3e\n",
                cell_line, timepoint, sum(comparable),
                if (block_on_replicate) "within-replicate mean" else "difference of group means",
                discrepancy))
  }

  invisible(worst)
}

cat(sprintf("design fitted: ~ 0 + group%s\n\n",
            if (BLOCK_ON_REPLICATE) " + replicate" else ""))
for (cell_line in names(limma_results)) {
  worst <- verify_contrast_definition(limma_input, limma_results[[cell_line]], cell_line)
  stopifnot(is.na(worst) || worst < 1e-9)
}
checkpoint("contrast definition verified for all cell lines")
```

A second sanity check: the limma log fold change against the mean-based `log2:FC` the Python pipeline
already computed. What to expect depends on the design, and this is the clearest illustration of what
the block term actually does:

- **Unblocked** — `limmaFC` and `log2:FC` are the *same estimator*. `dia_compute_fold_change()`
  computes `log2:mean(t) − log2:mean(starve)` over the observed replicates, which is exactly the
  difference of observed group means the unblocked fit estimates. Agreement should be exact (~1e-15)
  on **every** site, sparse ones included. limma is then adding only the variance model — the
  moderated *t* and *F*, their p-values and the FDR — not a different fold change.
- **Blocked** — the two agree exactly only where the site's data is balanced. On incomplete sites
  they genuinely diverge, and the size of that divergence *is* the replicate-offset correction. The
  `all` and `complete` numbers printed below separate the two.

```{r verify-against-pipeline}
for (cell_line in names(limma_results)) {
  results <- limma_results[[cell_line]]

  matched   <- match(results$site, limma_input$site)
  pipeline  <- limma_input[[paste0(cell_line, "_log2:FC_", CONDITIONS[1], "_5")]][matched]
  new_value <- results[[paste0(cell_line, "_log2:limmaFC_", CONDITIONS[1], "_5")]]

  own_columns <- grep(paste0("^", cell_line, "_log2:abs_"), colnames(limma_input), value = TRUE)
  is_complete <- rowSums(is.na(limma_input[own_columns])) == 0
  complete    <- is_complete[matched]
  comparable  <- !is.na(pipeline) & !is.na(new_value)

  cat(sprintf("%-14s vs log2:FC at %s 5 min: all r=%.4f (n=%d, max |diff| = %.2e) | complete %s\n",
              cell_line, CONDITIONS[1],
              cor(pipeline, new_value, use = "complete.obs"),
              sum(comparable),
              max(abs(pipeline[comparable] - new_value[comparable])),
              if (sum(complete, na.rm = TRUE) > 2) {
                sprintf("r=%.4f (n=%d, max |diff| = %.2e)",
                        cor(pipeline[complete], new_value[complete], use = "complete.obs"),
                        sum(complete, na.rm = TRUE),
                        max(abs(pipeline[complete] - new_value[complete]), na.rm = TRUE))
              } else "n/a"))
}
```

```{r verify-ranges}
p_cols <- grep("pvalue|FDR", colnames(limma_merged), value = TRUE)
p_cols <- grep("adjusted", p_cols, value = TRUE, invert = TRUE)
values <- unlist(limma_merged[, p_cols], use.names = FALSE)

stopifnot(all(is.na(values) | (values >= 0 & values <= 1)))
cat(sprintf("%d sites x %d columns | all %d p-value / FDR columns within [0, 1]\n",
            nrow(limma_merged), ncol(limma_merged), length(p_cols)))

# With one condition the ALL omnibus is by construction the same test as the EGF omnibus. Asserted
# rather than assumed, because the column is emitted for naming compatibility and a silent
# divergence would mean the contrast sets drifted apart.
for (cell_line in names(limma_results)) {
  egf <- limma_merged[[paste0(cell_line, "_log2:Fpvalue_EGF_omnibus")]]
  all_ <- limma_merged[[paste0(cell_line, "_log2:Fpvalue_ALL_omnibus")]]
  stopifnot(isTRUE(all.equal(egf, all_)))
}
cat("ALL omnibus is identical to the EGF omnibus in every cell line, as expected with one condition\n")
```

# Write the results

Results are written to a **separate** file keyed by `site`, in the same folder the input came from,
never back over the input (`limma_output_path()` raises if the two paths ever collide). Merging them
into the Python frames (`merge_limma_results()` in `src/transformations.py`, or
`pandas.merge(df, limma, on="site", how="left")`) is a downstream step, not part of this notebook.

A run made with `USE_SAMPLE <- TRUE` writes into `data/` and is a development artefact — do not merge
it into a real analysis.

```{r write-results}
out_path <- limma_output_path()
invisible(stage(paste("write", basename(out_path)),
                write.table(limma_merged,
                            file = out_path,
                            sep = "\t",
                            row.names = FALSE,
                            quote = FALSE,
                            na = "NA")))
cat("     wrote", out_path,
    "  (", nrow(limma_merged), "sites x", ncol(limma_merged), "columns,",
    round(file.size(out_path) / 1e6), "MB )\n")
checkpoint("results written")
```

# If the run is slow

The arithmetic is not the expensive part: on the TMT files (68 channels, 20 coefficients, ~25%
missing) `lmFit` plus all `eBayes` fits take ~7 s at 50 000 sites. This dataset fits **eight** models
of 27 runs each, which is smaller per fit, but the per-site loop is used throughout (NAs are
everywhere in DIA), so expect the cost to scale with sites × cell lines. If it takes far longer than
a few minutes, the checkpoints above name the stage — and the likely causes are, in order:

1. **Warning flooding in the IDE.** `lmFit` can emit one warning per site, and with per-run
   missingness nearly every site qualifies. `muffle_warnings()` collects and summarises them; if
   this was the cause, the fix is already in place.
2. **Swapping.** Watch the `memory:` lines. `read.delim` can transiently use several times the file
   size, and the file is read once and reused by all eight fits for that reason.
3. **A much wider file than expected.** The `parsed: N rows x M columns` line says so directly.
4. **Missing values stored as `0` rather than `NA`.** This does not make the run slow — it makes it
   *fast and wrong*, because limma sees a complete matrix and takes its single-QR path. The
   `% exactly zero` diagnostic in `read_dataset()` catches it.

To time it cheaply before committing to a full run, set `MAX_SITES <- 2000` (and optionally
`MAX_CELL_LINES <- 2`) in the parameters chunk. Prefer
`Rscript -e 'rmarkdown::render("limma_for_pvalues_diapasef.rmd")'` over stepping through chunks in
the IDE — the console rendering, not the statistics, is the part that scales badly.

# Output columns

One block per cell line. For each timepoint, contrasted against `starve` (within replicate only if
`BLOCK_ON_REPLICATE` is TRUE):

| Column | Meaning |
|---|---|
| `WT_log2:limmaFC_EGF_5` | log2 fold change vs starve, estimated inside the model. Named `limmaFC` rather than `FC` so it does not collide with the mean-based `log2:FC` from the Python pipeline — under the default unblocked design the two are numerically the same estimator |
| `WT_log2:pvalue_EGF_5` | moderated *t* p-value |
| `WT_log2:FDR_EGF_5` | Benjamini-Hochberg FDR, applied across the sites **tested in that cell line** for that contrast |
| `WT_log2:adjustedFDR_EGF_5` | `-log10(FDR)` |

Omnibus *F*-tests, per site per cell line:

| Column | Meaning |
|---|---|
| `WT_log2:Fpvalue_EGF_omnibus` | is this site different from starve at **any** EGF timepoint? |
| `WT_log2:Fpvalue_ALL_omnibus` | identical to the above while EGF is the only condition; kept so downstream code written for the TMT tables resolves |
| `..._FFDR_..._omnibus`, `..._adjustedFFDR_..._omnibus` | BH-FDR and `-log10(FDR)` of the corresponding F-test |

Two things to keep in mind when reading the table:

- **NA means "not tested", never "not significant".** A site can be NA because it failed the
  coverage filter in that cell line, because that timepoint was never observed, or — for the F
  columns — because *some* timepoint was never observed. Never `.fillna(0)` these columns: 0 reads
  as perfectly significant and every untested site would pass any FDR filter.
- The `full` timepoint gets its own *t* contrast but is deliberately **excluded from the F-tests**:
  it is a media control, not a stimulation timepoint, so including it would let a site qualify as
  "responsive" purely because full media differs from starve.

```{r session-info}
sessionInfo()
```
