---
title: "Trendy — segmented-regression patterns of each phosphosite time course"
output: html_notebook
---

# Purpose

For **one stimulation condition** (EGF by default), describe the shape of every phosphosite's time
course as a sequence of **linear segments** joined at **breakpoints**, using the Bioconductor
package [**Trendy**](https://bioconductor.org/packages/Trendy/) (Bacher *et al.* 2018,
*BMC Bioinformatics* 19:380).

The `full` timepoint is **excluded** — it is a media control, not a point on the post-stimulation
clock. `starve` is kept as time zero, because it is the state the cells are in when EGF is added.

The output is one row per site carrying a **pattern string** (`up-down`, `up-flat`, `flat`, …), the
fitted breakpoint times in minutes, the slope and significance of each segment, and the adjusted
R² of the fit.

## What this adds to what the project already has

| approach | where | what it gives | what it assumes |
|---|---|---|---|
| model-free descriptors | `src/response_shapes.py` | peak time, retention, AUC — read off the grid | nothing, but the timing is **quantised** to the sampled timepoints |
| shape census | `src/response_shapes.py` | one of six classes per site (`transient`, `sustained`, …) | a noise band; classes are rule-based, not fitted |
| anchored sigmoid | `src/curve_fitting.py`, `notebooks/06_sigmoids/` | continuous T50 and amplitude | a **monotone** S-shape — structurally excludes the transient majority |
| **Trendy (this file)** | here | a fitted **piecewise-linear** shape: how many turning points, when, and in which direction | linearity within each segment; enough samples per segment |

Trendy fills the specific gap the sigmoid work left open: it is the first method here that can
**fit** a rise-and-fall — the 69% of EGF-responsive sites that `clustering_method_decision.md` §9.5
recorded as transient or biphasic, and that a logistic cannot represent. It does it without
committing to a parametric peak shape, by letting a breakpoint mark the turn.

It is *not* a replacement for the sigmoid fits. A segmented fit gives a **turning time**; the
sigmoid gives a **half-response time**. Those are different quantities and both are useful.

# What Trendy actually does

## Segmented (breakpoint) regression

Ordinary linear regression fits one straight line, $y = a + bt$, through all the points.
**Segmented regression** fits several straight lines joined end to end:

```
     y                          .-----------      segment 3: slope ~ 0  ("flat")
       |                     .-'
       |                  .-'                     segment 2: slope > 0  ("up")
       |    .------------'
       |  -'                                      segment 1: slope > 0
       +------------------------------------ t
              ^            ^
          breakpoint 1  breakpoint 2
```

- a **breakpoint** (also *changepoint*, and `psi` in the underlying `segmented` package) is a time
  at which the slope changes. It is **fitted**, not chosen from the sampling grid, so it can land
  between two sampled timepoints;
- a **segment** is the stretch between two consecutive breakpoints;
- each segment has a **slope** and a p-value for that slope, and Trendy converts the pair into a
  **trend**: `1` = up, `-1` = down, `0` = no change (slope not significant at `pvalCut`);
- the sequence of trends is the site's **pattern** — the thing this notebook is for.

The fit is continuous: the segments join at the breakpoints rather than jumping. So a transient
response is `up` then `down`, and the breakpoint between them is an estimate of **when the site
turned over**, which no method currently in this project produces.

## How Trendy chooses the number of breakpoints

For each site it fits every model from 0 breakpoints (plain linear regression) up to `maxK`, and
picks between them with the **BIC** (Bayesian Information Criterion):

$$\mathrm{BIC} = -2\log L + p\log n$$

— misfit plus a penalty for the number of parameters $p$, so a breakpoint has to earn its keep.
Lower is better. Two rules then override the BIC choice (they are in `fitSegBIC`, and both matter
on a design this small):

1. every segment must contain at least `minNumInSeg` **samples**, otherwise the number of
   breakpoints is reduced until it does;
2. no breakpoint may fall before the `minNumInSeg`-th sample;
3. if the best breakpoint model still does not beat plain linear regression on BIC, the linear fit
   is returned with `Breakpoints = NA`.

So "no breakpoints" is a real, reachable answer, not a failure.

## ⚠️ How a segment is called `up`, `down` or `flat` — and why the test is unusually strict

A **p-value** is: if this segment's slope were really zero, how often would noise alone produce a
slope at least this steep? Small p = hard to explain as noise. Trendy takes the slope's *t*
statistic (slope divided by its standard error) and converts it with

```r
slp.pval <- pt(-abs(slp.t), 1)        # fitSegBIC(), Trendy source
```

— a one-sided tail of a *t* distribution with **1 degree of freedom**, hard-coded, not the residual
df of the fit (which is ~20 here). A *t* on 1 df is the Cauchy distribution: extremely heavy tails,
so it demands a much larger statistic before it yields a small p. At the default `pvalCut = 0.1` a
segment needs $|t| > 3.08$ to be called `up` or `down`, where the same one-sided test on the fit's
actual 20 df would need only $|t| > 1.33$.

Two consequences to carry into the census:

- `pvalCut = 0.1` here is **not** a 10% test in the usual sense. It is deliberately conservative,
  which is presumably why the package default is 0.1 rather than 0.05. Do not "correct" it to 0.05
  on the reflex that 0.05 is standard — that makes an already strict test stricter.
- **`flat` is a conservative call**, so the flat fraction below is an *upper* bound on the number of
  genuinely non-responding sites. The limma FFDR column, which uses a properly moderated variance
  with the right df, is the better authority on responsiveness, and it is carried into the output
  so the two can be compared directly.

## Adjusted R²

$R^2$ is the fraction of the variance in the data explained by the fit. **Adjusted** $R^2$ subtracts
a penalty for the number of parameters, so it does not automatically rise when a segment is added.
`topTrendy(adjR2Cut = ...)` uses it as the quality filter, exactly as `fit_ok` gates the sigmoid
fits. It is a *fit-quality* measure, not a shape test: a large clean response scores well even if the
segmentation is arguable.

# ⚠️ What this design can and cannot support

**Read this before trusting a pattern.** Trendy was written for ordered profiling experiments with
many time levels — the package vignette uses tens of points. This dataset has **six**
(`starve`, 2, 5, 10, 15, 90 min), and no fitter creates information that the design did not collect.

## The degrees-of-freedom ledger

Per site, per condition: 6 timepoints × 4 replicates = **24 samples**, 6 distinct **time levels**.

| model | free parameters | samples | residual df |
|---|---|---|---|
| linear (0 breakpoints) | 2 | 24 | 22 |
| 1 breakpoint | 4 (2 slopes, intercept, breakpoint) | 24 | 20 |
| 2 breakpoints | 6 | 24 | 18 |

Those residual df look comfortable, and they are **misleading in exactly the way
`council_note_3_temporal_curve_modelling.md` documented for the impulse model**: replicates buy
*pure-error* df, not *lack-of-fit* df. The number that limits what shape can be resolved is the
count of distinct **time levels**, which is 6. Three segments need six time levels at two apiece —
i.e. `maxK = 2` is the arithmetic ceiling and leaves nothing over to check the shape with.

`MAX_K` is therefore capped at 2 here and defaults to **2**, and a 2-breakpoint call should be
treated as descriptive rather than established. The same reasoning is why `minNumInSeg = 5`: with
4 replicates per timepoint, requiring 5 samples per segment forces **at least two distinct
timepoints** into every segment. Lowering it to 4 would let a segment be fitted through a single
timepoint's replicates — a slope estimated from four measurements taken at the same time, which is
not a slope.

## Trendy refuses missing values

`trendy()` stops on the first `NA`. TMT missingness is plex-wise, so the workable filter is
`n:reps == 4` — sites quantified in all four plexes — which is applied below and reported. Sites
with 2–3 plexes are **not** analysable by this method at all; they are absent from the output, and
that absence is not evidence about their behaviour.

## The time axis, and why the default is not minutes

Trendy fits *straight lines in whatever units `tVectIn` carries*, so the axis is a modelling choice
with consequences:

| axis | breakpoints reported in | problem |
|---|---|---|
| `minutes` | minutes directly | the grid is log-spaced, so the 90-min point carries leverage 0.979 (`clustering_method_decision.md`) and one measurement dominates every segment reaching it |
| `log` = $\log_{10}(t+1)$ | log units, back-transformed | Trendy's internal starting values are `round()`ed to integers (`breakpointFit`), and this axis spans only 0–1.96, so every starting guess collapses onto 0/1/2 |
| **`index`** (default) | timepoint index 1–6, interpolated back to minutes | none of the above; the spacing is uniform |

`index` is the default because uniform spacing on this particular grid is **near-identical to
log-time weighting** — `clustering_method_decision.md` measured $r = 0.987$ between the two,
because `{0, 2, 5, 10, 15, 90}` is already approximately log-spaced. It also matches how Trendy was
designed to be used ("ordered conditions"), and keeps the integer starting values sensible.

Breakpoints come back on the index axis and are converted to minutes by interpolating between the
flanking timepoints (`BREAKPOINT_INTERP`, in log-minute space by default, consistent with the axis
choice). Both are reported, so nothing is hidden behind the conversion.

## Two further caveats

- **No between-sample normalisation upstream.** The pipeline applies none (CLAUDE.md → Known
  issues; `NORMALIZE <- "none"` in `limma_for_pvalues.rmd`). Since Trendy fits each site
  separately, a *site-constant* offset is harmless — it is absorbed by the intercept. What is not
  harmless is a **per-channel** offset, which enters as slope-distorting noise. `NORMALIZE` and
  `PLEX_CENTER` below both address it and are reported side by side.
- **Trendy has no blocking term.** `limma_for_pvalues.rmd` fits `~ 0 + group + plex` because each
  replicate is one TMT plex with its own offset. Trendy cannot express that, so `PLEX_CENTER`
  removes the per-plex offset from each site before fitting. This is the segmented-regression
  equivalent of limma's `plex` term and it costs 3 df per site (four plex means constrained to the
  site mean).

# Installation

Trendy is **not** installed in the environment this file was written against (R 4.6.1). Install it
once, then re-run:

```{r install-trendy, eval=FALSE}
# Run once, interactively. Not evaluated when the notebook is knitted.
# if (!requireNamespace("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }
# BiocManager::install("Trendy")
```

# Parameters

Paths are resolved from the **project root**, located by walking up from the working directory,
exactly as in `limma_for_pvalues.rmd` — knitting runs with this file's folder as the working
directory while the IDE console runs from the project root, and no single relative path serves
both.

```{r setup}
suppressPackageStartupMessages({
  library(Trendy)
})

find_project_root <- function(start = getwd(),
                              markers = c("data", "src", "notebooks")) {
  # Locate the project root by walking up the directory tree from `start`.
  #
  # Args:
  #   start: directory to start searching from, default the current working directory.
  #   markers: directory names that must all be present for a folder to count as the root.
  #
  # Returns:
  #   Absolute path to the project root. Raises an error if no ancestor directory qualifies.

  path <- normalizePath(start, mustWork = TRUE)

  repeat {
    if (all(dir.exists(file.path(path, markers)))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("find_project_root: no ancestor of '", start, "' contains all of: ",
           paste(markers, collapse = ", "),
           ". Set PROJECT_ROOT manually to the TMT_Data_analysis folder.")
    }
    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()
DATA_DIR     <- file.path(PROJECT_ROOT, "Experiment")
PROCESSED_SUBDIR <- file.path("Data", "Processed")

# ── What to analyse ─────────────────────────────────────────────────────────────────────
DATASET   <- "hme1_2"      # dataset key = folder name under Experiment/
CELL_LINE <- "WT"
CONDITION <- "EGF"         # ONE condition per run; rerun for INS / EGFnINS and compare
CONTROL   <- "starve"      # kept as t = 0
EXCLUDE   <- "full"        # media control — never a point on the post-stimulation clock

# Candidate input files, tried in order. The first that exists and carries the required
# log2:abs columns is used, so the notebook survives the processed file being renamed.
INPUT_CANDIDATES <- c("20260807_hTERT_HEM1_2_processed_phPlus.tsv",
                      "20260807_hTERT_HEM1_2_log2_processed.tsv")

# ── Which sites are eligible ────────────────────────────────────────────────────────────
REQUIRE_COMPLETE <- TRUE   # Trendy cannot accept NA — see the caveats above
MIN_REPS         <- 4      # n:reps; must be the full plex count while REQUIRE_COMPLETE is TRUE
USE_FFDR         <- TRUE   # restrict to limma-responsive sites (much faster, and interpretable)
MAX_FFDR         <- 0.05   # BH-FDR of the omnibus F-test for this condition

# ── Pre-processing applied before fitting ───────────────────────────────────────────────
NORMALIZE   <- "none"      # "none" | "median" — between-sample (column) centring
PLEX_CENTER <- TRUE        # remove each replicate's per-site offset (limma's `plex` term)

# ── The Trendy fit ──────────────────────────────────────────────────────────────────────
TIME_AXIS          <- "index"   # "index" | "minutes" | "log"  — see the caveats above
BREAKPOINT_INTERP  <- "log"     # "log" | "linear" — how an index breakpoint becomes minutes
MAX_K              <- 2         # maximum breakpoints. 2 is the arithmetic ceiling here
MIN_NUM_IN_SEG     <- 5         # samples per segment; 5 forces >= 2 distinct timepoints
PVAL_CUT           <- 0.10      # slope p-value below which a segment counts as up/down
NUM_TRY            <- 5         # optimiser restarts inside `segmented`
MEAN_CUT           <- -Inf      # Trendy's own low-expression filter; disabled, we filter ourselves
ADJ_R2_CUT         <- 0.50      # quality gate, applied as a flag rather than as a hard filter
NCORES             <- 1         # 1 = reproducible (see the note on jitter below)
SEED               <- 1

# ── Output ──────────────────────────────────────────────────────────────────────────────
OUT_SUFFIX   <- paste0("_trendy_", CONDITION, "_patterns.tsv")
SAVE_TABLE   <- TRUE
SAVE_FIGURES <- FALSE

# ── Debugging ───────────────────────────────────────────────────────────────────────────
MAX_SITES <- Inf   # e.g. 500 to smoke-test the whole file in seconds
VERBOSE   <- TRUE

cat("Trendy version   :", as.character(packageVersion("Trendy")), "\n")
cat("segmented version:", as.character(packageVersion("segmented")), "\n")
cat("project root     :", PROJECT_ROOT, "\n")
```

## Resolving the input file

```{r resolve-input}
resolve_input_file <- function(dataset,
                               candidates,
                               data_dir = DATA_DIR,
                               subdir = PROCESSED_SUBDIR) {
  # Pick the first candidate processed file that exists for this dataset.
  #
  # Args:
  #   dataset: dataset key, e.g. "hme1_2"; also the folder name under Experiment/.
  #   candidates: character vector of file names to try, in priority order.
  #   data_dir: the Experiment/ directory.
  #   subdir: path of the processed folder inside the dataset folder.
  #
  # Returns:
  #   Absolute path to the first existing candidate. Raises listing the folder if none exists.

  folder <- file.path(data_dir, dataset, subdir)
  if (!dir.exists(folder)) {
    stop("resolve_input_file: '", folder, "' does not exist.")
  }

  for (candidate in candidates) {
    path <- file.path(folder, candidate)
    if (file.exists(path)) {
      return(path)
    }
  }

  stop("resolve_input_file: none of the candidates exist in '", folder, "'.\n",
       "  tried   : ", paste(candidates, collapse = ", "), "\n",
       "  present : ", paste(list.files(folder, pattern = "\\.tsv$"), collapse = ", "))
}


trendy_output_path <- function(input_path,
                               out_suffix = OUT_SUFFIX) {
  # Build the output path next to the input, reusing the input's date-prefixed stem.
  #
  # Args:
  #   input_path: path of the processed .tsv that was read.
  #   out_suffix: suffix appended to the stem, including the extension.
  #
  # Returns:
  #   Absolute path of the .tsv to write. Raises if it would collide with the input.

  stem <- basename(input_path)
  stem <- sub("_processed_phPlus\\.tsv$", "", stem)
  stem <- sub("_log2_processed\\.tsv$", "", stem)
  stem <- sub("\\.tsv$", "", stem)

  output_path <- file.path(dirname(input_path), paste0(stem, out_suffix))
  if (identical(output_path, input_path)) {
    stop("trendy_output_path: refusing to overwrite the input file '", input_path, "'.")
  }
  output_path
}

INPUT_FILE  <- resolve_input_file(DATASET, INPUT_CANDIDATES)
OUTPUT_FILE <- trendy_output_path(INPUT_FILE)

cat("input : ", INPUT_FILE, "\n")
cat("output: ", OUTPUT_FILE, "\n")
```

# Progress checkpoints

Same pattern as `limma_for_pvalues.rmd`: every stage announces itself **before** it runs, so a run
that stalls leaves the name of the stage it is stuck in as the last line printed. Trendy is the
slow part of this notebook — it fits `MAX_K + 1` regression models per site — so knowing which
stage is running matters.

```{r checkpoints}
.timing <- new.env(parent = emptyenv())

stage <- function(label,
                  expr) {
  # Run one stage, announcing it before evaluation and timing it after.
  #
  # `expr` is passed unevaluated and forced inside, so the "starting" line is printed and flushed
  # before any work begins.
  #
  # Args:
  #   label: short description of the stage.
  #   expr: the expression to evaluate.
  #
  # Returns:
  #   The value of `expr`.

  if (VERBOSE) {
    cat(sprintf("[%s]   -> %-46s ", format(Sys.time(), "%H:%M:%S"), paste0(label, " ...")))
    flush.console()
  }

  started <- proc.time()[["elapsed"]]
  value   <- force(expr)
  elapsed <- proc.time()[["elapsed"]] - started

  if (VERBOSE) {
    cat(sprintf("done in %7.1fs\n", elapsed))
    flush.console()
  }
  value
}
```

# Functions

```{r functions}
sort_timepoints <- function(timepoints,
                            control = CONTROL) {
  # Order timepoint labels biologically: the control first, then the numeric points ascending.
  #
  # Args:
  #   timepoints: character vector of timepoint labels, e.g. c("10", "2", "starve").
  #   control: label of the control timepoint, placed first.
  #
  # Returns:
  #   The labels in plotting / fitting order.

  numeric_points <- setdiff(timepoints, control)
  c(control[control %in% timepoints],
    numeric_points[order(as.numeric(numeric_points))])
}


timepoint_minutes <- function(labels,
                              control = CONTROL) {
  # Translate timepoint labels into minutes, with the control at 0.
  #
  # Args:
  #   labels: character vector of timepoint labels.
  #   control: label mapped to 0.
  #
  # Returns:
  #   Numeric vector of minutes, same length and order as `labels`.

  minutes <- suppressWarnings(as.numeric(labels))
  minutes[labels == control] <- 0
  if (anyNA(minutes)) {
    stop("timepoint_minutes: cannot convert to minutes: ",
         paste(labels[is.na(minutes)], collapse = ", "))
  }
  minutes
}


parse_condition_columns <- function(column_names,
                                    cell_line = CELL_LINE,
                                    condition = CONDITION,
                                    exclude = EXCLUDE,
                                    control = CONTROL) {
  # Find and parse the log2:abs replicate columns of one cell line and condition.
  #
  # The `full` timepoint is dropped here rather than downstream, so nothing later has to remember
  # to exclude it. Columns are returned sorted onto the minute axis (control first) — the on-disk
  # order is not trusted, because Trendy assumes `tVectIn` is ascending.
  #
  # Args:
  #   column_names: all column names of the dataset.
  #   cell_line: cell line prefix, e.g. "WT".
  #   condition: single condition name, e.g. "EGF".
  #   exclude: timepoint label to drop entirely, e.g. "full".
  #   control: timepoint label kept as t = 0.
  #
  # Returns:
  #   data.frame with one row per sample column, ordered by time then replicate, and the columns
  #   `column`, `timepoint`, `plex`, `minutes`.

  pattern <- paste0("^", cell_line, "_log2:abs_", condition, "_([^_]+)_(r[0-9]+)$")
  matched <- grep(pattern, column_names, value = TRUE)
  if (length(matched) == 0) {
    stop("parse_condition_columns: no log2:abs columns matched '", pattern, "'.")
  }

  fields <- regmatches(matched, regexec(pattern, matched))
  info <- data.frame(column    = matched,
                     timepoint = vapply(fields, `[`, character(1), 2),
                     plex      = vapply(fields, `[`, character(1), 3),
                     stringsAsFactors = FALSE)

  info <- info[!info$timepoint %in% exclude, , drop = FALSE]
  info$minutes <- timepoint_minutes(info$timepoint, control)
  info <- info[order(info$minutes, info$plex), , drop = FALSE]
  rownames(info) <- NULL
  info
}


normalize_matrix <- function(M,
                             method) {
  # Apply between-sample (column) normalisation to the log2 matrix.
  #
  # Args:
  #   M: numeric matrix, sites in rows and samples in columns, on the log2 scale.
  #   method: "none" (unchanged) or "median" (centre every column on the grand median).
  #
  # Returns:
  #   The normalised matrix, same dimensions and dimnames.

  if (method == "none") {
    return(M)
  }
  if (method == "median") {
    column_medians <- apply(M, 2, median, na.rm = TRUE)
    return(sweep(M, 2, column_medians) + mean(column_medians))
  }
  stop("normalize_matrix: unknown method '", method, "'.")
}


center_plex <- function(M,
                        plex) {
  # Remove each replicate's per-site offset — the segmented-regression stand-in for limma's
  # `plex` blocking term.
  #
  # Each TMT plex has its own overall intensity offset. limma absorbs it with a `plex` factor;
  # Trendy has no blocking term, so a plex that ran 0.4 log2 units high adds a step to the
  # residuals of every site and flattens real slopes. For each site and plex the mean over that
  # plex's timepoints is subtracted and the site's grand mean added back, so the level of each
  # site is preserved while the between-plex offsets are removed.
  #
  # Costs 3 degrees of freedom per site (four plex means constrained to the site mean); with 24
  # samples that is affordable, and it is reported in the output as `plex_centered`.
  #
  # Args:
  #   M: numeric matrix, sites in rows, samples in columns, no NA.
  #   plex: character or factor vector giving the plex of each column of M.
  #
  # Returns:
  #   The centred matrix, same dimensions and dimnames.

  plex <- as.character(plex)
  grand <- rowMeans(M)
  out <- M

  for (level in unique(plex)) {
    columns <- which(plex == level)
    offset  <- rowMeans(M[, columns, drop = FALSE]) - grand
    out[, columns] <- M[, columns, drop = FALSE] - offset
  }
  out
}


build_time_vector <- function(sample_info,
                              axis = TIME_AXIS) {
  # Build the numeric time vector Trendy fits against, on the requested axis.
  #
  # Args:
  #   sample_info: data.frame from parse_condition_columns(), already time-ordered.
  #   axis: "index" (1..T by timepoint order), "minutes" (raw), or "log" (log10(minutes + 1)).
  #
  # Returns:
  #   Named numeric vector, one entry per sample column, names = column names.

  minutes_levels <- sort(unique(sample_info$minutes))

  values <- switch(axis,
                   index   = match(sample_info$minutes, minutes_levels),
                   minutes = sample_info$minutes,
                   log     = log10(sample_info$minutes + 1),
                   stop("build_time_vector: unknown axis '", axis, "'."))

  values <- as.numeric(values)
  names(values) <- sample_info$column
  values
}


axis_to_minutes <- function(x,
                            minutes_levels,
                            axis = TIME_AXIS,
                            interp = BREAKPOINT_INTERP) {
  # Convert a fitted breakpoint from the fitting axis back to minutes.
  #
  # On the "index" axis a breakpoint at, say, 3.4 sits 40% of the way from the 3rd to the 4th
  # timepoint. "40% of the way" needs a scale: `interp = "log"` interpolates in log10(t + 1),
  # which is consistent with the reason the index axis was chosen (uniform index spacing is
  # near-equivalent to log-time on this grid), while "linear" interpolates in raw minutes and
  # will place the same breakpoint much later.
  #
  # Args:
  #   x: numeric vector of breakpoints on the fitting axis (may contain NA).
  #   minutes_levels: sorted unique timepoints in minutes, e.g. c(0, 2, 5, 10, 15, 90).
  #   axis: the axis the fit used.
  #   interp: "log" or "linear", used only for the "index" axis.
  #
  # Returns:
  #   Numeric vector of breakpoint times in minutes, same length as x.

  if (axis == "minutes") {
    return(x)
  }
  if (axis == "log") {
    return(10 ^ x - 1)
  }
  if (axis != "index") {
    stop("axis_to_minutes: unknown axis '", axis, "'.")
  }

  index_levels <- seq_along(minutes_levels)
  if (interp == "linear") {
    return(approx(index_levels, minutes_levels, xout = x, rule = 2)$y)
  }
  if (interp == "log") {
    logged <- approx(index_levels, log10(minutes_levels + 1), xout = x, rule = 2)$y
    return(10 ^ logged - 1)
  }
  stop("axis_to_minutes: unknown interp '", interp, "'.")
}


trend_labels <- function(trends) {
  # Translate Trendy's numeric trend codes into words.
  #
  # Args:
  #   trends: numeric vector of -1 / 0 / 1, possibly padded with NA.
  #
  # Returns:
  #   Character vector of "down" / "flat" / "up", with NA preserved.

  out <- rep(NA_character_, length(trends))
  out[which(trends ==  1)] <- "up"
  out[which(trends == -1)] <- "down"
  out[which(trends ==  0)] <- "flat"
  out
}


trendy_result_table <- function(fit_list,
                                minutes_levels,
                                max_k = MAX_K,
                                adj_r2_cut = ADJ_R2_CUT) {
  # Flatten Trendy's per-site list into one tidy row per fitted site.
  #
  # Built from the raw fit list rather than from topTrendy() on purpose: topTrendy drops every
  # site below the adjusted-R2 cutoff, and a dropped site is not the same thing as a site with no
  # pattern. Here every fitted site gets a row and the cutoff becomes a flag, matching how the
  # sigmoid notebook reports `fit_ok` alongside the sites it excludes.
  #
  # Args:
  #   fit_list: the list returned by results(trendy(...)), one element per site.
  #   minutes_levels: sorted unique timepoints in minutes, for the breakpoint conversion.
  #   max_k: maximum number of breakpoints, fixing the width of the output table.
  #   adj_r2_cut: adjusted-R2 threshold recorded in `r2_ok`.
  #
  # Returns:
  #   data.frame with one row per site: identity, pattern, breakpoints (axis units and minutes),
  #   per-segment slopes / trends / p-values, adjusted R2 and quality flags.

  n_seg_max <- max_k + 1

  rows <- lapply(names(fit_list), function(site) {
    fit <- fit_list[[site]]

    trends <- as.numeric(fit$Segment.Trends)
    slopes <- as.numeric(fit$Segment.Slopes)
    pvals  <- as.numeric(fit$Segment.Pvalues)

    breaks <- fit$Breakpoints
    breaks <- if (all(is.na(breaks))) numeric(0) else as.numeric(breaks[!is.na(breaks)])

    words <- trend_labels(trends)
    row <- data.frame(site           = site,
                      n_segments     = length(trends),
                      n_breakpoints  = length(breaks),
                      pattern        = paste(words, collapse = "-"),
                      first_trend    = words[1],
                      last_trend     = words[length(words)],
                      responsive     = any(trends != 0),
                      adj_r2         = as.numeric(fit$AdjustedR2),
                      stringsAsFactors = FALSE)

    # Fixed-width columns, padded with NA, so every site has the same schema regardless of how
    # many segments its own fit ended up with.
    for (i in seq_len(n_seg_max)) {
      row[[paste0("segment", i, "_slope")]]  <- if (i <= length(slopes)) slopes[i] else NA_real_
      row[[paste0("segment", i, "_trend")]]  <- if (i <= length(words))  words[i]  else NA_character_
      row[[paste0("segment", i, "_pvalue")]] <- if (i <= length(pvals))  pvals[i]  else NA_real_
    }
    for (i in seq_len(max_k)) {
      value <- if (i <= length(breaks)) breaks[i] else NA_real_
      row[[paste0("breakpoint", i, "_axis")]] <- value
      row[[paste0("breakpoint", i, "_min")]]  <- axis_to_minutes(value, minutes_levels)
    }
    row
  })

  out <- do.call(rbind, rows)
  out$r2_ok      <- out$adj_r2 >= adj_r2_cut
  out$pattern_ok <- out$r2_ok & out$responsive
  rownames(out) <- NULL
  out
}


pattern_census <- function(patterns,
                           groups = NULL) {
  # Tabulate how many sites follow each pattern, and what fraction that is.
  #
  # The direct analogue of `shape_census` in src/response_shapes.py, so the two summaries can be
  # read side by side.
  #
  # Args:
  #   patterns: character vector of pattern strings, one per site.
  #   groups: optional named list of logical vectors defining subsets to report separately.
  #
  # Returns:
  #   data.frame with one row per pattern: counts and percentages, most frequent first.

  counts <- sort(table(patterns), decreasing = TRUE)
  out <- data.frame(pattern = names(counts),
                    n       = as.integer(counts),
                    pct     = round(100 * as.integer(counts) / length(patterns), 1),
                    stringsAsFactors = FALSE)

  if (!is.null(groups)) {
    for (group_name in names(groups)) {
      subset_counts <- table(patterns[groups[[group_name]]])
      out[[paste0("n_", group_name)]] <- as.integer(subset_counts[out$pattern])
      out[[paste0("pct_", group_name)]] <-
        round(100 * as.integer(subset_counts[out$pattern]) / sum(groups[[group_name]]), 1)
    }
  }
  out[is.na(out)] <- 0
 # out
}
```

# Sanity checks

Five checks on synthetic data with a known answer, before any real site is fitted. Each one would
fail loudly if a wiring assumption were wrong, and together they are the reason the patterns below
can be believed. **Synthetic data** = a curve whose true shape is known, plus realistic noise;
recovering the known answer is the only way to test an estimator, since real data has no known
answer.

```{r sanity-checks}
set.seed(SEED)

# The design this notebook actually fits: 6 timepoints x 4 replicates, on the index axis.
.minutes_levels <- c(0, 2, 5, 10, 15, 90)
.n_rep          <- 4
.sample_minutes <- rep(.minutes_levels, each = .n_rep)
.tvect          <- rep(seq_along(.minutes_levels), each = .n_rep)
names(.tvect)   <- paste0("s", seq_along(.tvect))
.sigma          <- 0.184   # measured across-replicate SD in hme1_2

# 1. The axis conversion is an exact round trip on the sampled timepoints themselves.
stopifnot(all.equal(axis_to_minutes(seq_along(.minutes_levels), .minutes_levels, "index", "log"),
                    .minutes_levels))
cat("axis round trip  : index i -> minutes[i], exactly\n")

# 2. A breakpoint between two timepoints lands between them, and the two interpolations differ
#    in the direction the caveat above predicts (log earlier than linear).
.log_mid    <- axis_to_minutes(4.5, .minutes_levels, "index", "log")
.linear_mid <- axis_to_minutes(4.5, .minutes_levels, "index", "linear")
stopifnot(.log_mid > 10, .log_mid < 15, .linear_mid > 10, .linear_mid < 15,
          .log_mid < .linear_mid)
cat(sprintf("interpolation    : index 4.5 -> %.2f min (log) vs %.2f min (linear)\n",
            .log_mid, .linear_mid))

# 3. A known up-then-down site must be recovered as a turning pattern, with the turn placed
#    between the sampled timepoints that bracket the true peak (index 3 = 5 min).
.true_profile <- c(0, 0.8, 1.5, 1.1, 0.8, 0.2)
.transient    <- 18 + rep(.true_profile, each = .n_rep) + rnorm(length(.tvect), 0, .sigma)

# 4. A monotone site must not acquire a spurious reversal.
.monotone <- 18 + rep(c(0, 0.3, 0.7, 1.1, 1.4, 1.9), each = .n_rep) + rnorm(length(.tvect), 0, .sigma)

# 5. THE FALSE-POSITIVE CHECK. Flat sites must not acquire patterns. One flat site would be a
#    coin toss, so 40 of them are pushed through and the *rate* is checked — the quantity that
#    actually matters for the census below. Because the segment p-value is computed on 1 df (see
#    the section above), the observed rate should come out well BELOW pvalCut, not at it.
.n_flat <- 40
.flat <- t(vapply(seq_len(.n_flat),
                  function(i) 18 + rnorm(length(.tvect), 0, .sigma),
                  numeric(length(.tvect))))
rownames(.flat) <- paste0("flat", seq_len(.n_flat))

.check_matrix <- rbind(transient = .transient, monotone = .monotone, .flat)
colnames(.check_matrix) <- names(.tvect)

.check_fit <- Trendy::trendy(Data        = .check_matrix,
                             tVectIn     = .tvect,
                             maxK        = MAX_K,
                             minNumInSeg = MIN_NUM_IN_SEG,
                             pvalCut     = PVAL_CUT,
                             numTry      = NUM_TRY,
                             meanCut     = MEAN_CUT,
                             NCores      = 1)
.check_tbl <- trendy_result_table(Trendy::results(.check_fit), .minutes_levels)
print(.check_tbl[.check_tbl$site %in% c("transient", "monotone"),
                 c("site", "pattern", "n_breakpoints", "breakpoint1_min", "adj_r2")],
      row.names = FALSE)

.transient_pattern <- .check_tbl$pattern[.check_tbl$site == "transient"]
.monotone_pattern  <- .check_tbl$pattern[.check_tbl$site == "monotone"]
.transient_bp      <- .check_tbl$breakpoint1_min[.check_tbl$site == "transient"]
stopifnot(grepl("up", .transient_pattern), grepl("down", .transient_pattern))
stopifnot(!grepl("down", .monotone_pattern))
stopifnot(.transient_bp > 2, .transient_bp < 15)
cat(sprintf("shape recovery   : transient -> '%s', turn at %.1f min (true peak 5 min)\n",
            .transient_pattern, .transient_bp))
cat(sprintf("                   monotone  -> '%s', no spurious reversal\n", .monotone_pattern))

.flat_rows    <- grepl("^flat", .check_tbl$site)
.false_rate   <- mean(.check_tbl$responsive[.flat_rows])
.false_pattern <- mean(.check_tbl$pattern[.flat_rows] != "flat")
cat(sprintf("flat-site null   : %.1f%% of %d pure-noise sites got a trend, %.1f%% a non-flat pattern\n",
            100 * .false_rate, .n_flat, 100 * .false_pattern))
cat(sprintf("                   (pvalCut = %.2f is an upper bound: the slope test uses 1 df)\n",
            PVAL_CUT))
stopifnot(.false_rate < 4 * PVAL_CUT)
```

⚠️ **On reproducibility.** `fitSegBIC` calls `jitter()` on the time vector whenever there are
replicated times — which is always here — so two runs of the same data give breakpoints that differ
by a small amount. With `NCORES = 1` and a `set.seed` this is reproducible; with `NCORES > 1`,
`BiocParallel` is registered by Trendy itself without an RNG seed, so it is **not**. The jitter is
about 0.02 of a time unit, so the effect on any conclusion is negligible, but a breakpoint quoted to
three decimals is quoting noise.

# Read the data

```{r read-data}
df <- stage(sprintf("read.delim (%.0f MB)", file.size(INPUT_FILE) / 1e6),
            read.delim(INPUT_FILE,
                       sep              = "\t",
                       check.names      = FALSE,
                       na.strings       = c("NA", "NaN", ""),
                       stringsAsFactors = FALSE))

cat(sprintf("parsed: %d sites x %d columns\n", nrow(df), ncol(df)))
stopifnot(!anyDuplicated(df$site), "n:reps" %in% colnames(df))

samples <- parse_condition_columns(colnames(df))
cat(sprintf("samples for %s %s: %d columns = %d timepoints x %d replicates\n",
            CELL_LINE, CONDITION, nrow(samples),
            length(unique(samples$minutes)), length(unique(samples$plex))))
cat("timepoints (min):", paste(sort(unique(samples$minutes)), collapse = ", "),
    sprintf("   ('%s' excluded)\n", EXCLUDE))

# Missing values must be NA, not 0. A fillna(0) that reached disk would be read here as a real
# log2 intensity of 0 against a background of ~18, which Trendy would happily fit a giant
# breakpoint to. This is the defect recorded in clustering_method_decision.md section 1.
.values <- as.matrix(df[, samples$column, drop = FALSE])
.n_zero <- sum(.values == 0, na.rm = TRUE)
cat(sprintf("log2:abs cells: %.1f%% NA, %.1f%% exactly zero\n",
            100 * sum(is.na(.values)) / length(.values),
            100 * .n_zero / length(.values)))
if (.n_zero > 0) {
  cat("  WARNING: exact zeros in log2:abs — if these are missing values written as 0,\n",
      "           fix upstream before trusting any pattern below.\n", sep = "")
}
```

## Selecting the sites

Two filters, for two different reasons.

**`n:reps == 4` is not optional.** `trendy()` stops on the first `NA`, and TMT missingness is
plex-wise, so a site seen in three plexes is missing an entire block of six samples. Sites below the
threshold are absent from the output entirely; that absence says nothing about their biology.

**The limma responsiveness filter is optional but recommended.** Segmenting a site that does not
respond produces a `flat` pattern, which is correct but adds nothing, and Trendy is the slow step.
`USE_FFDR` restricts the fit to sites the omnibus F-test already called responsive for this
condition — the same gate `Clustering.ipynb` and the sigmoid notebook use, so the three analyses
describe the same set of sites.

```{r select-sites}
ffdr_column <- paste0(CELL_LINE, "_log2:FFDR_", CONDITION, "_omnibus")

eligible <- !is.na(df[["n:reps"]]) & df[["n:reps"]] >= MIN_REPS
cat(sprintf("n:reps >= %d                      : %d of %d sites\n",
            MIN_REPS, sum(eligible), nrow(df)))

if (REQUIRE_COMPLETE) {
  complete <- stats::complete.cases(.values)
  cat(sprintf("+ complete across all %2d samples  : %d\n", nrow(samples), sum(eligible & complete)))
  eligible <- eligible & complete
}

if (USE_FFDR) {
  if (!ffdr_column %in% colnames(df)) {
    stop("select-sites: '", ffdr_column, "' is not in the input file. Run ",
         "limma_for_pvalues.rmd and merge its output, or set USE_FFDR <- FALSE.")
  }
  responsive <- !is.na(df[[ffdr_column]]) & df[[ffdr_column]] < MAX_FFDR
  cat(sprintf("+ limma FFDR < %.2f               : %d   <- fitted\n",
              MAX_FFDR, sum(eligible & responsive)))
  eligible <- eligible & responsive
}

if (is.finite(MAX_SITES) && sum(eligible) > MAX_SITES) {
  eligible[which(eligible)[-seq_len(MAX_SITES)]] <- FALSE
  cat("  NOTE: MAX_SITES is set — only", MAX_SITES, "sites will be fitted.\n")
}
stopifnot(sum(eligible) > 0)
```

## Building the matrix Trendy fits

```{r build-matrix}
M <- as.matrix(df[eligible, samples$column, drop = FALSE])
rownames(M) <- df$site[eligible]
storage.mode(M) <- "double"
stopifnot(!anyNA(M))

M <- stage(paste0("normalise (", NORMALIZE, ")"), normalize_matrix(M, NORMALIZE))

if (PLEX_CENTER) {
  before <- apply(M, 2, median)
  M <- stage("centre plex offsets", center_plex(M, samples$plex))
  after <- apply(M, 2, median)
  cat(sprintf("     per-channel median spread: %.3f log2 before -> %.3f after\n",
              diff(range(before)), diff(range(after))))
}

tvect <- build_time_vector(samples, TIME_AXIS)
minutes_levels <- sort(unique(samples$minutes))

cat(sprintf("matrix: %d sites x %d samples | axis '%s', values %s\n",
            nrow(M), ncol(M), TIME_AXIS,
            paste(sprintf("%g", sort(unique(tvect))), collapse = ", ")))
stopifnot(!is.unsorted(tvect))   # Trendy assumes tVectIn ascends
```

# Run Trendy

`maxK` is capped at whatever this design can actually support, and Trendy applies its own caps on
top (`floor(nSamples / minNumInSeg) - 1`, and `nUniqueTimes - 2`) — both are reported so it is
visible when the requested value was not the one used.

```{r run-trendy}
max_k_supported <- min(MAX_K,
                       floor(ncol(M) / MIN_NUM_IN_SEG) - 1,
                       length(minutes_levels) - 2)
if (max_k_supported < MAX_K) {
  cat(sprintf("NOTE: MAX_K reduced from %d to %d by the design ",
              MAX_K, max_k_supported),
      sprintf("(%d samples, %d time levels, minNumInSeg = %d)\n",
              ncol(M), length(minutes_levels), MIN_NUM_IN_SEG))
}
stopifnot(max_k_supported >= 1)

set.seed(SEED)
trendy_out <- stage(sprintf("Trendy: %d sites, maxK = %d", nrow(M), max_k_supported),
                    Trendy::trendy(Data        = M,
                                   tVectIn     = tvect,
                                   maxK        = max_k_supported,
                                   minNumInSeg = MIN_NUM_IN_SEG,
                                   pvalCut     = PVAL_CUT,
                                   numTry      = NUM_TRY,
                                   meanCut     = MEAN_CUT,
                                   NCores      = NCORES))

trendy_fits <- Trendy::results(trendy_out)
cat(sprintf("fitted %d sites\n", length(trendy_fits)))
```

```{r assemble}
patterns <- stage("assemble the pattern table",
                  trendy_result_table(trendy_fits, minutes_levels, max_k_supported, ADJ_R2_CUT))

cat(sprintf("\nsites fitted                 : %d\n", nrow(patterns)))
cat(sprintf("with at least one breakpoint : %d (%.1f%%)\n",
            sum(patterns$n_breakpoints > 0),
            100 * mean(patterns$n_breakpoints > 0)))
cat(sprintf("adjusted R2 >= %.2f           : %d (%.1f%%)\n",
            ADJ_R2_CUT, sum(patterns$r2_ok), 100 * mean(patterns$r2_ok)))
cat(sprintf("non-flat AND R2 gate passed  : %d (%.1f%%)\n",
            sum(patterns$pattern_ok), 100 * mean(patterns$pattern_ok)))

head(patterns[, c("site", "pattern", "n_breakpoints", "breakpoint1_min",
                  "breakpoint2_min", "adj_r2", "pattern_ok")], 10)
```

# The pattern census

The headline result: how many sites follow each temporal pattern. Read it the way the shape census
in `src/response_shapes.py` is read — the point is the *distribution*, and the fraction of sites a
monotone model could not have described.

```{r pattern-census}
census <- pattern_census(patterns$pattern,
                         groups = list(r2_ok = patterns$r2_ok))
print(census, row.names = FALSE)

# The comparison that matters against the sigmoid work: how many sites turn over?
turning <- grepl("up-down|down-up", patterns$pattern)
cat(sprintf("\nsites with a direction reversal (up-down or down-up): %d (%.1f%%)\n",
            sum(turning), 100 * mean(turning)))
cat("  ^ these are the sites a monotone sigmoid structurally cannot represent\n")
cat(sprintf("sites fitted as a single flat segment                 : %d (%.1f%%)\n",
            sum(patterns$pattern == "flat"), 100 * mean(patterns$pattern == "flat")))
```

```{r census-plot, fig.width=8, fig.height=4.5}
top <- head(census[order(-census$n), ], 12)
old_par <- par(mar = c(9, 4, 3, 1))
barplot(top$n,
        names.arg = top$pattern,
        las       = 2,
        col       = "steelblue",
        ylab      = "sites",
        main      = sprintf("Trendy pattern census — %s %s (n = %d)",
                            CELL_LINE, CONDITION, nrow(patterns)))
par(old_par)
```

## When do sites turn over?

The breakpoint distribution is the quantity no other method in this project produces: a *fitted*
turning time, not a grid position. The sampled timepoints are drawn on the plot, because a
breakpoint distribution that simply reproduces the sampling grid is telling you about the design
rather than about the biology.

```{r breakpoint-distribution, fig.width=9, fig.height=4}
bp_minutes <- c(patterns$breakpoint1_min, patterns$breakpoint2_min)
bp_minutes <- bp_minutes[is.finite(bp_minutes)]

old_par <- par(mfrow = c(1, 2))
hist(bp_minutes,
     breaks = 40,
     col    = "grey70",
     xlab   = "breakpoint (min)",
     main   = "All breakpoints, in minutes")
abline(v = minutes_levels, col = "crimson", lty = 3)

hist(patterns$breakpoint1_axis[is.finite(patterns$breakpoint1_axis)],
     breaks = 30,
     col    = "grey70",
     xlab   = sprintf("first breakpoint (%s axis)", TIME_AXIS),
     main   = "First breakpoint, on the fitting axis")
par(old_par)

cat(sprintf("median first breakpoint: %.1f min   (IQR %.1f - %.1f)\n",
            median(patterns$breakpoint1_min, na.rm = TRUE),
            quantile(patterns$breakpoint1_min, 0.25, na.rm = TRUE),
            quantile(patterns$breakpoint1_min, 0.75, na.rm = TRUE)))
```

## Trendy's own summaries

`topTrendy` keeps the sites above the adjusted-R² cutoff and reshapes them into the matrix form the
package's plotting functions expect. `trendHeatmap` then orders sites by the direction of their
first segment and by when they first change, which is the closest thing Trendy has to a clustering:
sites that turn on at the same time end up adjacent.

```{r top-trendy, fig.width=7, fig.height=8}
top_trendy <- tryCatch(Trendy::topTrendy(trendy_fits, adjR2Cut = ADJ_R2_CUT),
                       error = function(e) {
                         cat("topTrendy:", conditionMessage(e), "\n")
                         NULL
                       })

if (!is.null(top_trendy)) {
  cat(sprintf("topTrendy kept %d sites at adjusted R2 >= %.2f\n",
              length(top_trendy$AdjustedR2), ADJ_R2_CUT))
  trend_order <- Trendy::trendHeatmap(top_trendy)
  cat(sprintf("first segment up: %d | down: %d | no change: %d\n",
              length(trend_order$firstup),
              length(trend_order$firstdown),
              length(trend_order$firstnochange)))
}
```

## Sites following a specific pattern

`extractPattern` searches the fitted trend sequences for a pattern and returns the sites that
follow it together with the breakpoint at which it starts. `up-down` is the transient response —
the shape the sigmoid work had to exclude.

```{r extract-patterns}
for (wanted in list(c("up", "down"), c("down", "up"), c("up", "same"))) {
  found <- tryCatch(Trendy::extractPattern(trendy_fits,
                                           Pattern  = wanted,
                                           adjR2Cut = ADJ_R2_CUT),
                    error = function(e) NULL)
  label <- paste(wanted, collapse = " -> ")
  if (is.null(found) || nrow(found) == 0) {
    cat(sprintf("%-16s : no sites\n", label))
  } else {
    cat(sprintf("%-16s : %d sites, median turn at %.1f min\n",
                label, nrow(found),
                median(axis_to_minutes(as.numeric(found[, 2]), minutes_levels), na.rm = TRUE)))
  }
}
```

## Individual fits

`plotFeature` draws the observed replicates with the fitted segments on top. Look at these before
believing any census: the fits either visibly track the data or they do not.

```{r plot-examples, fig.width=10, fig.height=7}
pathway_proteins <- c("EGFR", "SOS1", "GAB1", "BRAF", "RAF1", "MAP2K1", "MAPK1", "MAPK3",
                      "SHC1", "GRB2", "SHOC2", "RPS6KA1", "RPS6KA3")

is_pathway <- df$protein_name[eligible] %in% pathway_proteins
show_sites <- rownames(M)[is_pathway]
show_sites <- head(show_sites[show_sites %in% names(trendy_fits)], 6)

if (length(show_sites) > 0) {
  old_par <- par(mfrow = c(2, 3))
  Trendy::plotFeature(Data          = M,
                      tVectIn       = tvect,
                      featureNames  = show_sites,
                      trendyOutData = trendy_fits,
                      simple        = TRUE,
                      xlab          = sprintf("time (%s axis)", TIME_AXIS),
                      ylab          = "log2 intensity")
  par(old_par)
} else {
  cat("no MAPK-pathway sites among the fitted set\n")
}
```

```{r plot-best-worst, fig.width=10, fig.height=7}
# The best- and worst-fitting sites that still produced a pattern. Looking at both ends of the
# quality distribution is the cheapest way to catch a systematic problem the census averages away.
with_pattern <- patterns[patterns$responsive, ]
with_pattern <- with_pattern[order(-with_pattern$adj_r2), ]
extremes <- c(head(with_pattern$site, 3), tail(with_pattern$site, 3))

old_par <- par(mfrow = c(2, 3))
Trendy::plotFeature(Data          = M,
                    tVectIn       = tvect,
                    featureNames  = extremes,
                    trendyOutData = trendy_fits,
                    simple        = TRUE,
                    xlab          = sprintf("time (%s axis)", TIME_AXIS),
                    ylab          = "log2 intensity")
par(old_par)
cat("top row: highest adjusted R2 | bottom row: lowest, among sites with a non-flat pattern\n")
```

# Verification

Three checks against things already established elsewhere in the project. None of them is a
formality: each one has a way of failing that would invalidate the census above.

```{r verify-against-limma}
# 1. Trendy and limma must broadly agree about which sites do something. They are different tests
#    (segment slopes vs contrasts against starve), so agreement will not be perfect — but if sites
#    Trendy calls flat were NOT enriched in limma-non-significant sites, the fit is not tracking
#    the data at all.
if (ffdr_column %in% colnames(df)) {
  ffdr <- df[[ffdr_column]][match(patterns$site, df$site)]
  flat <- patterns$pattern == "flat"
  cat(sprintf("median limma FFDR, sites Trendy calls flat    : %.3g  (n = %d)\n",
              median(ffdr[flat], na.rm = TRUE), sum(flat)))
  cat(sprintf("median limma FFDR, sites with a fitted trend  : %.3g  (n = %d)\n",
              median(ffdr[!flat], na.rm = TRUE), sum(!flat)))
} else {
  cat("no FFDR column in this file — skipping the limma comparison\n")
}
```

```{r verify-breakpoint-range}
# 2. Every breakpoint must lie inside the observed window. A breakpoint outside it is an
#    extrapolation: a turn claimed at a time the experiment never sampled.
bp_all <- c(patterns$breakpoint1_min, patterns$breakpoint2_min)
bp_all <- bp_all[is.finite(bp_all)]
outside <- sum(bp_all < min(minutes_levels) | bp_all > max(minutes_levels))
cat(sprintf("breakpoints outside [%g, %g] min: %d of %d\n",
            min(minutes_levels), max(minutes_levels), outside, length(bp_all)))
stopifnot(outside == 0)

# Breakpoints landing in the 15-90 min gap are interpolations into unsampled time: legal, but
# they are the fit's opinion rather than an observation, and should be reported as such.
in_gap <- sum(bp_all > 15 & bp_all < 90)
cat(sprintf("breakpoints inside the unsampled 15-90 min gap: %d (%.1f%%)  <- interpolated, not observed\n",
            in_gap, 100 * in_gap / length(bp_all)))
```

```{r verify-r2}
# 3. The adjusted R2 distribution. A pile near zero means the segments are not describing the
#    data, whatever pattern strings came out.
cat(sprintf("adjusted R2: median %.3f | IQR %.3f - %.3f | %% negative %.1f\n",
            median(patterns$adj_r2, na.rm = TRUE),
            quantile(patterns$adj_r2, 0.25, na.rm = TRUE),
            quantile(patterns$adj_r2, 0.75, na.rm = TRUE),
            100 * mean(patterns$adj_r2 < 0, na.rm = TRUE)))
hist(patterns$adj_r2, breaks = 50, col = "grey70",
     xlab = "adjusted R2", main = "Fit quality across sites")
abline(v = ADJ_R2_CUT, col = "crimson", lty = 2)
```

# Write the results

One row per **fitted** site, written next to the input in
`Experiment/{dataset}/Data/Processed/`, never over it. Sites that were not fitted (missing
replicates, or filtered out by `USE_FFDR`) are absent — the annotation columns carried through make
it possible to see which.

```{r write-results}
annotation_columns <- intersect(c("site", "protein_name", "protein_Id", "n:reps", ffdr_column),
                                colnames(df))
output <- merge(df[eligible, annotation_columns, drop = FALSE],
                patterns,
                by  = "site",
                all = FALSE)

# Provenance: what this table was produced with, so a stale file cannot be mistaken for a fresh one.
output$condition      <- CONDITION
output$cell_line      <- CELL_LINE
output$time_axis      <- TIME_AXIS
output$max_k          <- max_k_supported
output$min_num_in_seg <- MIN_NUM_IN_SEG
output$pval_cut       <- PVAL_CUT
output$plex_centered  <- PLEX_CENTER
output$normalize      <- NORMALIZE

cat(sprintf("output table: %d rows x %d columns\n", nrow(output), ncol(output)))

if (SAVE_TABLE) {
  stage(paste("write", basename(OUTPUT_FILE)),
        write.table(output,
                    file      = OUTPUT_FILE,
                    sep       = "\t",
                    row.names = FALSE,
                    quote     = FALSE,
                    na        = "NA"))
  cat("wrote", OUTPUT_FILE, "\n")
} else {
  cat("SAVE_TABLE is FALSE — nothing written.\n")
}
```

# Output columns

| Column | Meaning |
|---|---|
| `site`, `protein_name`, `protein_Id`, `n:reps` | identity, carried from the input |
| `pattern` | the result: segment trends joined by `-`, e.g. `up-down`, `up-flat`, `flat` |
| `n_segments`, `n_breakpoints` | how many pieces the fit has; `n_breakpoints = 0` means plain linear regression won on BIC |
| `first_trend`, `last_trend` | the first and last segment's direction — convenient for filtering |
| `responsive` | TRUE if any segment has a significant slope (`pvalCut`) |
| `adj_r2`, `r2_ok` | adjusted R² of the fit, and whether it clears `ADJ_R2_CUT` |
| `pattern_ok` | `responsive & r2_ok` — the flag to filter on before interpreting a pattern |
| `segmentN_slope` | slope of segment N, in log2 units per unit of the fitting axis |
| `segmentN_trend` | `up` / `down` / `flat` |
| `segmentN_pvalue` | p-value of that segment's slope |
| `breakpointN_axis` | fitted breakpoint on the fitting axis (index / minutes / log) |
| `breakpointN_min` | the same breakpoint converted to **minutes** — the interpretable one |
| `condition` … `normalize` | provenance: the settings this table was produced with |

⚠️ `segmentN_slope` is in units of the **fitting axis**. On the default `index` axis it is "log2
units per timepoint step", which is comparable *between sites* but is not a rate per minute. Convert
via `breakpointN_min` if a rate is needed, and state which axis was used.

# Caveats to carry into any interpretation

1. **Six time levels.** Everything above rests on six distinct times. A 2-breakpoint call uses all
   of them and leaves nothing to check the shape with. Treat `n_breakpoints = 2` as descriptive.
2. **Only complete sites are analysable.** Trendy refuses `NA`, so this is an analysis of the
   `n:reps == 4` subset. That subset is biased towards abundant, well-behaved peptides.
3. **Breakpoints in the 15–90 min gap are interpolations.** The verification chunk counts them.
   Nothing was measured between 15 and 90 minutes, and a turn placed at 40 min is the fit's opinion.
4. **`flat` is not the same as "no response", and it is a conservative call.** A site that rises and
   returns symmetrically within one segment is reported flat, and so is any site whose slope misses
   the 1-df significance test described above (roughly $|t| < 3.1$). Cross-check against the limma
   FFDR column, which is in the output for exactly this reason.
5. **Loading normalisation is still unfixed upstream** (CLAUDE.md → Known issues). `PLEX_CENTER`
   removes the plex-level offset, not the per-channel one. Re-run with `NORMALIZE <- "median"` to
   bracket the effect, as the sigmoid notebook does with its census.
6. **The pattern alphabet is coarse.** `up-down` covers a spike at 2 min and a broad hump peaking at
   15 min. The breakpoint time and the segment slopes distinguish them; the pattern string does not.

# Where this goes next

- **Cross-tabulate against the shape census.** `src/response_shapes.py` classifies the same sites
  into `transient` / `sustained` / `monotonic` / `biphasic` from the fold-change profile with no
  fitting at all. The two should agree on the gross split, and where they disagree is informative:
  Trendy adds a fitted turning time to the classes the census can only name.
- **Compare the turning time to T50.** For sites that are both `sigmoid_legitimate` and fitted here
  as `up-flat`, the first breakpoint and the sigmoid T50 are estimating related quantities by
  different routes — a genuine mutual validation, in the spirit of the CurveCurator cross-check.
- **Crosstalk.** Rerun with `CONDITION <- "INS"` and `"EGFnINS"` and compare patterns per site: a
  site whose pattern is `up-down` under EGF but `up-flat` under co-stimulation has had its
  *shut-off* altered, which is a different claim from an amplitude change and is invisible to the
  clustering.
- **The mutant lines.** The same fit on `hme1_lfq` (9 timepoints) is a much better-posed problem —
  `maxK = 3` becomes defensible, and the breakpoint standard errors from `segmented` become worth
  reporting.
- **Breakpoint uncertainty.** `segmented` returns a standard error for each `psi`. Trendy discards
  it (`keepFit = FALSE` drops the fit object). Setting `keepFit = TRUE` and pulling
  `fit$Fit$psi[, "St.Err"]` would give per-site breakpoint errors — necessary before any claim that
  two sites turn over at different times.

```{r session-info}
sessionInfo()
```
